\chapter{Hamiltonian Mechanics And Phase Space}
\label{chap:hamiltonian-phase-space}

Phase space is a parameter space for histories. A point
\((q,p)\in T^*Q\) is one possible state at one instant; the Hamiltonian flow
extends it forward and backward in time. Two histories that meet at the same
phase-space point are the same history, at least for as long as the equations
have a unique solution.

The Lagrangian chapter described motion as stationary paths in configuration
space. Hamiltonian mechanics reorganizes the same dynamics on phase space and
brings the symplectic form, Poisson brackets, canonical transformations, and
phase-space volume into the foreground. These tools are the language in which
integrability, action-angle variables, perturbation theory, and KAM theory are
most naturally stated.

Throughout this chapter \(Q\) is a smooth configuration manifold of dimension
\(n\), with local coordinates \(q^i\). The cotangent bundle \(T^*Q\) has
induced coordinates \((q^i,p_i)\). When a Lagrangian is regular, the Legendre
map carries \((q,\dot q)\) to \((q,p)\), and the Hamiltonian \(H\) is the
corresponding energy function written in these phase-space coordinates. The
assumptions behind this passage are stated explicitly below, because singular
Lagrangians and gauge constraints require extra reduction machinery that
belongs to more advanced treatments.

\paragraph{Roadmap.}
The chapter has four layers. First, the Legendre transform explains when
velocity data on \(TQ\) can be replaced by momentum data on \(T^*Q\). Second,
the canonical one-form and symplectic form derive Hamilton's equations from
geometry. Third, Poisson brackets, the phase-space action, and canonical
transformations identify the maps that preserve Hamiltonian structure. Fourth,
the numerical lesson explains why a simulation should preserve geometry as well
as approximate trajectories.

Keep one distinction in view throughout the chapter. A point of phase space is
instantaneous data, while a Hamiltonian flow is the rule that moves that data.
The symplectic form supplies the structure that turns the differential of
\(H\) into a vector field and decides which coordinate changes are canonical.

\section{Legendre Transform And Hamiltonian}
\label{sec:legendre-hamiltonian}

The Legendre transform sends \((q,\dot q)\in TQ\) to \((q,p)\in T^*Q\) by
\[
  p_i=\frac{\partial L}{\partial \dot q^i}.
\]
For a regular Lagrangian this can be inverted to express \(\dot q^i\) as a
function of \((q,p,t)\). Concretely, regularity means that the velocity Hessian
\[
  W_{ij}(q,\dot q,t)=
  \frac{\partial^2L}{\partial\dot q^i\partial\dot q^j}
\]
is nonsingular. The inverse function theorem then says that, locally in phase
space, the equations \(p_i=\partial L/\partial\dot q^i\) can be solved for
\(\dot q^i=\dot q^i(q,p,t)\).
The Hamiltonian is
\[
  H(q,p,t)=p_i\dot q^i(q,p,t)-L(q,\dot q(q,p,t),t).
\]

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1.0, >=Latex, font=\small]
  \node[draw, rounded corners=2pt, minimum width=2.9cm, minimum height=0.9cm,
    fill=blue!5] (tq) at (0,0) {\(TQ:\ (q,\dot q)\)};
  \node[draw, rounded corners=2pt, minimum width=2.9cm, minimum height=0.9cm,
    fill=red!5] (tsq) at (6.2,0) {\(T^*Q:\ (q,p)\)};
  \draw[->, thick] (tq.east) -- node[below=5pt, fill=white, inner sep=2pt]
    {\(\mathbb F L:\ p_i=\partial L/\partial\dot q^i\)} (tsq.west);
  \draw[->, thick]
    ([yshift=5pt]tsq.west) to[out=150,in=30]
    node[above=5pt, fill=white, inner sep=2pt] {regular inverse}
    ([yshift=5pt]tq.east);
  \node[draw, rounded corners=2pt, fill=gray!6, align=center,
    text width=0.46\textwidth, inner sep=4pt] at (3.1,-1.40)
    {regular iff \(\det(W)\neq 0\)};
\end{tikzpicture}
\caption{The Legendre transform from tangent-bundle data to cotangent-bundle
data. Singular Lagrangians require constraints; the Hamiltonian construction in
this chapter assumes the regular case.}
\label{fig:legendre-transform}
\end{figure}

\begin{derivation}[Hamilton's equations from the Legendre transform]
In this calculation \(H\) is regarded as a function of \((q,p,t)\). The
velocity \(\dot q=\dot q(q,p,t)\) is the inverse image of \((q,p)\) under the
Legendre map, so its differential appears temporarily and then cancels. This is
where regularity is used.
Differentiate \(H\):
\[
\begin{aligned}
  \dd H
  &=
  \dot q^i\,\dd p_i
  +
  p_i\,\dd \dot q^i
  -
  \frac{\partial L}{\partial q^i}\dd q^i
  -
  \frac{\partial L}{\partial \dot q^i}\dd \dot q^i
  -
  \frac{\partial L}{\partial t}\dd t\\
  &=
  \dot q^i\,\dd p_i
  -
  \frac{\partial L}{\partial q^i}\dd q^i
  -
  \frac{\partial L}{\partial t}\dd t,
\end{aligned}
\]
because \(p_i=\partial L/\partial \dot q^i\). Thus
\[
  \frac{\partial H}{\partial p_i}=\dot q^i,
  \qquad
  \frac{\partial H}{\partial q^i}=-\frac{\partial L}{\partial q^i}.
\]
Using the Euler-Lagrange equation \(\dot p_i=\partial L/\partial q^i\), we get
\[
  \boxed{
    \dot q^i=\frac{\partial H}{\partial p_i},
    \qquad
    \dot p_i=-\frac{\partial H}{\partial q^i}
  }.
\]
\end{derivation}

The cancellation of the \(\dd\dot q\) terms is the algebraic heart of the
Legendre transform. It says that once momentum has been defined by
\(p_i=\partial L/\partial\dot q^i\), the Hamiltonian differential contains
only \(\dd q,\dd p,\dd t\). Hamilton's equations are therefore a first-order
system on \(T^*Q\), with \((q,p)\) as the instantaneous phase-space data.

The Legendre transform changes the bookkeeping of initial data. A Lagrangian
state \((q,\dot q)\) records a position and a velocity; a Hamiltonian state
\((q,p)\) records a position and the covector that measures how the action
changes under endpoint displacement. The symplectic form below is built from
that endpoint one-form, so Hamiltonian mechanics retains the variational origin
of momentum.

Singular systems begin when the velocity Hessian has null directions. Then some
velocities cannot be solved from the momenta, and some momenta become
constraints. Electromagnetic gauge theory, relativistic
reparametrization-invariant particles, and ideal incompressible fluids all have
versions of this phenomenon. The present chapter stays in the regular case so
that the geometry of \(T^*Q\) can be developed before Dirac constraints enter.

\section{Phase Space As Initial Data And Histories}
\label{sec:phase-space-solutions}

For a deterministic Hamiltonian system, phase space parametrizes the solutions.
If \(H\) is smooth and the Hamiltonian vector field is complete on the time
interval under study, then every point
\[
  z_0=(q_0,p_0)\in T^*Q
\]
determines a unique trajectory \(z(t)=\Phi^t_H(z_0)\), and every trajectory is
obtained this way from its value at any one time. Thus phase space is the set
of all possible instantaneous states, but it is also a coordinate system on the
family of all classical histories. The caveat is that this identification uses
existence and uniqueness of the equations of motion; singular potentials,
collisions, constraints, or finite-time blowup can invalidate the global
statement even when the local Hamiltonian equations make sense. For example,
the attractive Kepler Hamiltonian has regular equations away from \(r=0\), but
a collision orbit reaches \(r=0\) in finite time unless one adds a separate
regularization. Completeness is therefore a statement about the vector field on
the chosen phase space, not a formal consequence of writing Hamilton's
equations.

\section{The Symplectic Form}
\label{sec:symplectic-form}

The cotangent bundle \(T^*Q\) carries a canonical one-form, often called the
Liouville one-form,
\[
  \Theta=p_i\,\dd q^i.
\]
The name credits Liouville's nineteenth-century work on volume preservation and
integrability in Hamiltonian systems; the geometric object is the tautological
one-form on \(T^*Q\).
This formula is intrinsic under changes of coordinates on the base manifold
\(Q\), although it is written in induced cotangent coordinates. If
\(\tilde q^j=\tilde q^j(q)\) and the induced cotangent coordinates satisfy
\[
  \tilde p_j=\frac{\partial q^i}{\partial\tilde q^j}p_i,
\]
then the chain rule gives
\[
  \tilde p_j\,\dd\tilde q^j
  =
  \frac{\partial q^i}{\partial\tilde q^j}p_i
  \frac{\partial\tilde q^j}{\partial q^k}\dd q^k
  =
  p_i\,\dd q^i.
\]
Thus \(\Theta\) is the same one-form in every coordinate system on \(Q\).
The canonical symplectic form is
\[
  \omega=-\dd\Theta=\dd q^i\wedge\dd p_i.
\]
The sign comes from the antisymmetry of the wedge product:
\[
  -\dd(p_i\,\dd q^i)
  =
  -\dd p_i\wedge\dd q^i
  =
  \dd q^i\wedge\dd p_i.
\]
This sign convention fixes the later Hamiltonian equation
\(\iota_{X_H}\omega=\dd H\). Some standard references, including Arnold's
\emph{Mathematical Methods of Classical Mechanics}, use the opposite sign for
the canonical symplectic form and compensate by the corresponding convention
for Hamiltonian vector fields or Poisson brackets. The geometry is unchanged;
only the sign bookkeeping changes.
It is closed because
\[
  \dd\omega=-\dd^2\Theta=0.
\]
In arbitrary local phase-space coordinates
\[
  x^m,\qquad m=1,\ldots,2n,
\]
one can write
\[
  \omega=\frac12\omega_{mn}(x)\,\dd x^m\wedge\dd x^n,
\]
where \(\omega_{mn}=-\omega_{nm}\), \(\det(\omega_{mn})\neq0\), and
\[
  \dd\omega=0.
\]
The closedness condition is what makes \(\omega\) locally equivalent to the
canonical form; this is Darboux's theorem (1882). Nondegeneracy is what lets
the Hamiltonian determine a unique vector field. In canonical coordinates
\(x=(q,p)\),
\[
  (\omega_{mn})=
  \begin{pmatrix}
  0 & I_n\\
  -I_n & 0
  \end{pmatrix}.
\]
Darboux's proof can be outlined as a local induction. Choose a function
\(q^1\) with nonzero differential and define \(Y_1\) by
\(\iota_{Y_1}\omega=-\dd q^1\). Nondegeneracy makes \(Y_1\) unique, and
\(\dd q^1(Y_1)=\omega(Y_1,Y_1)=0\), so \(Y_1\) is tangent to the level sets of
\(q^1\). Choose a local hypersurface transverse to \(Y_1\), set \(p_1=0\) on it,
and extend \(p_1\) by requiring \(Y_1p_1=1\) along the \(Y_1\)-flow. Then
\(\{q^1,p_1\}=1\), with the sign convention below. The common kernel of
\(\dd q^1\) and \(\dd p_1\), modulo the two directions already constructed,
carries a nondegenerate two-form of dimension \(2n-2\). Closedness is inherited
because restriction and quotient pullbacks commute with \(\dd\); this is the
point at which \(\dd\omega=0\) enters the induction. Repeating the
construction produces a coordinate chart in which
\(\omega=\dd q^i\wedge\dd p_i\).
The Poisson bracket can be expressed intrinsically by the bivector inverse to
the symplectic form, with the sign convention fixed by Hamilton's equations
below. Once \(\omega\) is given, the bracket is the corresponding Lie operation
on functions.
In arbitrary coordinates, let \(\pi^{mn}\) be defined by
\[
  \pi^{mn}\omega_{nk}=\delta^m{}_k.
\]
With the convention \(\iota_{X_H}\omega=\dd H\), this gives
\[
  \Poisson{f}{g}
  =
  -\pi^{mn}\partial_m f\,\partial_n g,
\]
which reduces to the displayed canonical formula in coordinates
\((q^i,p_i)\).
Indeed, for
\[
  \omega=
  \begin{pmatrix}
  0&I\\ -I&0
  \end{pmatrix},
  \qquad
  \pi=\omega^{-1}=
  \begin{pmatrix}
  0&-I\\ I&0
  \end{pmatrix},
\]
the extra minus sign gives
\(\Poisson{f}{g}=\partial_{q^i}f\,\partial_{p_i}g
-\partial_{p_i}f\,\partial_{q^i}g\).

The two defining properties of \(\omega\) have different jobs. Nondegeneracy
lets one solve \(\iota_{X_H}\omega=\dd H\) for a unique vector field, just as a
mass matrix lets one solve force equals mass times acceleration. Closedness is
the integrability condition behind canonical coordinates, Stokes-type
arguments, and the preservation of symplectic area. Hamiltonian mechanics needs
both.

The Hamiltonian vector field \(X_H\) is defined by
\[
  \iota_{X_H}\omega=\dd H.
\]
Here \(\iota_X\) denotes interior contraction: if \(\omega\) is a two-form, then
\((\iota_X\omega)(Y)=\omega(X,Y)\) for every vector field \(Y\).
The coordinate-free formulation became standard in the twentieth century
through Lichnerowicz's symplectic-geometry program, Kirillov's coadjoint-orbit
work, and Souriau's \emph{Structure des systemes dynamiques} (1970). The
formula is not an additional physical postulate; it is Hamilton's equations
written without choosing canonical coordinates.
If
\[
  X_H=A^i\frac{\partial}{\partial q^i}
  +
  B_i\frac{\partial}{\partial p_i},
\]
then
\[
  \iota_{X_H}\omega=A^i\,\dd p_i-B_i\,\dd q^i.
\]
Equating this with
\[
  \dd H=\frac{\partial H}{\partial q^i}\dd q^i
  +
  \frac{\partial H}{\partial p_i}\dd p_i
\]
gives
\[
  A^i=\frac{\partial H}{\partial p_i},
  \qquad
  B_i=-\frac{\partial H}{\partial q^i}.
\]
Thus \(X_H\) is Hamilton's equation written geometrically.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1.0, >=Latex]
  \draw[->] (-0.25,0) -- (6.9,0) node[right] {\(q\)};
  \draw[->] (3.35,-2.85) -- (3.35,2.85) node[above] {\(p\)};
  \node[below] at (0.25,0) {\(-\pi\)};
  \node[below] at (3.35,0) {\(0\)};
  \node[below] at (6.45,0) {\(\pi\)};
  \foreach \E/\qmax in {0.35/0.86,0.90/1.47,1.50/2.09}
  {
    \draw[blue!70!black, thick, domain=-\qmax:\qmax, samples=120]
      plot ({\x+3.35},{sqrt(2*(\E-1+cos(\x r)))});
    \draw[blue!70!black, thick, domain=-\qmax:\qmax, samples=120]
      plot ({\x+3.35},{-sqrt(2*(\E-1+cos(\x r)))});
  }
  \draw[red!70!black, thick, domain=-3.12:3.12, samples=180]
    plot ({\x+3.35},{2.0*cos(0.5*\x r)});
  \draw[red!70!black, thick, domain=-3.12:3.12, samples=180]
    plot ({\x+3.35},{-2.0*cos(0.5*\x r)});
  \draw[gray!65, thick, domain=-3.12:3.12, samples=180]
    plot ({\x+3.35},{sqrt(2*(2.35-1+cos(\x r)))});
  \draw[gray!65, thick, domain=-3.12:3.12, samples=180]
    plot ({\x+3.35},{-sqrt(2*(2.35-1+cos(\x r)))});
  \draw[->, thick] (4.02,0.90) -- (4.60,0.60);
  \node[fill=white, inner sep=1pt] at (4.95,0.88) {\(X_H\)};
  \node[blue!70!black, fill=white, inner sep=1pt] at (1.25,2.28)
    {libration levels};
  \node[fill=white, inner sep=1pt] at (3.35,-0.28) {\(0\)};
  \node[red!70!black, fill=white, inner sep=1pt] at (5.25,-2.10)
    {separatrix};
  \node[gray!65!black, fill=white, inner sep=1pt] at (1.25,-2.35)
    {rotation levels};
  \node[fill=white, inner sep=1.5pt] at (3.35,-3.08)
    {Exact level sets of \(H(q,p)=\frac12p^2+1-\cos q\).};
\end{tikzpicture}
\caption{The pendulum phase plane. Libration curves have
\(0<E<2\), the separatrix has \(E=2\), and rotation curves have \(E>2\).
The Hamiltonian vector field is tangent to energy level sets because
\(\dot H=\{H,H\}=0\).}
\label{fig:pendulum-phase-plane}
\end{figure}

\section{Poisson Brackets And Time Evolution}
\label{sec:poisson-time-evolution}

For \(f,g\in C^\infty(T^*Q)\), define
\[
  \Poisson{f}{g}
  =
  \frac{\partial f}{\partial q^i}
  \frac{\partial g}{\partial p_i}
  -
  \frac{\partial f}{\partial p_i}
  \frac{\partial g}{\partial q^i}.
\]
If \(z(t)=(q(t),p(t))\) follows Hamilton's equation, then
\[
  \frac{\dd}{\dd t}f(q(t),p(t),t)
  =
  \frac{\partial f}{\partial t}+\Poisson{f}{H}.
\]
A time-independent observable \(f\) is conserved exactly when
\[
  \Poisson{f}{H}=0.
\]

\begin{proposition}[Liouville volume preservation]
Hamiltonian flow preserves the phase-space volume form
\[
  \Omega=\frac{1}{n!}\omega^n.
\]
\end{proposition}

\begin{proof}
Liouville's 1838 volume-preservation theorem is a direct consequence of
Cartan's formula for Lie derivatives.\footnote{For any differential form
\(\alpha\) and vector
field \(X\), \(\mathcal L_X\alpha=\dd(\iota_X\alpha)+\iota_X\dd\alpha\). One
proves it first on functions and one-forms, then uses that both sides are
graded derivations of the exterior algebra.}
\[
  \Lie_{X_H}\omega=\dd(\iota_{X_H}\omega)+\iota_{X_H}\dd\omega.
\]
Since \(\iota_{X_H}\omega=\dd H\) and \(\dd\omega=0\),
\[
  \Lie_{X_H}\omega=\dd\dd H=0.
\]
Therefore \(\Lie_{X_H}\omega^n=0\), so the associated volume form is preserved.
\end{proof}

\section{Phase-Space Action}
\label{sec:phase-space-action}

Hamilton's equations can also be derived from the phase-space action
\[
  S_H[q,p]=\int_{t_0}^{t_1}\left(p_i\dot q^i-H(q,p,t)\right)\dd t.
\]
Varying \(q\) and \(p\) independently, with \(\delta q(t_0)=\delta q(t_1)=0\),
gives
\[
  \delta S_H
  =
  \int_{t_0}^{t_1}
  \left[
    \left(\dot q^i-\frac{\partial H}{\partial p_i}\right)\delta p_i
    -
    \left(\dot p_i+\frac{\partial H}{\partial q^i}\right)\delta q^i
  \right]\dd t.
\]
Since \(\delta q\) and \(\delta p\) are arbitrary in the interior, the
stationary curves are exactly Hamiltonian trajectories.

\section{Canonical Transformations}
\label{sec:canonical-transformations}

A diffeomorphism \(\Phi:T^*Q\to T^*Q\) is canonical, or symplectic, if
\[
  \Phi^*\omega=\omega.
\]
Canonical transformations preserve Poisson brackets:
\[
  \Poisson{f\circ\Phi}{g\circ\Phi}
  =
  \Poisson{f}{g}\circ\Phi.
\]
Indeed, \(\Phi^*\omega=\omega\) implies that the inverse bivector is also
preserved, \(\Phi_*\pi=\pi\). Applying the bivector formula for the bracket to
\(\dd(f\circ\Phi)=\Phi^*(\dd f)\) and
\(\dd(g\circ\Phi)=\Phi^*(\dd g)\) gives the displayed identity.
This is why canonical coordinates may change but Hamiltonian mechanics keeps
the same form. Later, action-angle variables will be canonical coordinates
adapted to integrable motion.

\begin{derivation}[Generating function for a canonical transformation]
Let \((q,p)\) be old canonical coordinates and \((Q,P)\) new ones. A convenient
type of generating function depends on the new momenta and old positions:
\[
  F=F(P,q).
\]
The starting point is preservation of the canonical two-form:
\[
  \dd q^i\wedge\dd p_i=\dd Q^i\wedge\dd P_i.
\]
Equivalently, the one-form \(p_i\dd q^i+Q^i\dd P_i\) is locally exact, because
\[
  \dd(p_i\dd q^i+Q^i\dd P_i)
  =
  -\dd q^i\wedge\dd p_i+\dd Q^i\wedge\dd P_i
  =
  0.
\]
By the Poincare lemma, a closed one-form on a contractible coordinate patch is
the differential of a local function. The type-II generating function is this
local potential.
Impose
\[
  Q^i=\frac{\partial F}{\partial P_i}\bigg|_q,
  \qquad
  p_i=\frac{\partial F}{\partial q^i}\bigg|_P.
\]
Then
\[
  \dd F=Q^i\,\dd P_i+p_i\,\dd q^i.
\]
Taking another exterior derivative gives
\[
  0=\dd Q^i\wedge\dd P_i+\dd p_i\wedge\dd q^i.
\]
Therefore
\[
  \dd Q^i\wedge\dd P_i
  =
  \dd q^i\wedge\dd p_i.
\]
The canonical symplectic form is preserved, so the transformation is canonical.
In coordinates this is the same cancellation one sees by expanding
\[
  \dd Q^i\wedge\dd P_i
  =
  \left(
    \frac{\partial^2F}{\partial P_i\partial q^j}\dd q^j
    +
    \frac{\partial^2F}{\partial P_i\partial P_j}\dd P_j
  \right)
  \wedge \dd P_i,
\]
where the pure \(\dd P_i\wedge\dd P_j\) term vanishes because second
derivatives are symmetric while the wedge product is antisymmetric.
\end{derivation}

The generating function is useful because it replaces the condition
\(\Phi^*\omega=\omega\), which is a two-form equation, by the exactness of a
one-form. Locally, a canonical transformation is often found by guessing the
right exact differential. Globally, the same calculation must be patched across
coordinate charts, which is why action-angle generating functions can be
multivalued even when the symplectic form is perfectly well defined.

The type-II function used above is one member of the standard local family,
systematized in Jacobi's Hamiltonian lectures of the 1840s. The four common
choices are distinguished by which old and new variables are treated as
independent:
\[
\begin{array}{c|c|c}
\text{type} & \text{generating function} & \text{differential convention}\\
\hline
F_1(q,Q) & F_1 & p\,\dd q-P\,\dd Q=\dd F_1\\
F_2(q,P) & F_2 & p\,\dd q+Q\,\dd P=\dd F_2\\
F_3(p,Q) & F_3 & -q\,\dd p-P\,\dd Q=\dd F_3\\
F_4(p,P) & F_4 & -q\,\dd p+Q\,\dd P=\dd F_4
\end{array}
\]
Each line is a different Legendre transform of the same exact one-form.
The signs follow from preserving \(p\,\dd q-P\,\dd Q\) up to an exact
differential.

\section{Numerical Lesson}
\label{sec:numerical-lesson}

The equations in this chapter are exact statements about smooth systems. A
numerical integrator creates a discrete map
\[
  z_{n+1}=\Psi_{\Delta t}(z_n).
\]
For long-time Hamiltonian problems, whether \(\Psi_{\Delta t}\) preserves a
symplectic form or a modified Hamiltonian can matter more than the local order
of the method. More precisely, a symplectic method of order \(p\) often has a
formal modified Hamiltonian
\[
  \widetilde H_{\Delta t}
  =
  H+\Delta t^p H_p+\Delta t^{p+1}H_{p+1}+\cdots
\]
whose exact time-\(\Delta t\) flow agrees with the numerical map up to the
relevant asymptotic order. The series is generally asymptotic rather than
convergent, but for analytic problems it explains why energy error often stays
bounded over very long times instead of drifting monotonically. The pendulum,
standard map, and asteroid demos are therefore experiments in preserving
structure.
Symplectic integrators were developed in modern numerical form in the 1980s
and 1990s. Ruth (1983) gave the first explicit schemes; Channell-Scovel and
Sanz-Serna systematized them; Yoshida (1990) constructed high-order composition
methods. Hairer, Lubich, and Wanner later assembled the backward-error analysis
that explains why these methods preserve a modified Hamiltonian over long
times.

The practical lesson is diagnostic. A good mechanics simulation reports not
only a trajectory but also the quantities that the continuous problem says
should be preserved or nearly preserved. Energy drift, angular-momentum drift,
phase-space area, and event criteria are part of the model audit trail. Without
them, a numerical plot can look convincing while solving the wrong mechanical
problem.

\section{What To Take Forward}
\label{sec:hamiltonian-take-forward}

The rest of the notes repeatedly reuse three lessons sharpened in this
chapter.
\begin{itemize}
\item The Legendre transform is a structural step, not a change of notation:
it is valid only where velocities can be recovered from momenta.
\item The canonical one-form records the endpoint momentum information from
the Lagrangian action; its exterior derivative is the symplectic form that
controls Hamiltonian evolution.
\item Hamiltonian mechanics is geometry on \(T^*Q\): the symplectic form
determines Hamilton's equations, volume preservation, Poisson brackets, and the
meaning of a canonical numerical method.
\end{itemize}

\section{Chapter-End Laboratories And Exercises}
\label{sec:hamiltonian-chapter-end-labs}

The laboratories below test when a Lagrangian can be converted into a
Hamiltonian system and how the resulting phase portrait organizes motion.

\subsection{A Nonstandard Lagrangian And Regularity}
\label{sec:lab-nonstandard-lagrangian}

This laboratory separates two ideas that are often conflated. A Lagrangian may
produce a familiar-looking equation of motion and still fail to be globally
regular. Regularity is a statement about whether velocities can be solved from
momenta. Without it, the Legendre transform can fold phase space and the
Hamiltonian description must be handled with care.

Consider the one-dimensional Lagrangian
\[
  L(x,\dot x)
  =
  \frac{m^2\dot x^4}{12}
  +
  m\dot x^2V(x)
  -
  V(x)^2.
\]
The momentum is
\[
  p=\frac{\partial L}{\partial\dot x}
  =
  \frac{m^2}{3}\dot x^3+2mV\dot x.
\]
The Euler-Lagrange equation gives
\[
\begin{aligned}
  0
  &=
  \frac{\dd}{\dd t}
  \left(
    \frac{m^2}{3}\dot x^3+2mV\dot x
  \right)
  -
  \left(
    m\dot x^2V'-2VV'
  \right)\\
  &=
  (m\dot x^2+2V)(m\ddot x+V').
\end{aligned}
\]
There are two branches. The Newtonian branch is
\[
  m\ddot x=-V'(x).
\]
The second branch,
\[
  m\dot x^2+2V(x)=0,
\]
is a singular energy-surface condition. The velocity Hessian
\[
  \frac{\partial^2L}{\partial\dot x^2}
  =
  m(m\dot x^2+2V)
\]
vanishes on this second branch. This example shows how a Lagrangian
can produce familiar equations on a regular branch while being singular on
another branch. Regularity is the hypothesis that lets the Legendre transform
define a Hamiltonian system.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1.0, >=Latex, font=\small]
  \draw[->] (-3.0,0) -- (3.15,0) node[right] {\(\dot x\)};
  \draw[->] (0,-2.2) -- (0,2.25) node[above] {\(p\)};
  \draw[blue!70!black, thick, domain=-2.55:2.55, samples=180]
    plot (\x,{0.33*\x*\x*\x-0.9*\x});
  \draw[dashed, red!75!black] (-0.95,-1.9) -- (-0.95,1.9);
  \draw[dashed, red!75!black] (0.95,-1.9) -- (0.95,1.9);
  \fill[red!75!black] (-0.95,0.57) circle (1.3pt);
  \fill[red!75!black] (0.95,-0.57) circle (1.3pt);
  \node[black, fill=white, inner sep=1pt] (wleft) at (-1.45,1.72) {\(W=0\)};
  \node[black, fill=white, inner sep=1pt] (wright) at (1.55,-1.72) {\(W=0\)};
  \draw[thin, ->] (wleft.south) -- (-0.95,1.15);
  \draw[thin, ->] (wright.north west) -- (0.95,-1.15);
  \draw[decorate, decoration={brace, amplitude=6pt}, thick]
    (1.45,-0.57) -- (1.45,0.57);
  \draw[->, gray!70] (2.42,0.30) -- (1.50,0.28);
  \node[align=left, fill=white, inner sep=1.5pt] at (2.52,0.88)
    {same \(p\),\\multiple \(\dot x\)};
  \node[blue!70!black, align=center, fill=white, inner sep=2pt] at (2.05,1.45)
    {folded\\Legendre map};
\end{tikzpicture}
\caption{Nonstandard Lagrangian laboratory. For a representative region with
\(V<0\), the momentum-velocity map can fold where the velocity Hessian
\(W=\partial^2L/\partial\dot x^2\) vanishes. At such points the Legendre
transform is not locally invertible, even though the Euler-Lagrange equation has
a familiar Newtonian branch.}
\label{fig:lab-nonstandard-legendre-fold}
\end{figure}
\FloatBarrier

\subsection{Pendulum Phase Space And The Separatrix}
\label{sec:detailed-pendulum}

The simple pendulum is the canonical one-degree-of-freedom Hamiltonian example.
Let \(\theta\) be the angle from the downward vertical. The Lagrangian is
\[
  L=\frac12m\ell^2\dot\theta^2+mg\ell\cos\theta.
\]
The momentum and Hamiltonian are
\[
  p_\theta=m\ell^2\dot\theta,
  \qquad
  H(\theta,p_\theta)
  =
  \frac{p_\theta^2}{2m\ell^2}-mg\ell\cos\theta.
\]
Hamilton's equations are
\[
  \dot\theta=\frac{p_\theta}{m\ell^2},
  \qquad
  \dot p_\theta=-mg\ell\sin\theta.
\]
The equilibria are
\[
  (\theta,p_\theta)=(0,0)
  \quad\text{and}\quad
  (\pi,0)
\]
modulo \(2\pi\). The downward equilibrium is elliptic; the upward equilibrium
is hyperbolic.

The separatrix is the energy level through the hyperbolic point:
\[
  H=mg\ell.
\]
On this level,
\[
  \frac{p_\theta^2}{2m\ell^2}
  =
  mg\ell(1+\cos\theta)
  =
  2mg\ell\cos^2\frac{\theta}{2}.
\]
Therefore
\[
  p_\theta
  =
  \pm2m\ell\sqrt{g\ell}\cos\frac{\theta}{2}
\]
for \(-\pi\leq\theta\leq\pi\). The separatrix divides libration orbits, which
oscillate around \(\theta=0\), from rotation orbits, which wind around the
circle. This picture is the one-degree-of-freedom ancestor of separatrix
splitting in perturbed Hamiltonian systems.

\section{Associated Demos}
\label{sec:hamiltonian-associated-demos}

The associated phase-space demo is
\path{demos/python/hamiltonian_pendulum.py}.
It plots pendulum trajectories and the separatrix in \((\theta,p_\theta)\).
Read it as a check on this chapter's central claim: phase portraits display
Hamiltonian level sets and the symplectic flow on phase space.
