\chapter{Perturbation, KAM, And Breakdown Of Invariant Tori}
\label{chap:perturbation-chaos}

\section{What This Chapter Studies}
\label{sec:perturbation-chapter-purpose}

A motivating physical example is a massless asteroid perturbed by Jupiter while
orbiting the Sun. The unperturbed problem is Kepler motion, one of the most
integrable systems in classical mechanics. The perturbation introduces
resonances, broken invariant tori, and chaotic transport.  The full restricted
three-body formulation, including Lagrange points and ejection statistics, is
developed separately in Chapter~\ref{chap:three-body-problem}.

Historically, this is the question that broke the dream of solving mechanics
by writing enough first integrals. Poincare's work on the three-body problem
showed that a small perturbation of an integrable system can fold stable and
unstable manifolds into a tangle. The modern story is more nuanced than
``small perturbations make chaos.'' Kolmogorov, Arnold, and Moser found the
survival theorem hidden inside the failure: sufficiently irrational invariant
tori persist, while resonant regions become the sites of islands, separatrix
splitting, and transport. The standard map is the compact laboratory in this
chapter; the asteroid problem is the gravitational setting to which we return
after the perturbative mechanisms are clear.

\begin{assumption}[Celestial perturbation model]
When a celestial example is needed, this chapter uses a planar restricted
model:
\begin{enumerate}
\item the Sun has mass \(M_\odot\);
\item Jupiter has mass \(M_J\) and follows a prescribed circular orbit;
\item asteroids are massless test particles;
\item asteroid-asteroid interactions are neglected;
\item all motion is confined to one plane.
\end{enumerate}
This model is not a high-precision Solar System ephemeris. It is a controlled
Hamiltonian laboratory for perturbation, resonance, and transport.
\end{assumption}

\paragraph{Roadmap.}
The chapter begins with the Kepler problem because perturbation theory needs an
integrable reference system. It then uses the action-angle machinery from
Section~\ref{sec:action-angle-variables} in perturbative form:
near-identity generating functions try to remove angle dependence, while small
divisors explain why this attempt is delicate. Diophantine conditions then
identify the tori for which the calculation can be controlled. The KAM theorem
gives the positive result: many sufficiently irrational tori persist under
small perturbation. The breakdown section gives the complementary mechanisms:
resonant normal forms, island widths, resonance overlap, separatrix splitting,
and cantori. The Poincare-map and standard-map sections then make these
mechanisms concrete in two-dimensional return maps. The three-body chapter then
uses the same mechanisms to interpret a numerical asteroid ejection experiment.

There are three levels of description to keep separate. The analytic level asks
whether a formal canonical transformation can remove angle dependence. The
geometric level asks which invariant tori, islands, and separatrices organize
phase space. The numerical level asks how an ensemble moves through that
geometry over a finite time. The asteroid ejection probability in
Chapter~\ref{chap:three-body-problem} belongs to the third level, but it is
meaningful only because the first two levels explain where transport channels
come from.

\paragraph{How approximations are labelled.}
Perturbation theory is useful only when the bookkeeping is explicit. In this
chapter \(H_0\) denotes the exactly integrable reference problem, \(\eps H_1\)
denotes a controlled perturbation, \(O(\eps^2)\) denotes terms dropped by a
first-order calculation, and resonance conditions such as \(k\cdot\omega=0\)
identify where that first-order calculation is least trustworthy. Numerical
experiments later in the chapter do not replace these estimates; they test the
transport suggested by them over a stated finite time horizon.

\section{The Kepler Problem}
\label{sec:kepler-perturbation}

For two bodies of masses \(m_1,m_2\), let
\[
  r=x_1-x_2
\]
be the relative position and
\[
  \mu=\frac{m_1m_2}{m_1+m_2}
\]
the reduced mass. The relative Hamiltonian is
\[
  H(r,p)=\frac{\norm{p}^2}{2\mu}-\frac{Gm_1m_2}{\norm r}.
\]
For a massless asteroid around the Sun, one usually divides by asteroid mass and
works with the specific Hamiltonian
\[
  h(r,v)=\frac12\norm v^2-\frac{GM_\odot}{r},
  \qquad r=\norm r.
\]

\begin{derivation}[Kepler orbit equation]
The angular momentum per unit mass is
\[
  \ell=r^2\dot\theta.
\]
This \(\ell\) is the specific angular momentum. In
Chapter~\ref{chap:lagrangian}, the two-body Kepler derivation used the massful
angular momentum \(\ell_{\mathrm{mass}}=\mu r^2\dot\theta\); the two symbols
differ by the reduced mass because this chapter has divided the Hamiltonian by
the asteroid mass. Section~\ref{subsec:detailed-binet} gives the same Binet
equation with an arbitrary particle mass \(m\); setting \(m=1\) gives the
specific convention used here.
For a central potential, \(\ell\) is conserved. The radial equation is
\[
  \ddot r-r\dot\theta^2=-\frac{GM_\odot}{r^2}.
\]
Set \(u=1/r\). Using
\[
  \frac{\dd}{\dd t}=\dot\theta\frac{\dd}{\dd\theta}
  =
  \ell u^2\frac{\dd}{\dd\theta},
\]
the radial equation loses its explicit time dependence and becomes Binet's
equation for the orbit shape:
\[
  \frac{\dd^2u}{\dd\theta^2}+u=\frac{GM_\odot}{\ell^2}.
\]
Therefore
\[
  u(\theta)=\frac{GM_\odot}{\ell^2}\left(1+e\cos(\theta-\varpi)\right),
\]
or
\[
  \boxed{
  r(\theta)=\frac{a(1-e^2)}{1+e\cos(\theta-\varpi)}
  }.
\]
Here \(a\) is the semimajor axis, \(e\) is eccentricity, and \(\varpi\) is the
longitude of periapse.
\end{derivation}

Bound Kepler orbits have energy
\[
  h=-\frac{GM_\odot}{2a}.
\]
Kepler's third law follows from the period \(T\):
\[
  n^2a^3=GM_\odot,\qquad n=\frac{2\pi}{T}.
\]

\section{Action-Angle View Of Kepler Motion}
\label{sec:kepler-action-angle}

The Kepler problem is more than solvable; it is superintegrable. For
perturbation theory, the most useful idea is that unperturbed motion can be
written in action-angle form:
\[
  H_0=H_0(I),
  \qquad
  \dot I=0,
  \qquad
  \dot\theta=\omega(I).
\]
For a near-integrable Hamiltonian,
\[
  H(I,\theta)=H_0(I)+\eps H_1(I,\theta),
\]
the actions are no longer exactly constant:
\[
  \dot I=-\eps\frac{\partial H_1}{\partial \theta},
  \qquad
  \dot\theta=\frac{\partial H_0}{\partial I}
  +
  \eps\frac{\partial H_1}{\partial I}.
\]
The small parameter \(\eps\) should be interpreted physically. In the
Sun-Jupiter-asteroid problem, a natural scale is \(M_J/M_\odot\), though close
encounters and resonances can amplify its effect.

\section{Resonance}
\label{sec:resonance}

\begin{definition}[Resonance]
A resonance occurs when there is a nonzero integer vector \(k\in\Z^n\) such
that
\[
  k\cdot\omega(I)=0.
\]
Near resonance, phase combinations that normally average out become slow and
can accumulate over long times.
\end{definition}

For mean-motion resonance with Jupiter, a \(p:q\) resonance means
\[
  \frac{n}{n_J}\approx \frac{p}{q}.
\]
Using Kepler's law,
\[
  \frac{n}{n_J}\approx
  \left(\frac{a_J}{a}\right)^{3/2},
\]
so the nominal resonance location is
\[
  \boxed{
  a_{p:q}=a_J\left(\frac{q}{p}\right)^{2/3}
  }.
\]
This formula is only the center of a resonance zone. The width of the zone
depends on the perturbation strength, eccentricity, and the order of the
resonance. The point of marking the nominal location is to identify where slow
angle combinations can form; the later chaos discussion explains what happens
inside and near those zones.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[x=0.72cm,y=1cm, >=Latex, font=\small]
  \draw[->] (0,0) -- (13.8,0) node[right] {\(a\) [AU]};
  \foreach \x/\lab in {0/2.0,2/2.5,4/3.0,6/3.5,8/4.0,10/4.5,12/5.0}
    \draw (\x,0.07) -- (\x,-0.07) node[below] {\lab};
  \draw[ultra thick, blue!35] (0.1,0) -- (7.0,0);
  \node[blue!60!black, fill=white, inner sep=1.5pt] at (3.6,0.42)
    {asteroid-belt-like annulus};
  \foreach \x/\lab/\y in {2.008/3:1/1.25,3.302/5:2/1.70,3.833/7:3/1.35,5.114/2:1/1.70}
  {
    \draw[red!75!black, thick] (\x,-0.18) -- (\x,0.92);
    \draw[red!75!black, thin] (\x,0.92) -- (\x,\y-0.10);
    \node[red!75!black, fill=white, inner sep=1pt] at (\x,\y) {\lab};
  }
  \fill[orange!80!black] (12.82,0) circle (2.2pt);
  \node[orange!80!black, above=3pt] at (12.82,0.12) {Jupiter};
\end{tikzpicture}
\caption{Nominal mean-motion resonance locations with Jupiter, shown
across an asteroid-belt-like range. The resonance positions use
\(a_{p:q}=a_J(q/p)^{2/3}\); the figure now includes Jupiter's orbital radius to
show that the listed resonances lie interior to Jupiter.}
\label{fig:jupiter-resonance-locations}
\end{figure}

\section{Perturbing An Integrable System}
\label{sec:perturbing-integrable-system}

The central perturbation question is the following. Suppose the unperturbed
Hamiltonian is integrable and has action-angle coordinates:
\[
  H_0=H_0(I),
  \qquad
  (I,\theta)\in U\times\mathbb T^N,
  \qquad
  \dot I=0,\quad \dot\theta=\omega_0(I),
\]
where
\[
  \omega_0(I)=\frac{\partial H_0}{\partial I}.
\]
Now add a small generic perturbation:
\[
  H(I,\theta)=H_0(I)+\eps H_1(I,\theta),
  \qquad
  0<\eps\ll 1.
\]
What happens to the invariant tori \(I=\text{constant}\)?

Historically, two tempting answers were both too simple. One might hope that
the perturbed system is still integrable after a clever change of variables.
One might also expect that a generic perturbation destroys the tori and makes a
typical orbit wander ergodically through the available energy surface. KAM
theory gives the subtler answer: for small analytic perturbations, most
sufficiently irrational tori persist, while resonant regions create islands,
separatrices, and chaotic transport.

The historical landmarks are the Fermi-Pasta-Ulam-Tsingou numerical experiment
of 1953, which contradicted naive ergodicity, and the
Kolmogorov-Arnold-Moser work from roughly 1954--1963, which explained why
small perturbations can leave most invariant tori intact while still allowing
chaotic regions.

\section{Canonical Perturbation Theory And Small Divisors}
\label{sec:canonical-perturbation-small-divisors}

We first ask whether a canonical transformation can remove the angle dependence
order by order. The system is no longer exactly separable, so the best one can
usually hope for is a near-identity generating function that makes the new
Hamiltonian independent of angles up to a controlled order in \(\eps\). Use a
type-II generating function
\[
  S(I',\theta)=I'\cdot\theta+\eps S_1(I',\theta)+\eps^2S_2+\cdots.
\]
The transformation is
\[
  I_i=\frac{\partial S}{\partial \theta^i},
  \qquad
  \theta'^i=\frac{\partial S}{\partial I'_i}.
\]
To first order,
\[
  I=I'+\eps\frac{\partial S_1}{\partial \theta}+O(\eps^2).
\]
If the transformed Hamiltonian is angle-independent to order \(\eps\),
\[
  H'(I')=H(I,\theta),
\]
then expanding gives
\[
  H'(I')
  =
  H_0(I')
  +
  \eps\left[
    \omega_0(I')\cdot\frac{\partial S_1}{\partial \theta}
    +
    H_1(I',\theta)
  \right]
  +
  O(\eps^2).
\]
Thus we need
\[
  \omega_0(I')\cdot\frac{\partial S_1}{\partial \theta}
  =
  f_1(I')-H_1(I',\theta),
\]
where \(f_1(I')\) will become the first correction to the new Hamiltonian.
Since \(S_1\) must be periodic in each angle, the average over the torus of the
left side is zero. Therefore
\[
  f_1(I')=\langle H_1(I',\theta)\rangle_\theta.
\]

Expand the perturbation in Fourier modes:
\[
  H_1(I,\theta)=\sum_{k\in\Z^N} H_{1k}(I)e^{ik\cdot\theta}.
\]
The \(k=0\) mode is the average. For \(k\neq0\), the homological equation gives
\[
  i(k\cdot\omega_0(I'))S_{1k}(I')=-H_{1k}(I'),
\]
or
\[
  \boxed{
  S_{1k}(I')
  =
  \frac{iH_{1k}(I')}{k\cdot\omega_0(I')}
  \qquad (k\neq0).
  }
\]
The denominators \(k\cdot\omega_0\) are the small divisors.

\begin{definition}[Resonant torus]
The unperturbed torus with action \(I\) is resonant if there exists
\[
  k\in\Z^N\setminus\{0\}
\]
such that
\[
  k\cdot\omega_0(I)=0.
\]
On such a torus the phase \(k\cdot\theta\) is constant along the unperturbed
motion, so the corresponding perturbation does not average away.
\end{definition}

Even if no divisor is exactly zero, the Fourier series for \(S_1\) can diverge
if some \(k\cdot\omega_0\) are too small too often. This is the analytic heart
of KAM theory.

\section{Fourier Decay And The Diophantine Condition}
\label{sec:diophantine-condition}

Small divisors can be overcome when the numerators decay rapidly. If
\(H_1(I,\theta)\) is real analytic in \(\theta\), it extends to a complex strip
\[
  |\operatorname{Im}\theta^i|<\sigma.
\]
Shifting the Fourier integral into the complex strip gives the estimate
\[
  |H_{1k}(I)|\leq C(I)e^{-\sigma |k|_1},
  \qquad
  |k|_1=|k_1|+\cdots+|k_N|.
\]
Thus large Fourier modes are exponentially suppressed.
The estimate is the analytic reason KAM is possible: small divisors grow at
worst polynomially under a Diophantine condition, while analytic Fourier
coefficients shrink exponentially. Smooth but nonanalytic perturbations require
more delicate bookkeeping because the numerator may decay only as a power of
\(|k|\).

\begin{example}[Fourier decay distinguishes analytic and nonanalytic functions]
One-angle examples make the difficulty visible. The series
\[
  \sum_{k=1}^{\infty}\frac{\sin k\theta}{k}
\]
has Fourier coefficients that decay only like \(1/k\); its graph has a jump and
is not real analytic. Likewise
\[
  \sum_{k=1}^{\infty}\frac{\cos k\theta}{k^2}
  =
  \frac14(\theta-\pi)^2-\frac{\pi^2}{12},
  \qquad |\theta|<2\pi,
\]
is periodic but not analytic at the endpoints of a period. A genuinely
real-analytic example is
\[
  \sum_{k=-\infty}^{\infty}\cos(k\theta)e^{-k^2/2},
\]
a theta-function-type series with rapidly suppressed Fourier coefficients.
\end{example}

The frequency vector must also avoid being too well approximated by resonances.

\begin{definition}[Diophantine frequency]
A vector \(\omega\in\R^N\) is Diophantine with constants \(\alpha>0\) and
\(\tau>N-1\) if
\[
  |k\cdot\omega|
  \geq
  \frac{\alpha}{|k|_1^\tau}
  \qquad
  \text{for all } k\in\Z^N\setminus\{0\}.
\]
\end{definition}

For Diophantine \(\omega_0(I')\),
\[
  |S_{1k}(I')|
  \leq
  \frac{C(I')}{\alpha}|k|_1^\tau e^{-\sigma |k|_1},
\]
so the first-order generating function is well-defined. The polynomial factor
from the small divisor cannot beat the exponential decay from analyticity.

\begin{example}[How irrational is irrational?]
For \(N=2\), the Diophantine inequality can be recast as a rational
approximation condition for
\[
  x=\frac{\omega_1}{\omega_2}.
\]
It asks, roughly, that
\[
  \left|x-\frac{p}{q}\right|
  \geq
  \frac{A}{q^{\tau+1}}
\]
for all rational approximants \(p/q\). For \(x=\sqrt2\), use
\[
  f(x)=x^2-2=0.
\]
If \(p/q=x-\eta\), then
\[
  f(p/q)=\frac{p^2-2q^2}{q^2}.
\]
The numerator is a nonzero integer unless \(p/q=\sqrt2\), so its absolute value
is at least one. Linearizing
\[
  f(x-\eta)=f(x)-\eta f'(x)+O(\eta^2)
\]
with \(f'(x)=2\sqrt2\) gives
\[
  |\eta|\geq \frac{A}{q^2}
\]
for some \(A>0\). Thus \(\sqrt2\) satisfies a Diophantine-type lower bound.
More generally, algebraic irrational numbers obey polynomial rational-
approximation bounds. By contrast, numbers such as
\[
  x=\sum_{n=0}^{\infty}\frac{1}{2^{n!}}
\]
are too well approximated by rationals and fail every fixed lower bound of this
kind.
\end{example}

The Diophantine condition excludes a dense set of resonant hyperplanes, but the
excluded set has small measure when \(\alpha\) is small. To see the estimate,
fix a bounded frequency domain \(\Omega\subset\R^N\). For a given
\(k\neq0\), the bad set is the slab
\[
  R_k
  =
  \left\{
    \omega\in\Omega:
    |k\cdot\omega|<
    \frac{\alpha}{|k|_1^\tau}
  \right\}.
\]
Its thickness normal to the hyperplane \(k\cdot\omega=0\) is of order
\[
  \frac{\alpha}{|k|_1^{\tau+1}},
\]
because the normal vector has length comparable to \(|k|_1\). Hence
\[
  \operatorname{Vol}(R_k\cap\Omega)
  \leq
  C_\Omega\frac{\alpha}{|k|_1^{\tau+1}}.
\]
Summing over \(k\neq0\) converges when \(\tau+1>N\), i.e. \(\tau>N-1\). Thus
the non-Diophantine set can be made to have arbitrarily small measure by taking
\(\alpha\) small, even though resonances are dense.

\section{KAM Theorem: Statement And Sketch}
\label{sec:kam-theorem-sketch}

At this point the perturbation problem has two competing facts. Analyticity
makes high Fourier modes small, which helps the perturbation series. Small
divisors make near-resonant modes large, which threatens the perturbation
series. KAM theory is the result of balancing these facts: it restricts
attention to frequencies that are irrational in a quantified way, and it asks
the unperturbed Hamiltonian to twist enough that frequency can be adjusted when
the torus is deformed.

The theorem is best read as a statement about labelled invariant tori, not
about individual trajectories. An unperturbed action value \(I\) determines a
frequency \(\omega_0(I)=\partial H_0/\partial I\). If that frequency is
Diophantine and the twist condition below holds, then the perturbed system has
a nearby torus carrying quasiperiodic motion with the corresponding frequency.
Points on resonant or nearly resonant tori are precisely the points for which
this labelling argument breaks down.

The theorem also has a negative pedagogical message. A small perturbation does
not give one uniform conclusion for all initial conditions. It produces a
mixed phase space: most sufficiently irrational tori survive for small
\(\eps\), low-order resonances open into island chains, separatrices split,
and gaps between surviving tori become possible transport channels. The rest
of the chapter is organized around this mixed picture.

\begin{assumption}[Kolmogorov nondegeneracy]
The integrable Hamiltonian \(H_0(I)\) has a locally nondegenerate frequency map:
\[
  \det\left(\frac{\partial^2H_0}{\partial I_i\partial I_j}\right)\neq0.
\]
Equivalently, nearby actions give genuinely different frequency vectors.
\end{assumption}

\begin{proposition}[KAM theorem, lecture-note form]
Let
\[
  H(I,\theta)=H_0(I)+\eps H_1(I,\theta)
\]
be a sufficiently smooth, typically analytic, nearly integrable Hamiltonian on
\(U\times\mathbb T^N\). Assume \(H_0\) is Kolmogorov nondegenerate. For
sufficiently small \(|\eps|\), every unperturbed torus whose frequency is
Diophantine with fixed constants \((\alpha,\tau)\) persists as a nearby
invariant torus of the perturbed system, provided \(\eps\) is small enough
relative to \(\alpha\). The perturbed tori fill phase space up to a set whose
measure tends to zero as \(\eps\to0\).
\end{proposition}

The proof is not ordinary first-order perturbation theory. The first-order
calculation above shows the mechanism, but higher orders produce new small
divisors and require control of the coordinate change itself:
\[
  I=I'+\eps\frac{\partial S_1}{\partial\theta}+\cdots,
  \qquad
  \theta'=\theta+\eps\frac{\partial S_1}{\partial I'}+\cdots.
\]
The important scale warning is:
\[
  \frac{\partial S_1}{\partial I'},\
  \frac{\partial S_1}{\partial \theta}
  =
  O(\alpha^{-1}),
\]
so the coordinate change is small only when \(\eps/\alpha\) is small. A common
lecture-note form is therefore: for \(|\eps|<\delta\alpha^2\), the tori whose
frequencies satisfy the Diophantine inequality with constant \(\alpha\) persist
and fill phase space up to \(O(\alpha)\) measure.

The point of the exponent and smallness conditions is bookkeeping, not
mysticism. If \(H_1\) is analytic in a strip of width \(r\), its Fourier
coefficients satisfy a bound of the form
\[
  |H_{1,k}|\leq C e^{-r|k|}.
\]
Solving the homological equation divides by \(k\cdot\omega\), and the
Diophantine condition gives
\[
  \left|\frac{H_{1,k}}{k\cdot\omega}\right|
  \leq
  \frac{C}{\alpha}|k|^\tau e^{-r|k|}.
\]
The polynomial factor \(|k|^\tau\) is harmless compared with exponential
decay, provided one gives up a little analyticity width. This is the analytic
heart of the proof: small divisors are dangerous, but not fatal, on the
Diophantine set.

KAM uses a Newton-style iteration. At each step, one solves a homological
equation like the first-order equation, removes most of the angle dependence,
and shrinks the analytic domain to pay for derivative losses. The error is made
quadratically smaller at each step. Diophantine estimates keep the small
divisors under control; nondegeneracy lets one adjust the action label so the
desired frequency remains fixed.

In proof-outline form, the iteration has four repeating moves:
\[
\begin{array}{rcl}
\text{linearize} &:& \text{separate resonant size from nonresonant modes},\\
\text{solve} &:& \text{invert } \omega\cdot\partial_\theta
    \text{ using the Diophantine bound},\\
\text{renormalize} &:& \text{change the action label using nondegeneracy},\\
\text{shrink} &:& \text{reduce the analytic domain so derivatives stay controlled}.
\end{array}
\]
The theorem is difficult because all four moves must be made quantitatively at
once. The conceptual message is simpler: sufficiently irrational frequencies
avoid the worst resonances, while twist lets the torus move to the nearby
frequency that the perturbed system actually supports.

One useful way to read the theorem is as a filtered survival statement. Start
with all invariant tori of the integrable system. Remove narrow neighborhoods
of the low-order resonance surfaces \(k\cdot\omega=0\). The remaining set is
not an interval of actions but a Cantor-like family: it is full of holes at all
orders, yet the total measure of the holes can be made small. The KAM iteration
then proves that each torus in this filtered family has a nearby perturbed
torus with conjugate quasiperiodic motion.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1.0, >=Latex, font=\small]
  \draw[->] (-0.2,0) -- (7.2,0) node[right] {action label \(I\)};
  \draw[thick, blue!70!black] (0.4,0) -- (1.25,0);
  \draw[thick, blue!70!black] (1.55,0) -- (2.35,0);
  \draw[thick, blue!70!black] (2.62,0) -- (3.55,0);
  \draw[thick, blue!70!black] (3.85,0) -- (4.18,0);
  \draw[thick, blue!70!black] (4.52,0) -- (5.42,0);
  \draw[thick, blue!70!black] (5.78,0) -- (6.65,0);
  \foreach \x/\w in {1.4/0.18,2.48/0.10,3.7/0.18,4.34/0.16,5.60/0.18}
  {
    \draw[red!70!black, very thick] (\x-\w,0.18) -- (\x+\w,-0.18);
    \draw[red!70!black, very thick] (\x-\w,-0.18) -- (\x+\w,0.18);
  }
  \node[blue!70!black, align=center, fill=white, inner sep=2pt] at (2.0,0.85)
    {Diophantine\\surviving labels};
  \node[red!70!black, align=center, fill=white, inner sep=2pt] at (5.30,0.90)
    {resonant gaps\\and chaotic layers};
  \draw[decorate, decoration={brace, amplitude=4pt}, thick]
    (0.4,-0.52) -- (6.65,-0.52)
    node[midway, below=10pt, fill=white, inner sep=1.5pt]
    {large measure remains when \(\eps\) is small};
\end{tikzpicture}
\caption{KAM survival is Cantor-like. Resonant action labels are removed in
thin layers around \(k\cdot\omega=0\); the surviving Diophantine labels are
nowhere all of the line, but their total measure approaches the integrable
measure as \(\eps\to0\).}
\label{fig:kam-cantor-survival}
\end{figure}

The theorem does not say that every torus survives. Resonant tori are destroyed
or reorganized into lower-dimensional structures. Nor does it say that all
surviving tori are easy to see numerically. In systems with \(N=2\), surviving
one-dimensional invariant curves on a Poincare section can block transport. In
higher-dimensional systems, the geometry leaves more room for slow drift
through resonance channels.

\section{Breakdown Of Invariant Tori}
\label{sec:breakdown-invariant-tori}

KAM theory is a survival theorem, but a mechanics course also has to explain
what the theorem deliberately leaves unprotected. A surviving irrational torus
is a barrier in a two-dimensional section; a resonant torus is instead a place
where perturbations can accumulate coherently. The route from regular motion
to chaotic transport begins by studying these unprotected resonant tori one by
one, then asking when neighboring resonances overlap and open a connected
transport channel.

The tori most exposed to perturbation are those whose frequencies are
rationally related. On such a torus there exists a nonzero integer vector
\(k\in\Z^N\) with
\[
  k\cdot\omega_0(I_*)=0,
\]
where \(I_*\) is the resonant action and \(\omega_0(I)=\partial H_0/\partial I\)
is the unperturbed frequency. The Fourier mode \(e^{ik\cdot\theta}\) is no
longer rapidly oscillating on that torus; it is phase locked. Averaging does
not remove it. This is the first and most important distinction:
nonresonant terms average away to leading order, while resonant terms build a
new slow dynamics.

\subsection{What It Means For A Torus To Be A Barrier}
\label{subsec:torus-barrier}

The word ``breakdown'' can mean several related things, so we state the
geometric setting carefully.

\begin{assumption}[Two-degree-of-freedom reduction]
When discussing transport barriers, assume either:
\begin{enumerate}
\item a Hamiltonian flow with two degrees of freedom, restricted to one fixed
energy surface and viewed on a transverse Poincare section; or
\item an area-preserving twist map of the annulus
\[
  A=\mathbb T\times I_{\mathrm{ann}},
  \qquad
  (\theta,J)\in\mathbb T\times I_{\mathrm{ann}},
\]
where \(I_{\mathrm{ann}}\subset\R\) is an interval.
\end{enumerate}
In either case the section is two-dimensional, and a surviving invariant torus
appears as a one-dimensional invariant curve.
\end{assumption}

An invariant curve on the annulus is a set
\[
  \Gamma=\{(\theta,J_\Gamma(\theta)):\theta\in\mathbb T\}
\]
that is mapped to itself. If an orbit starts below \(\Gamma\), it cannot later
appear above \(\Gamma\), because crossing would contradict uniqueness and
invariance. Thus, in a two-dimensional Poincare section, invariant curves are
literal transport barriers.

This codimension-one fact is special. For a Hamiltonian with \(N\) degrees of
freedom, a KAM torus has dimension \(N\), while a fixed energy surface has
dimension \(2N-1\). The codimension inside the energy surface is therefore
\[
  (2N-1)-N=N-1.
\]
For \(N=2\), this is codimension one and the torus can separate the energy
surface. For \(N\geq3\), the codimension is at least two, so a single torus
does not divide the energy surface into disconnected regions. This is why
Arnold diffusion is possible in higher-dimensional nearly integrable systems:
motion can drift through connected resonance channels even when many KAM tori
survive.

\subsection{Resonant Normal Form And Island Width}
\label{subsec:resonant-normal-form-island-width}

Now consider a nearly integrable Hamiltonian
\[
  H(I,\theta)=H_0(I)+\eps H_1(I,\theta),
  \qquad
  H_1(I,\theta)=\sum_{m\in\Z^N}H_{1m}(I)e^{im\cdot\theta}.
\]
Let \(I_*\) satisfy the primitive resonance condition
\[
  k\cdot\omega_0(I_*)=0,
  \qquad
  k\in\Z^N,\quad \gcd(k_1,\ldots,k_N)=1.
\]
Here primitive means that the resonance has not been written as an integer
multiple of a lower-order resonance.

Near this resonance, choose a canonical linear change of angle-action
variables in which
\[
  \psi=k\cdot\theta
\]
is one of the new angles. Let \(J\) be the action conjugate to \(\psi\), and
let \(J_*=J(I_*)\). The remaining angles are fast away from secondary
resonances. Averaging over those fast angles leaves, to leading order,
\[
  H_{\mathrm{res}}(\Delta J,\psi)
  =
  \frac12 a_k(\Delta J)^2
  +
  \eps |V_k|\cos(\psi-\phi_k),
  \qquad
  \Delta J=J-J_*.
\]
All symbols in this expression are local data at the resonance:
\begin{itemize}
\item \(\psi\) is the slow resonant phase;
\item \(\Delta J\) is the displacement of the resonant action;
\item \(a_k\neq0\) is the effective twist, obtained by restricting the Hessian
\(\partial_I^2H_0(I_*)\) to the resonant direction;
\item \(V_k\) is the resonant Fourier coefficient of \(H_1\), after the same
canonical change and fast-angle averaging;
\item \(\phi_k\) is the phase of \(V_k\).
\end{itemize}

This is the Hamiltonian of a pendulum. Its equations are
\[
  \dot\psi=a_k\Delta J,
  \qquad
  \dot{\Delta J}
  =
  \eps |V_k|\sin(\psi-\phi_k).
\]
Equilibria of this pendulum are the centers and saddles of a resonant island
chain. The separatrix of the pendulum gives a first estimate of the width of
the resonance.

To compute that width, shift \(\psi\) so that the stable center is at
\(\psi=0\), and replace \(a_k\) by \(|a_k|\) if necessary by reversing
\(\Delta J\). Up to an irrelevant additive constant, write
\[
  H_{\mathrm{pend}}
  =
  \frac12 |a_k|(\Delta J)^2
  -
  \eps |V_k|\cos\psi .
\]
The center has energy \(-\eps |V_k|\), and the saddle at \(\psi=\pi\) has
energy \(+\eps |V_k|\). On the separatrix, the largest value of
\(|\Delta J|\) occurs at \(\psi=0\). Therefore
\[
  \frac12 |a_k|(\Delta J_{\max})^2-\eps |V_k|
  =
  +\eps |V_k|.
\]
Solving gives the resonance half-width
\[
  \boxed{
  \Delta J_{\max}
  =
  2\sqrt{\frac{\eps |V_k|}{|a_k|}}.
  }
\]
The full pendulum width in the resonant action is twice this value. The
important scaling is \(\Delta J_{\max}=O(\sqrt{\eps})\), not \(O(\eps)\). A
small perturbation can therefore carve visible resonant layers.

\subsection{Chirikov Overlap}
\label{subsec:chirikov-overlap}

Suppose two resonances, labelled by \(k\) and \(\ell\), occur at resonant action
values \(J_k\) and \(J_\ell\) along the same local action direction. Let
\(\Delta J_k\) and \(\Delta J_\ell\) denote their half-widths, estimated by the
pendulum calculation above. The Chirikov overlap criterion says that large
chaotic layers should be expected when
\[
  \boxed{
  \Delta J_k+\Delta J_\ell
  \gtrsim |J_k-J_\ell|.
  }
\]
This is not a theorem of the same kind as KAM. It is a physically powerful
criterion for when the separatrix layers of neighboring resonances can no
longer be treated independently. Before overlap, islands may be surrounded by
surviving invariant curves. After overlap, hyperbolic manifolds from different
resonances intersect and produce a connected chaotic sea.

The criterion also explains why low-order resonances dominate pictures. A
Fourier coefficient associated with a large vector \(k\) is usually small in an
analytic Hamiltonian, while the resonance width depends on
\(\sqrt{|V_k|}\). Low-order Fourier modes tend to have the largest islands, so
they create the first visible transport channels.

\subsection{Poincare-Birkhoff Splitting Of A Rational Curve}
\label{subsec:poincare-birkhoff-splitting}

The pendulum normal form explains the local shape of one resonant layer. The
Poincare-Birkhoff theorem explains why rational invariant curves of
area-preserving twist maps do not usually survive as smooth invariant curves.

\begin{proposition}[Poincare-Birkhoff theorem, informal form]
Let \(F\) be an area-preserving twist map of an annulus. If a rational
invariant curve of rotation number \(p/q\) is destroyed by a perturbation while
nearby boundary curves twist in opposite directions under \(F^q\), then
\(F\) has at least two periodic orbits of period \(q\) in the resonant annulus.
Generically these periodic points alternate between elliptic and hyperbolic
type around the annulus.
\end{proposition}

The hypothesis about opposite twist is the section-map version of being on
opposite sides of a resonance. Elliptic points become island centers. Hyperbolic
points have stable and unstable manifolds. If those manifolds coincide, they
form a separatrix. If they split and intersect transversely, the separatrix is
replaced by a homoclinic tangle.

This theorem is the precise mathematical reason a rational torus usually turns
into an island chain. The next section derives this mechanism directly for a
Poincare map, including the area-preservation argument that forces fixed
points of the resonant iterate.

\subsection{Separatrix Splitting And Homoclinic Tangles}
\label{subsec:separatrix-splitting-homoclinic-tangles}

A resonant pendulum has a separatrix because the stable and unstable manifolds
of the saddle coincide. This coincidence is fragile. Under a perturbation,
there are usually two different curves:
\[
  \begin{aligned}
  W^s&=\text{stable manifold of the hyperbolic orbit},\\
  W^u&=\text{unstable manifold of the hyperbolic orbit}.
  \end{aligned}
\]
If \(W^s\) and \(W^u\) cross transversely once, area preservation forces an
infinite sequence of further intersections. The resulting homoclinic tangle is
the geometric core of deterministic chaos: strips of phase space are stretched,
folded, and reinserted near the saddle.

One first-order way to measure separatrix splitting is the Melnikov function.
Assume a one-degree-of-freedom separatrix Hamiltonian \(H_{\mathrm{sep}}(z)\),
with separatrix orbit \(z_0(t)\), is perturbed periodically:
\[
  H(z,t)=H_{\mathrm{sep}}(z)+\eps H_{\mathrm{pert}}(z,t).
\]
The phase \(t_0\) records where the perturbation is in its period when the
orbit passes near the saddle. The leading signed distance between \(W^u\) and
\(W^s\) is proportional to
\[
  M(t_0)
  =
  \int_{-\infty}^{\infty}
  \Poisson{H_{\mathrm{sep}}}{H_{\mathrm{pert}}}
  \bigl(z_0(t-t_0),t\bigr)\,\dd t .
\]
If \(M(t_0)\) has a simple zero, then the two manifolds intersect
transversely for sufficiently small nonzero \(\eps\). Thus a single computable
integral can diagnose the birth of a chaotic layer near a separatrix.

\subsection{Cantori And Partial Barriers}
\label{subsec:cantori-partial-barriers}

For area-preserving maps, the strongest irrational invariant curves are often
those whose rotation numbers are hardest to approximate by rationals. The
golden-mean rotation number
\[
  \rho_g=\frac{\sqrt5-1}{2}
\]
is the standard example because its continued fraction is
\([1;1,1,1,\ldots]\), making it maximally poorly approximated among
irrationals in the usual sense.

When an irrational invariant curve finally breaks, it need not disappear
without trace. It can leave behind a Cantor-like invariant set called a
cantorus. A cantorus is not a continuous barrier. It has gaps. However, these
gaps can be small, so the cantorus acts as a partial barrier: transport occurs
through turnstile lobes, but the flux can be very slow.

For the standard map, a famous numerical benchmark is that the last invariant
curve with golden-mean rotation breaks near
\[
  K_c\approx0.971635.
\]
This number is obtained by numerical criteria such as Greene's residue
criterion, not by simply inspecting a finite orbit. If a rational periodic
orbit has period \(q\), the residue of the \(q\)-fold derivative is
\[
  R=\frac{2-\tr(DT_K^q)}{4}.
\]
For an elliptic periodic point \(0<R<1\); for a hyperbolic point \(R<0\) or
\(R>1\). Greene's idea is to approximate an irrational rotation number by
rationals \(p_n/q_n\) and watch the stability of the corresponding periodic
orbits as \(n\) grows. Bounded, critical behavior suggests a surviving
irrational curve; runaway hyperbolicity suggests breakdown.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=0.95, >=Latex, font=\small]
  \begin{scope}[shift={(0,0)}]
    \draw[->] (-0.2,0) -- (2.7,0) node[right] {\(\theta\)};
    \draw[->] (0,-0.15) -- (0,2.2) node[above] {\(J\)};
    \foreach \y/\amp in {0.45/0.04,0.85/0.06,1.25/0.05,1.65/0.04}
      \draw[blue!70!black, thick, domain=0.15:2.45, samples=80]
        plot (\x,{\y+\amp*sin(220*\x)});
    \draw[red!75!black, thick, ->] (0.55,0.63) -- (1.45,0.63);
    \node[align=center, text width=2.8cm] at (1.35,-0.70)
      {surviving invariant curves block crossing};
  \end{scope}

  \begin{scope}[shift={(4.0,0)}]
    \draw[->] (-0.2,0) -- (2.7,0) node[right] {\(\theta\)};
    \draw[->] (0,-0.15) -- (0,2.2) node[above] {\(J\)};
    \foreach \x in {0.55,1.35,2.15}
    {
      \draw[orange!85!black, thick] (\x,1.05) ellipse (0.28 and 0.17);
      \fill[orange!85!black] (\x,1.05) circle (1.1pt);
    }
    \foreach \x in {0.95,1.75}
    {
      \draw[purple!75!black, thick] (\x-0.10,0.95) -- (\x+0.10,1.15);
      \draw[purple!75!black, thick] (\x-0.10,1.15) -- (\x+0.10,0.95);
    }
    \draw[purple!65!black, densely dashed]
      (0.10,1.05) .. controls (0.55,1.55) and (1.05,1.55) .. (1.35,1.23)
      .. controls (1.70,1.55) and (2.20,1.55) .. (2.55,1.05);
    \draw[purple!65!black, densely dashed]
      (0.10,1.05) .. controls (0.55,0.55) and (1.05,0.55) .. (1.35,0.87)
      .. controls (1.70,0.55) and (2.20,0.55) .. (2.55,1.05);
    \node[align=center, text width=2.8cm] at (1.35,-0.70)
      {rational curve becomes islands and saddles};
  \end{scope}

  \begin{scope}[shift={(8.0,0)}]
    \draw[->] (-0.2,0) -- (2.7,0) node[right] {\(\theta\)};
    \draw[->] (0,-0.15) -- (0,2.2) node[above] {\(J\)};
    \draw[blue!70!black, thick, domain=0.15:0.65, samples=30]
      plot (\x,{1.05+0.05*sin(240*\x)});
    \draw[blue!70!black, thick, domain=0.86:1.36, samples=30]
      plot (\x,{1.05+0.05*sin(240*\x)});
    \draw[blue!70!black, thick, domain=1.58:2.28, samples=40]
      plot (\x,{1.05+0.05*sin(240*\x)});
    \draw[red!75!black, thick, ->] (0.72,0.70) .. controls (0.83,0.90) .. (0.95,1.04);
    \draw[red!75!black, thick, ->] (1.42,1.36) .. controls (1.52,1.19) .. (1.64,1.07);
    \foreach \x/\y in {0.40/0.55,0.75/1.55,1.20/0.75,1.75/1.42,2.25/0.62}
      \fill[black!55] (\x,\y) circle (0.9pt);
    \node[align=center, text width=2.8cm] at (1.35,-0.70)
      {cantorus leaves gaps and slow turnstiles};
  \end{scope}
\end{tikzpicture}
\caption{Three stages in the loss of transport barriers on a two-dimensional
section. Left: invariant curves separate the annulus. Middle: a rational curve
splits into an island chain with hyperbolic points. Right: an irrational curve
may leave a cantorus, which permits transport only through gaps.}
\label{fig:torus-breakdown-stages}
\end{figure}

The accompanying Python script
\path{demos/python/standard_map_torus_breakdown.py} uses the standard map on
the cylinder. It initializes an ensemble near the golden-mean unperturbed
rotation number, leaves \(p\) unwrapped, and measures finite-time momentum
spreading. A large spread is evidence of transport through broken barriers; a
small spread over finite time is not a proof of survival. The corresponding
Makefile target is \path{standard-map-breakdown-demo}, and the Wolfram
Language companion is
\path{demos/mathematica/StandardMapTorusBreakdown.wl}. This distinction is
essential: numerical trajectories illustrate the geometry, while KAM,
Poincare-Birkhoff theory, and separatrix-splitting criteria explain why the
geometry is there.

\section{From Resonance To Chaos: Poincare Map Picture}
\label{sec:poincare-resonance-chaos}

A Poincare section converts continuous Hamiltonian motion into an
area-preserving map. This loses the time parametrization but keeps the geometry
needed to see resonance. Near a resonant torus, some iterate of the unperturbed
map is the identity; after perturbation, that near-identity map unfolds into
fixed points, island chains, and separatrices.

This is why maps are so useful in a chapter about flows. A continuous orbit can
hide its organization in time; a section records each return to a transverse
surface and turns invariant tori into curves, resonances into island chains,
and chaotic layers into visible scatter. The standard map below is not a toy
because it is simple; it is useful because it isolates this return-map
geometry.

For two degrees of freedom, begin with an integrable Hamiltonian
\[
  H_0=H_0(I_1,I_2),
  \qquad
  (I_1,I_2,\theta_1,\theta_2)\in U\times\mathbb T^2,
\]
and fix an energy surface \(H_0=E\). Choose a Poincare section \(S\), for
example \(\theta_2=0\). Starting from \(x\in S\), follow the orbit until it
next intersects \(S\). This defines the return map
\[
  T:S\to S.
\]
The unperturbed flow has
\[
  \dot\theta_1=\omega_1(I),
  \qquad
  \dot\theta_2=\omega_2(I).
\]
Returning to \(\theta_2=0\) takes time \(2\pi/\omega_2\), so the return map
twists the remaining angle by
\[
  \Delta\theta_1=2\pi\frac{\omega_1}{\omega_2}.
\]
On a resonant torus with \(\omega_1/\omega_2=p/q\),
\[
  T^q=\id
\]
for the unperturbed return map \(T\). A concrete period-three resonance is
\[
  \frac{\omega_1}{\omega_2}=\frac13,
  \qquad
  T:\theta_1\mapsto\theta_1+\frac{2\pi}{3},
  \qquad
  T^3=\id.
\]
Let \(C\subset S\) be the invariant curve obtained by intersecting this
resonant torus with the section. Nearby invariant curves \(C_-\) and \(C_+\)
have slightly different rotation numbers. With the usual twist assumption,
\(T^3\) rotates points one way on \(C_-\) and the opposite way on \(C_+\).

Now perturb the Hamiltonian:
\[
  H=H_0+\eps H_1.
\]
The perturbed Poincare map \(T_\eps\) is still defined by first return to
\(S\), even though the invariant tori may no longer exist. Near
\(C\cap S\), \(T_\eps^3\) is close to the identity. At fixed energy it has the
local form
\[
  T_\eps^3(I_1,\theta_1)
  =
  \left(
    I_1+f_\eps(I_1,\theta_1),
    \theta_1+g_\eps(I_1,\theta_1)
  \right).
\]
For small \(\eps\), the opposite twists on \(C_-\) and \(C_+\) persist.
Therefore an intermediate curve \(C'\) has
\[
  g_\eps(I_1,\theta_1)=0;
\]
on \(C'\), the third-return map produces no angular displacement.

The missing radial displacement is controlled by area preservation. Hamiltonian
flow preserves \(\omega\), and hence the Poincare map preserves the induced
area form on the section. Equivalently, for a closed curve \(\gamma\) bounding
a disk \(D_\gamma\) in the section,
\[
  \oint_\gamma p\,\dd q
  =
  -\int_{D_\gamma}\omega,
\]
so the signed symplectic area enclosed by \(\gamma\) is invariant. If
\(f_\eps\) were nonzero with a fixed sign everywhere on \(C'\), the curve
would be displaced monotonically inward or outward under \(T_\eps^3\), changing
the enclosed area. Thus \(f_\eps\) must have at least a pair of zeros on
\(C'\). At such points
\[
  f_\eps=0,
  \qquad
  g_\eps=0,
\]
so \(T_\eps^3\) has fixed points. In the period-three example, if \(x\) is a
fixed point of \(T_\eps^3\), then \(T_\eps(x)\) is also a fixed point of
\(T_\eps^3\). Thus the resonant curve generically produces six fixed points:
three elliptic and three hyperbolic, alternating around the resonant layer.

Elliptic points are surrounded by island chains and new local invariant curves.
Hyperbolic points have stable and unstable manifolds. Once those manifolds
intersect transversely, the separatrix is replaced by a chaotic layer. Farther
from the resonance, sufficiently Diophantine KAM curves persist and confine
motion.

This is the local mechanism behind the pictures one sees in the standard map,
the driven pendulum, and the asteroid-belt resonances: KAM tori survive away
from strong resonances, while resonant tori break into islands and chaotic
separatrix webs.

\section{The Standard Map As A Minimal Model}
\label{sec:standard-map}

The standard map is
\[
\begin{aligned}
  p_{n+1} &= p_n+K\sin q_n,\\
  q_{n+1} &= q_n+p_{n+1}\pmod{2\pi}.
\end{aligned}
\]
This is naturally a map of the cylinder, with \(q\) taken modulo \(2\pi\) and
\(p\in\R\). For phase portraits on a compact square, the Python demo reduces
both \(q\) and \(p\) modulo \(2\pi\), giving the corresponding map on the
two-torus.
It is area-preserving because the Jacobian matrix has determinant one:
\[
  D\Psi=
  \begin{pmatrix}
  1+K\cos q_n & 1\\
  K\cos q_n & 1
  \end{pmatrix},
  \qquad
  \det D\Psi=1.
\]
The parameter \(K\) controls resonance overlap. For small \(K\), invariant
curves dominate. For larger \(K\), chaotic regions grow.
The compact two-torus portrait is useful for seeing islands, but it hides
whether an orbit has crossed many momentum cells on the cylinder. For transport
questions one should keep \(p\) unwrapped; this is the convention used in the
torus-breakdown diagnostic demo.

\begin{example}[Period-three resonance in the standard map]
A common drift-kick convention for the standard map is
\[
  T_K(\theta,p)=\left(\theta+p,\ p+K\sin(\theta+p)\right),
\]
which is symplectically equivalent to the convention above after shifting where
the kick is placed. At \(K=0\), the invariant curve
\[
  C:\quad p=\frac{2\pi}{3}
\]
has
\[
  T_0^3=\id
\]
because the angle advances by \(2\pi\) after three iterates.

Let
\[
  p=\frac{2\pi}{3}+\delta(\theta),
  \qquad
  \delta=O(K).
\]
To first order in \(K\) and \(\delta\),
\[
\begin{aligned}
  \theta_3-\theta-2\pi
  &=
  3\delta
  +
  2K\sin\left(\theta+\frac{2\pi}{3}\right)
  +
  K\sin\left(\theta+\frac{4\pi}{3}\right)
  +O(K^2).
\end{aligned}
\]
The nearby curve \(C'\) on which \(T_K^3\) fixes the angle therefore satisfies
\[
  \boxed{
  \delta(\theta)
  =
  \frac{K}{2}\sin\theta
  -
  \frac{K\sqrt3}{6}\cos\theta
  +
  O(K^2).
  }
\]
For example, at \(\theta=\pi/3\),
\[
  p=\frac{2\pi}{3}+\frac{K}{2\sqrt3}+O(K^2),
\]
which is the initial condition singled out in the corresponding numerical
exercise. At the next order, area preservation forces the radial displacement
on \(C'\) to have alternating zeros. These become the six period-three points:
three elliptic island centers and three hyperbolic points with separatrix
structure.
\end{example}

\begin{figure}[htbp]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=0.86\textwidth,
  height=0.64\textwidth,
  xmin=0, xmax=6.283185,
  ymin=0, ymax=6.283185,
  axis lines=left,
  axis line style={-Latex},
  xlabel={\(q\)},
  ylabel={\(p\)},
  xtick={0,1.570796,3.141593,4.712389,6.283185},
  xticklabels={\(0\),\(\pi/2\),\(\pi\),\(3\pi/2\),\(2\pi\)},
  ytick={0,2.094395,4.188790,6.283185},
  yticklabels={\(0\),\(2\pi/3\),\(4\pi/3\),\(2\pi\)},
  tick style={black},
  grid=both,
  major grid style={black!10},
  minor tick num=1,
  minor grid style={black!4},
  clip=true,
  legend style={
    at={(0.5,-0.16)},
    anchor=north,
    legend columns=2,
    draw=none,
    fill=white,
    font=\scriptsize,
    /tikz/every even column/.append style={column sep=0.35cm}
  },
]
  \addplot[only marks, mark=*, mark size=0.23pt, blue!45]
    table[x=q,y=p] {data/standard_map_kam_K065.dat};
  \addlegendentry{quasiperiodic curves}
  \addplot[only marks, mark=*, mark size=0.30pt, black!38]
    table[x=q,y=p] {data/standard_map_period3_layer_K065.dat};
  \addlegendentry{separatrix layer}
  \addplot[only marks, mark=*, mark size=0.32pt, red!65!black]
    table[x=q,y=p] {data/standard_map_period3_islands_K065.dat};
  \addlegendentry{period-three islands}
  \addplot[densely dashed, black!45]
    coordinates {(0,2.094395) (6.283185,2.094395)};
  \addplot[only marks, mark=*, mark size=1.45pt, red!75!black]
    coordinates {(0.880179,1.760359) (3.141593,2.261413) (5.403006,2.261413)};
  \addplot[only marks, mark=x, mark size=3.0pt, line width=0.8pt, purple!75!black]
    coordinates {(0.000000,1.888577) (1.888577,1.888577) (4.394609,2.506032)};
  \addlegendentry{hyperbolic points}
  \node[anchor=west, black!55, fill=white, inner sep=1pt] at (axis cs:0.18,2.34)
    {\(p=2\pi/3\)};
\end{axis}
\end{tikzpicture}
\caption{Actual iterates of the standard map on the compact two-torus at
\(K=0.65\). Blue points lie on quasiperiodic invariant curves away from the
resonance, red points trace orbits near the elliptic period-three island chain,
gray points sample the local separatrix layer, and purple crosses mark the
hyperbolic period-three orbit.}
\label{fig:standard-map-island-chain}
\end{figure}

\section{What To Take Forward}
\label{sec:perturbation-chaos-take-forward}

The main logical chain of this chapter is the following. Complete
integrability gives invariant tori carrying quasi-periodic motion. In
action-angle variables an integrable Hamiltonian depends only on the actions,
so the frequencies are gradients of \(H_0(I)\). A small perturbation introduces
angle-dependent terms. Canonical perturbation theory tries to remove those
terms order by order, but the homological equation divides by integer
frequency combinations \(k\cdot\omega(I)\). Small divisors are therefore not a
technical nuisance; they are the obstruction that distinguishes robust
quasi-periodic motion from resonant dynamics.

KAM theory says that sufficiently nonresonant tori with nondegenerate frequency
maps survive small smooth perturbations, though they deform. Resonant tori are
more fragile. Near resonance, the averaged dynamics reduces to pendulum-like
islands, separatrices, and stochastic layers. As perturbations grow, resonance
overlap destroys transport barriers and produces large chaotic regions.

The standard map is the simplest laboratory where these stages can be seen in
one picture: rotational invariant curves at small coupling, island chains
around rational rotation numbers, cantori and partial barriers near the breakup
threshold, and widespread diffusion when barriers are gone.

The three-body problem in Chapter~\ref{chap:three-body-problem} is a principal
physical arena for these ideas. It is not included here because its Newtonian
reduction, restricted limit, Lagrange points, and asteroid-belt ensembles
deserve a self-contained treatment.


\section{Chapter-End Laboratories And Exercises}
\label{sec:perturbation-chaos-chapter-end-labs}

The laboratories below are finite-dimensional tests of the perturbative logic:
slow resonant angles, pendulum normal forms, small divisors, island chains, and
the onset of chaotic regions.

\subsection{Driven One-Angle Hamiltonian And Resonant Normal Form}
\label{sec:detailed-driven-resonance}

Consider the one-angle time-periodic perturbation model
\[
  H(I,\theta,t)
  =
  \frac12I^2+\epsilon\cos(n\theta-\nu t).
\]
The unperturbed frequency is
\[
  \dot\theta=I.
\]
Resonance occurs when the phase
\[
  \psi=n\theta-\nu t
\]
is slow, i.e.
\[
  nI-\nu\approx0.
\]
Thus
\[
  I_{\mathrm{res}}=\frac{\nu}{n}.
\]
To make the system autonomous, introduce the extended phase-space pair
\((t,E_t)\), with Hamiltonian
\[
  K=E_t+\frac12I^2+\epsilon\cos(n\theta-\nu t).
\]
Use the canonical slow angle
\[
  \psi=n\theta-\nu t
\]
and an action \(P\) measuring displacement from resonance. Locally set
\[
  I=\frac{\nu}{n}+P.
\]
Dropping constants and terms independent of the slow dynamics, the resonant
Hamiltonian has the pendulum form
\[
  \boxed{
  K_{\mathrm{res}}(P,\psi)
  =
  \frac12P^2+\epsilon\cos\psi.
  }
\]
Up to scale factors depending on the precise canonical normalization, this is
the universal local model of a first-order resonance. Its phase portrait has
elliptic islands, hyperbolic fixed points, and a separatrix. This is why
planetary mean-motion resonances, the standard map, and the driven pendulum all
share the same local geometry.

\subsection{Henon-Heiles As The Finite-Dimensional Chaos Bridge}
\label{sec:lab-henon-heiles}

The Henon-Heiles Hamiltonian is
\[
  H_{\mathrm{HH}}
  =
  H_0+\lambda_{\mathrm{HH}}H_1,
\]
with
\[
  H_0=
  \frac12(p_x^2+p_y^2+x^2+y^2),
  \qquad
  H_1=x^2y-\frac13y^3.
\]
For the isotropic oscillator, introduce action-angle variables
\[
  x=\sqrt{2I_1}\sin\theta_1,
  \qquad
  p_x=\sqrt{2I_1}\cos\theta_1,
\]
\[
  y=\sqrt{2I_2}\sin\theta_2,
  \qquad
  p_y=\sqrt{2I_2}\cos\theta_2.
\]
Then
\[
  H_0=I_1+I_2,
  \qquad
  \omega_0=(1,1).
\]
The first-order homological equation for a generating function \(S_1\) is
\[
  \omega_0\cdot\partial_\theta S_1
  =
  \langle H_1\rangle-H_1.
\]
In Fourier components,
\[
  i(k_1+k_2)S_{1k}=-H_{1k}.
\]
The denominator vanishes whenever
\[
  k_1+k_2=0.
\]
At first order the cubic perturbation has no average and the dangerous modes
can be handled by symmetry, but at second order the canonical transformation
generates resonant Fourier modes with \(k_1+k_2=0\). Those modes cannot be
removed by a periodic \(S_2\). This is the small-divisor obstruction in its
most concrete classroom form: the unperturbed oscillator has equal
frequencies, so resonant combinations of angles do not average away.

The Henon-Heiles system is therefore the right finite-dimensional bridge from
integrability to Hamiltonian chaos. At low energy, Poincare sections show
deformed invariant curves. At higher energy, resonance islands and chaotic
regions appear. Unlike the asteroid problem, all variables and forces are
visible in a four-dimensional phase space.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1.0, >=Latex, font=\small]
  \begin{scope}
    \draw[->] (-2.1,0) -- (2.1,0) node[right] {\(x\)};
    \draw[->] (0,-1.9) -- (0,2.2) node[above] {\(y\)};
    \foreach \s in {0.45,0.75,1.05,1.35}
    {
      \draw[blue!70!black, thick, domain=0:360, samples=120, variable=\t]
        plot ({\s*cos(\t)*(1+0.12*sin(3*\t))},
              {\s*sin(\t)*(1-0.10*cos(3*\t))});
    }
    \draw[red!75!black, thick]
      (0,1.75) -- (-1.55,-0.95) -- (1.55,-0.95) -- cycle;
    \node[red!75!black, align=center, fill=white, inner sep=1pt] at (0,-1.47)
      {three escape\\valleys};
    \node[fill=white, inner sep=1pt] at (0.15,2.55) {configuration contours};
  \end{scope}
  \begin{scope}[xshift=5.2cm]
    \draw[->] (-1.8,0) -- (1.9,0) node[right] {\(x\)};
    \draw[->] (0,-1.8) -- (0,1.9) node[above] {\(p_x\)};
    \draw[blue!70!black, thick] (0,0) ellipse (1.2 and 0.65);
    \draw[blue!70!black, thick] (0,0) ellipse (0.72 and 0.38);
    \foreach \a in {0,120,240}
      \draw[red!75!black, thick] ({0.82*cos(\a)},{0.82*sin(\a)})
        ellipse (0.28 and 0.13);
    \foreach \x/\y in {-1.3/0.9,-1.1/-0.8,-0.4/1.15,0.25/-1.2,1.0/0.75,1.25/-0.55}
      \fill[black!55] (\x,\y) circle (0.9pt);
    \node[fill=white, inner sep=1pt] at (0.45,2.30) {Poincare-section cartoon};
  \end{scope}
\end{tikzpicture}
\caption{Henon-Heiles laboratory. Configuration-space contours show the
threefold cubic deformation of the oscillator potential; a Poincare section
then records the phase-space transition from invariant curves to islands and
chaotic scatter. Both panels are schematic diagnostics for the derivation.}
\label{fig:lab-henon-heiles}
\end{figure}
\FloatBarrier

\section{Questions For The Reader}
\label{sec:perturbation-chaos-questions}

\begin{enumerate}
\item In the homological equation
\(\omega(I)\cdot\partial_\theta S_1=-(H_1-\langle H_1\rangle)\), identify
exactly where the Diophantine condition is used. What fails if \(\omega\) is
rational?
\item For a one-degree-of-freedom resonant normal form, draw the phase portrait
and identify elliptic fixed points, hyperbolic fixed points, separatrices, and
island width. Which part of this picture persists in an area-preserving map?
\item Use the standard-map figure in this chapter to compare three types of
phase-space organization: invariant curves, resonance islands, and chaotic
layers. Which of these can block global radial transport in action?
\item Explain why Liouville volume preservation is compatible with chaos. Which
property of the dynamics changes when invariant tori break?
\end{enumerate}

\section{Associated Demos}
\label{sec:perturbation-chaos-demos}

The associated standard-map torus-breakdown demos are:
\begin{itemize}
\item \path{demos/python/standard_map_torus_breakdown.py}
\item \path{demos/mathematica/StandardMapTorusBreakdown.wl}
\end{itemize}
They reproduce the qualitative transition from invariant curves to resonance
islands and chaotic layers.  They are intentionally small: the point is to make
the phase portrait transparent enough that it can be compared directly with the
perturbative derivation.
