\chapter{The Three-Body Problem}
\label{chap:three-body-problem}

Two bodies under mutual gravity trace conic sections: ellipses, parabolas,
hyperbolas, and rectilinear collision orbits as degenerate limiting cases.
Three bodies do not admit a comparable closed-form solution, a universal
one-dimensional radial reduction, or generic integrability. What survives is a
rich phase-space geometry: rotating-triangle exact solutions, equilibrium
points in the rotating frame, resonant islands, chaotic scattering, and
long-time transport.

Sundman (1907 announcement, 1912 publication) constructed a convergent series
solution away from triple collision, but the result is not a substitute for
integrability. Binary collisions can be regularized by changing variables and
time scale. In the planar problem, the Levi-Civita change
\[
  z=u^2,\qquad \dd t=|u|^2\,\dd s
\]
turns the inverse-square collision singularity into a regular passage in the
\(u\)-plane: since \(|z|=|u|^2\), the singular potential \(1/|z|\) becomes
\(1/|u|^2\), and multiplication by the Sundman time factor
\(\dd t/\dd s=|u|^2\) cancels the leading pole. Sundman's construction combines
this local regularization with analytic continuation and a global time
reparametrization. Triple collision is the obstruction left out of the global
statement, and Belorizky emphasized in 1930 that Sundman's series converges far
too slowly for practical computation.

This chapter connects every phenomenon back to a controlled reduction of
Newton's law. We first formulate the full Newtonian three-body problem and
reduce translations using Jacobi coordinates. We then derive the circular
restricted three-body problem as a controlled limit and compute its
rotating-frame Hamiltonian, Jacobi constant, zero-velocity curves, and Lagrange
points. Finally we connect that model to the Sun-Jupiter-asteroid demo and to
ejection statistics.

\paragraph{Roadmap.}
The chapter moves through three models in a fixed order. The full three-body
problem supplies the Hamiltonian, the conserved quantities, and the reason
generic integrability fails. The circular restricted three-body problem supplies
the rotating frame, Jacobi constant, zero-velocity curves, and Lagrange points.
The heliocentric Sun-Jupiter-asteroid model supplies the finite-time ensemble
experiment. Each model keeps part of the previous one and discards part of it;
the text names those decisions so that numerical ejection is tied to a specified
model and time scale.

\paragraph{Model hierarchy.}
The reader should keep four levels separate:
\[
\begin{array}{rcl}
\text{two-body Kepler} &:& \text{integrable reference problem},\\
\text{full three-body} &:& \text{Newtonian Hamiltonian with all masses active},\\
\text{CR3BP} &:& \text{two primaries prescribed, one massless test body},\\
\text{asteroid ensemble} &:& \text{finite-time numerical statistic}.
\end{array}
\]
The later levels are controlled reductions chosen to isolate transport, escape
channels, and resonance-driven loss in a teachable setting.

Appendix~\ref{app:three-body-numerics} records the numerical protocol behind
the last two levels: coordinate choices, event definitions, resonance windows,
uncertainty estimates, and convergence checks. The chapter introduces these
choices where they first affect the mechanics; the appendix gathers them in one
place for reproducibility.

\paragraph{Historical context.}
Poincare's prize essay for King Oscar II's 1889 competition was substantially
revised after he discovered an error; Weierstrass, on the prize jury, demanded
a corrected version. The revision exposed homoclinic intersections and became
the foundation of \emph{Methodes nouvelles de la mecanique celeste}
(1892--1899). Birkhoff (1915) extended the program by making return maps and
periodic orbits operational tools of celestial mechanics. Painleve's 1895
singularity theorem clarified a separate issue: in the Newtonian three-body
problem, every finite-time singularity is caused by a collision. Xia's 1992
noncollision singularities for five bodies show that this is a special feature
of low \(N\), not a general safety theorem for Newtonian gravity.

\section{The Newtonian Three-Body Problem}
\label{sec:newtonian-three-body-problem}

The restricted problem will be derived as a limit of the Newtonian three-body
problem; we therefore start with the full problem. The classical conserved
quantities, momentum, angular momentum, and energy, survive, but they no longer
reduce the dynamics to a one-dimensional radial equation. The restricted
problem is not a different force law; it is a controlled limit and coordinate
choice inside the same Newtonian interaction.

Let \(q_i(t)\in\R^3\) be the inertial position of mass
\(m_i>0\), and write
\[
  \mathbf r_{ij}=q_i-q_j,\qquad r_{ij}=\norm{\mathbf r_{ij}}.
\]
The Lagrangian is
\[
  L(q,\dot q)
  =
  \sum_{i=1}^3 \frac12 m_i\norm{\dot q_i}^2
  +
  \sum_{1\leq i<j\leq 3}\frac{Gm_im_j}{r_{ij}},
\]
because \(L=T-V\) and the Newtonian potential energy is negative:
\[
  V(q)=-\sum_{i<j}\frac{Gm_im_j}{r_{ij}}.
\]
The Euler-Lagrange equation for \(q_i\) gives
\[
  m_i\ddot q_i
  =
  \sum_{j\neq i}
  Gm_im_j\frac{q_j-q_i}{\norm{q_j-q_i}^3}.
\]
After dividing by \(m_i\),
\[
  \boxed{
  \ddot q_i
  =
  \sum_{j\neq i}
  Gm_j\frac{q_j-q_i}{\norm{q_j-q_i}^3}.
  }
\]
This is a closed Hamiltonian system with canonical momenta \(p_i=m_i\dot q_i\)
and Hamiltonian
\[
  H(q,p)
  =
  \sum_{i=1}^3\frac{\norm{p_i}^2}{2m_i}
  -
  \sum_{i<j}\frac{Gm_im_j}{r_{ij}}.
\]

The center of mass
\[
  R_{\mathrm{cm}}=\frac{1}{M}\sum_i m_iq_i,
  \qquad
  M=m_1+m_2+m_3,
\]
moves uniformly because the internal gravitational forces cancel in pairs.
The total momentum, angular momentum, and energy are
\[
  P=\sum_i p_i,
  \qquad
  J=\sum_i q_i\times p_i,
  \qquad
  H=\text{constant}.
\]
These conserved quantities are essential, but they do not make the generic
three-body problem integrable. After removing center-of-mass motion, the
planar problem has four configuration degrees of freedom and an
eight-dimensional phase space.  Fixing angular momentum and quotienting the
remaining planar rotation removes two phase dimensions; a fixed-energy reduced
space is still five-dimensional.  This is vastly richer than the one-degree
radial problem left by the two-body reduction, leaving enough room for
separatrices, homoclinic tangles, chaotic scattering, and transport.

\subsection{Translation Reduction And Jacobi Coordinates}
\label{subsec:three-body-jacobi-coordinates}

The center-of-mass degree of freedom is dynamically trivial. Removing it shows
which variables remain in the interacting problem. Let
\[
  m_{12}=m_1+m_2,\qquad M=m_{12}+m_3,
\]
and define Jacobi coordinates
\[
  R=\frac{m_1q_1+m_2q_2+m_3q_3}{M},
  \qquad
  \rho_1=q_2-q_1,
  \qquad
  \rho_2=q_3-\frac{m_1q_1+m_2q_2}{m_{12}}.
\]
Here \(\rho_1\) is the relative vector inside the pair \((1,2)\), while
\(\rho_2\) points from the center of mass of that pair to the third body.  The
inverse transformation is
\[
\begin{aligned}
  q_1&=R-\frac{m_2}{m_{12}}\rho_1-\frac{m_3}{M}\rho_2,\\
  q_2&=R+\frac{m_1}{m_{12}}\rho_1-\frac{m_3}{M}\rho_2,\\
  q_3&=R+\frac{m_{12}}{M}\rho_2 .
\end{aligned}
\]
Substituting these expressions into the kinetic energy and collecting terms
gives the diagonal form
\[
  T=
  \frac12 M\norm{\dot R}^2
  +\frac12\mu_1\norm{\dot\rho_1}^2
  +\frac12\mu_2\norm{\dot\rho_2}^2,
  \qquad
  \mu_1=\frac{m_1m_2}{m_{12}},
  \qquad
  \mu_2=\frac{m_3m_{12}}{M}.
\]
The absence of cross terms is the point of Jacobi coordinates: the mass metric
has been diagonalized after translations are separated.

The attraction potential \(U=-V\) becomes
\[
\begin{aligned}
  U(\rho_1,\rho_2)
  &=
  \frac{Gm_1m_2}{\norm{\rho_1}}
  +\frac{Gm_1m_3}
  {\norm{\rho_2+\frac{m_2}{m_{12}}\rho_1}}
  +\frac{Gm_2m_3}
  {\norm{\rho_2-\frac{m_1}{m_{12}}\rho_1}} .
\end{aligned}
\]
In the center-of-mass frame \(R=0\), the relative Hamiltonian is therefore
\[
  H_{\mathrm{rel}}
  =
  \frac{\norm{\pi_1}^2}{2\mu_1}
  +\frac{\norm{\pi_2}^2}{2\mu_2}
  -U(\rho_1,\rho_2),
\]
where \(\pi_a=\mu_a\dot\rho_a\).  This formula shows what
changed from the two-body problem: there are now two coupled relative vectors,
and each inverse-square attraction depends on a different linear combination of
them.

\begin{derivation}[Lagrange-Jacobi identity]
In the center-of-mass frame define the moment of inertia
\[
  I_{\mathrm{cm}}
  =
  \sum_{i=1}^3 m_i\norm{q_i}^2
  =
  \mu_1\norm{\rho_1}^2+\mu_2\norm{\rho_2}^2 .
\]
Differentiate twice:
\[
  \ddot I_{\mathrm{cm}}
  =
  4T+2\sum_i q_i\cdot m_i\ddot q_i .
\]
Since \(m_i\ddot q_i=\partial U/\partial q_i\) and \(U\) is homogeneous of
degree \(-1\), Euler's theorem gives
\[
  \sum_i q_i\cdot\frac{\partial U}{\partial q_i}=-U.
\]
Using \(H=T-U\) gives
\[
  \boxed{\ddot I_{\mathrm{cm}}=4H+2U.}
\]
This identity is a diagnostic for the size of the configuration. For a bounded
noncollisional motion, the long-time average of \(\ddot I_{\mathrm{cm}}\)
vanishes, so time averaging gives the virial relation
\(2\langle T\rangle=\langle U\rangle\).
For escape or near-collision, \(I_{\mathrm{cm}}\) instead records the large-scale
expansion or collapse of the configuration.
\end{derivation}

\begin{remark}[Why the two-body reduction does not generalize]
For two bodies, the center-of-mass reduction leaves one relative vector in a
central potential. Angular momentum reduces the dynamics to a one-dimensional
radial problem, while the Laplace-Runge-Lenz vector gives the conic orbit
because the inverse-square potential has a hidden Kepler symmetry. Central
forces alone do not give that vector; generic central potentials still precess.
For three bodies there is no single relative radius and no central force after
reduction. Pairwise forces keep changing the effective center seen by each
body, so the Kepler-only constants are no longer constants of the full motion.
\end{remark}

\subsection{Central Configurations And Special Solutions}
\label{subsec:central-configurations}

Some exact motions survive this complexity. A configuration
\(a_1,a_2,a_3\), with center of mass at the origin, is a central configuration
if there is a scalar \(\lambda\) such that
\[
  \sum_{j\neq i}
  Gm_j\frac{a_j-a_i}{\norm{a_j-a_i}^3}
  =
  -\lambda a_i
  \qquad
  (i=1,2,3).
\]
Then the ansatz
\[
  q_i(t)=R(\omega t)a_i
\]
solves the equations in a plane provided \(\omega^2=\lambda\). The shape
rotates rigidly even though the force law is gravitational.

\begin{example}[Lagrange equilateral solution]
Place three bodies at the vertices of an equilateral triangle and choose the
center of mass as origin. The gravitational force on each body points toward
the center of mass, so the central-configuration equation holds. For three
equal masses \(m\) and side length \(s\), the force on one mass has magnitude
\[
  2\frac{Gm^2}{s^2}\cos\frac{\pi}{6}
  =
  \sqrt3\,\frac{Gm^2}{s^2}.
\]
The distance from a vertex to the centroid is \(s/\sqrt3\). Equating this to
the centripetal force \(m\omega^2s/\sqrt3\) gives
\[
  \omega^2=\frac{3Gm}{s^3}.
\]
This is a genuine three-body orbit with fixed shape.
\end{example}

Lagrange found the equilateral family in 1772; the collinear configurations
were Euler's, from 1767. Both are finite-mass ancestors of the Lagrange points
in the restricted problem.

\begin{example}[Euler collinear solutions]
There are also central configurations with the three bodies on one line. The
relative separations are fixed in a rotating frame, and the line rotates
uniformly. Unlike the equilateral case, the ratios of the two gaps depend on
the three masses and are determined by an algebraic equation. These collinear
solutions are the ancestors of \(L_1,L_2,L_3\) in the restricted problem.
For the ordering \(m_1,m_2,m_3\) on a line, set the distance \(m_2m_3\) equal
to \(x\) times the distance \(m_1m_2\). Euler's calculation gives the quintic
\[
\begin{aligned}
 &(m_1+m_2)x^5+(3m_1+2m_2)x^4+(3m_1+m_2)x^3\\
 &\qquad -(m_2+3m_3)x^2-(2m_2+3m_3)x-(m_2+m_3)=0.
\end{aligned}
\]
Its unique positive root gives the gap ratio for that ordering; the other two
collinear configurations come from the other orderings of the masses.

Here is the reduction behind the polynomial. Place the bodies at
\[
  q_1=0,\qquad q_2=d,\qquad q_3=d(1+x),
\]
and subtract the center-of-mass acceleration, since a central configuration is
allowed to translate uniformly before the rotating-frame description is chosen.
The central-configuration equation says that the gravitational acceleration of
each body differs from the common center-of-mass acceleration by
\(-\omega^2(q_i-q_{\mathrm{cm}})\). Taking the difference of the equations for
\((q_2,q_1)\) and for \((q_3,q_2)\) eliminates the center-of-mass acceleration
and leaves two expressions for the same \(\omega^2\). Equating them, clearing
the denominators \(d^2\), \(x^2d^2\), and \((1+x)^2d^2\), and collecting powers
of \(x\) gives the displayed quintic. Explicitly, with the common factor
\(G/d^2\) suppressed,
\[
\begin{aligned}
  a_2-a_1
  &=
  -(m_1+m_2)
  +m_3\left(\frac1{x^2}-\frac1{(1+x)^2}\right),\\
  a_3-a_2
  &=
  m_1\left(1-\frac1{(1+x)^2}\right)
  -\frac{m_2+m_3}{x^2}.
\end{aligned}
\]
The central-configuration condition requires \(a_3-a_2=x(a_2-a_1)\). Multiplying
this equation by \(x^2(1+x)^2\) produces the quintic above. Thus \(x\) is the
only shape variable, while \(d\) sets the overall scale.
\end{example}

Section~\ref{sec:lab-three-body} turns the equal-mass equilateral solution into
a numerical laboratory. There the exact rotating triangle is used as the
comparison orbit, and small perturbations of the initial velocities show how a
central configuration can organize nearby motion without making the full
three-body problem integrable.
Chenciner and Montgomery (2000) discovered the figure-eight periodic orbit:
three equal masses chase one another along a single planar curve, with zero
total angular momentum. It is not a central configuration, and like Lagrange's
and Euler's solutions it does not make the generic three-body problem
integrable. Special exact orbits can exist without integrability.

\section{Circular Restricted Three-Body Problem}
\label{sec:circular-restricted-three-body-problem}

The circular restricted three-body problem is the standard autonomous model
behind the Sun-Jupiter-asteroid story. Two massive primaries move on circular
Kepler orbits about their common center of mass, and a third body has zero
mass, so it feels the primaries but does not affect them. This is the setting
of Jacobi's integral (1836) and of Hill's lunar restricted problem (1878), and
the setting in which Birkhoff's periodic-orbit program made return maps and
periodic points operational. After choosing units
in which
\[
  G(m_1+m_2)=1,\qquad
  \text{primary separation}=1,\qquad
  \text{angular speed}=1,
\]
define the mass parameter
\[
  \mu_{\mathrm{CR3BP}}=\frac{m_2}{m_1+m_2},
  \qquad
  0<\mu_{\mathrm{CR3BP}}\leq \frac12.
\]
In the uniformly rotating barycentric frame the primaries are fixed at
\[
  P_1=(-\mu_{\mathrm{CR3BP}},0),
  \qquad
  P_2=(1-\mu_{\mathrm{CR3BP}},0).
\]
For compactness in this section write \(\mu=\mu_{\mathrm{CR3BP}}\). If the
test particle has rotating-frame position \(r=(x,y)\), set
\[
  r_1=\sqrt{(x+\mu)^2+y^2},
  \qquad
  r_2=\sqrt{(x-1+\mu)^2+y^2}.
\]
The effective potential is
\[
  \boxed{
  \Omega_{\mathrm{eff}}(x,y)
  =
  \frac12(x^2+y^2)
  +
  \frac{1-\mu}{r_1}
  +
  \frac{\mu}{r_2}.
  }
\]
The first term is the centrifugal potential; the last two are the gravitational
potentials of the fixed primaries.

\begin{derivation}[Rotating-frame equations]
Let \(\rho(t)=R(t)r(t)\) be the inertial position, where \(R(t)\) rotates with
angular velocity \(\hat z\). Differentiating twice gives
\[
  \ddot\rho
  =
  R(t)\left(
    \ddot r+2\hat z\times\dot r+\hat z\times(\hat z\times r)
  \right).
\]
In the plane, \(\hat z\times(\hat z\times r)=-r\). The inertial gravitational
acceleration, expressed in the rotating frame, is
\[
  - (1-\mu)\frac{r-P_1}{r_1^3}
  - \mu\frac{r-P_2}{r_2^3}
  =
  \nabla\left(\frac{1-\mu}{r_1}+\frac{\mu}{r_2}\right).
\]
Therefore
\[
  \ddot r+2\hat z\times\dot r-r
  =
  \nabla\left(\frac{1-\mu}{r_1}+\frac{\mu}{r_2}\right),
\]
equivalently
\[
  \boxed{
  \ddot r+2\hat z\times\dot r=\nabla\Omega_{\mathrm{eff}}(r).
  }
\]
In components,
\[
  \boxed{
  \ddot x-2\dot y=\Omega_x,
  \qquad
  \ddot y+2\dot x=\Omega_y.
  }
\]
The Coriolis force is velocity dependent and does no work in the rotating
frame, but the deeper reason a Jacobi integral exists is that the rotating-frame
Hamiltonian is autonomous. The conserved quantity is rotating-frame energy,
equivalently inertial energy minus \(\omega L_z\) in dimensional variables.
\end{derivation}

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1.25, >=Latex]
  \draw[->] (-2.25,0) -- (2.45,0) node[right] {\(x\)};
  \draw[->] (0,-1.72) -- (0,1.74) node[above] {\(y\)};
  \coordinate (pone) at (-0.55,0);
  \coordinate (ptwo) at (1.05,0);
  \coordinate (lone) at (0.62,0);
  \coordinate (ltwo) at (1.52,0);
  \coordinate (lthree) at (-1.62,0);
  \coordinate (lfour) at (0.25,1.386);
  \coordinate (lfive) at (0.25,-1.386);
  \fill[yellow!80!black] (pone) circle (3.5pt);
  \node[yellow!45!black, below=5pt] at (pone) {\(P_1\)};
  \fill[orange!80!black] (ptwo) circle (2.6pt);
  \node[orange!80!black, below=5pt] at (ptwo) {\(P_2\)};
  \foreach \pt/\lab/\anchor in {lone/L_1/above left,ltwo/L_2/above right,lthree/L_3/above}
  {
    \draw[red!75!black, thick] (\pt) ++(0,-0.08) -- ++(0,0.16);
    \node[red!75!black, \anchor=3pt] at (\pt) {\(\lab\)};
  }
  \foreach \pt/\lab/\anchor in {lfour/L_4/above right,lfive/L_5/below right}
  {
    \draw[red!75!black, thick] (\pt) ++(-0.08,-0.08) -- ++(0.16,0.16);
    \draw[red!75!black, thick] (\pt) ++(-0.08,0.08) -- ++(0.16,-0.16);
    \node[red!75!black, \anchor=3pt] at (\pt) {\(\lab\)};
  }
  \draw[blue!60!black, ->] (1.20,0.45) arc (20:95:1.05);
  \node[blue!60!black, fill=white, inner sep=1pt] at (1.70,1.18)
    {rotating frame};
\end{tikzpicture}
\caption{Circular restricted three-body geometry in rotating barycentric
coordinates. The primaries are fixed, while the test particle moves under
gravity, centrifugal force, and Coriolis deflection.}
\label{fig:cr3bp-rotating-frame}
\end{figure}

\begin{derivation}[Jacobi integral]
Dot the rotating-frame equation with \(\dot r\):
\[
  \dot r\cdot\ddot r
  +
  2\dot r\cdot(\hat z\times\dot r)
  =
  \dot r\cdot\nabla\Omega_{\mathrm{eff}}.
\]
The Coriolis term vanishes because \(\dot r\cdot(\hat z\times\dot r)=0\). Hence
\[
  \frac{\dd}{\dd t}\left(\frac12\norm{\dot r}^2\right)
  =
  \frac{\dd}{\dd t}\Omega_{\mathrm{eff}}(r(t)).
\]
Therefore
\[
  \frac12\norm{\dot r}^2-\Omega_{\mathrm{eff}}=\text{constant}.
\]
It is customary to multiply by \(-2\) and define the Jacobi constant
\[
  \boxed{
  C_J=2\Omega_{\mathrm{eff}}(x,y)-(\dot x^2+\dot y^2).
  }
\]
This is the energy integral of the autonomous rotating-frame problem.
Equivalently, before nondimensionalizing the angular speed to one, the
conserved rotating-frame Hamiltonian is the inertial energy shifted by
\(-\omega L_z\); the Jacobi constant is the customary sign-reversed version of
that autonomous Hamiltonian.
\end{derivation}

\begin{derivation}[Canonical Hamiltonian in the rotating frame]
The rotating-frame equations also come from a genuine autonomous Hamiltonian,
but the canonical momenta are not the velocity components.  The Lagrangian per
unit test-particle mass is
\[
  L_{\mathrm{rot}}
  =
  \frac12(\dot x^2+\dot y^2)
  +x\dot y-y\dot x
  +\Omega_{\mathrm{eff}}(x,y).
\]
The linear velocity term is the Coriolis term in variational form.  Hence
\[
  p_x=\frac{\partial L_{\mathrm{rot}}}{\partial\dot x}=\dot x-y,
  \qquad
  p_y=\frac{\partial L_{\mathrm{rot}}}{\partial\dot y}=\dot y+x.
\]
Solving for velocity gives \(\dot x=p_x+y\) and \(\dot y=p_y-x\).  The
Legendre transform gives
\[
\begin{aligned}
  H_{\mathrm{rot}}
  &=
  p_x\dot x+p_y\dot y-L_{\mathrm{rot}}\\
  &=
  \frac12(p_x^2+p_y^2)+yp_x-xp_y
  -\frac{1-\mu}{r_1}-\frac{\mu}{r_2}.
\end{aligned}
\]
Equivalently,
\[
  H_{\mathrm{rot}}
  =
  \frac12\big((p_x+y)^2+(p_y-x)^2\big)
  -\Omega_{\mathrm{eff}}(x,y).
\]
Along solutions
\[
  C_J=-2H_{\mathrm{rot}}.
\]
This equality is often the safest way to translate between the geometric
zero-velocity picture and the canonical phase-space picture: velocities are
natural for drawing allowed regions, while \((x,y,p_x,p_y)\) are natural for
Hamiltonian perturbation theory and Poincare maps.
\end{derivation}

The Jacobi constant gives an immediate geometric constraint. Since the squared
speed is nonnegative,
\[
  \dot x^2+\dot y^2
  =
  2\Omega_{\mathrm{eff}}(x,y)-C_J
  \geq0.
\]
Thus the allowed region in the rotating plane is
\[
  2\Omega_{\mathrm{eff}}(x,y)\geq C_J,
\]
and curves satisfying
\[
  2\Omega_{\mathrm{eff}}(x,y)=C_J
\]
are zero-velocity curves. They are not physical walls; they are projections of
the energy surface to configuration space. A test particle cannot cross a
forbidden region because that would require imaginary speed. When the
zero-velocity neck near \(L_1\) or \(L_2\) opens, transport between regions
becomes possible. This is the rotating-frame picture behind many capture,
escape, and ejection mechanisms.
The allowed components are often called Hill regions. Near a secondary primary,
the small allowed lobe gives the Hill-sphere approximation used in celestial
mechanics: it is the configuration-space shadow of the same Jacobi-energy
constraint, not a literal material boundary.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1.0, >=Latex]
  \begin{scope}
    \fill[gray!22] (-2.0,-1.15) rectangle (2.0,1.15);
    \draw[->] (-2.0,0) -- (2.25,0) node[right] {\(x\)};
    \draw[->] (0,-1.25) -- (0,1.25) node[above] {\(y\)};
    \draw[blue!70!black, thick] (-0.75,0) ellipse (0.62 and 0.45);
    \draw[blue!70!black, thick] (0.88,0) ellipse (0.34 and 0.27);
    \fill[white] (-0.75,0) ellipse (0.60 and 0.43);
    \fill[white] (0.88,0) ellipse (0.32 and 0.25);
    \fill[yellow!80!black] (-0.75,0) circle (2.8pt);
    \node[yellow!45!black, below=4pt, fill=white, inner sep=1pt,
      font=\small] at (-0.75,0) {\(P_1\)};
    \fill[orange!80!black] (0.88,0) circle (2.2pt);
    \node[orange!80!black, below=4pt, fill=white, inner sep=1pt,
      font=\small] at (0.88,0) {\(P_2\)};
    \draw[red!75!black, thick] (0.2,-0.09) -- (0.2,0.09);
    \node[red!75!black, above=3pt, fill=white, inner sep=1.5pt,
      font=\small] at (0.2,0.09) {\(L_1\)};
    \node[align=center] at (0,-1.65) {neck closed\\separate allowed lobes};
  \end{scope}
  \begin{scope}[xshift=5.6cm]
    \fill[gray!22] (-2.0,-1.15) rectangle (2.0,1.15);
    \draw[->] (-2.0,0) -- (2.25,0) node[right] {\(x\)};
    \draw[->] (0,-1.25) -- (0,1.25) node[above] {\(y\)};
    \draw[blue!70!black, thick]
      (-1.45,0) .. controls (-1.25,0.78) and (-0.36,0.82) .. (0.28,0.28)
      .. controls (0.88,0.66) and (1.38,0.52) .. (1.55,0)
      .. controls (1.38,-0.52) and (0.88,-0.66) .. (0.28,-0.28)
      .. controls (-0.36,-0.82) and (-1.25,-0.78) .. (-1.45,0);
    \fill[white]
      (-1.28,0) .. controls (-1.12,0.62) and (-0.34,0.66) .. (0.28,0.17)
      .. controls (0.78,0.50) and (1.18,0.40) .. (1.34,0)
      .. controls (1.18,-0.40) and (0.78,-0.50) .. (0.28,-0.17)
      .. controls (-0.34,-0.66) and (-1.12,-0.62) .. (-1.28,0);
    \fill[yellow!80!black] (-0.75,0) circle (2.8pt);
    \node[yellow!45!black, below=4pt, fill=white, inner sep=1pt,
      font=\small] at (-0.75,0) {\(P_1\)};
    \fill[orange!80!black] (0.88,0) circle (2.2pt);
    \node[orange!80!black, below=4pt, fill=white, inner sep=1pt,
      font=\small] at (0.88,0) {\(P_2\)};
    \draw[red!75!black, thick] (0.2,-0.09) -- (0.2,0.09);
    \node[red!75!black, above left=4pt, fill=white, inner sep=1.5pt,
      font=\small] at (0.2,0.09) {\(L_1\)};
    \draw[red!75!black, thick] (1.18,-0.09) -- (1.18,0.09);
    \node[red!75!black, above right=4pt, fill=white, inner sep=1.5pt,
      font=\small] at (1.18,0.09) {\(L_2\)};
    \draw[red!75!black, ->] (0.2,-0.52) -- (0.2,-0.14);
    \node[align=center] at (0,-1.65) {neck open\\transport channel};
  \end{scope}
\end{tikzpicture}
\caption{Schematic zero-velocity curves in the rotating CR3BP plane. Gray
regions are forbidden by \(2\Omega_{\mathrm{eff}}<C_J\); white regions are
allowed projections of the energy surface. Changing \(C_J\) can open a neck
near \(L_1\), creating a channel between previously separated regions.}
\label{fig:cr3bp-zero-velocity-neck}
\end{figure}

\subsection{Lagrange Points}
\label{subsec:lagrange-points}

Equilibria in the rotating frame satisfy
\[
  \dot x=\dot y=\ddot x=\ddot y=0,
  \qquad
  \nabla\Omega_{\mathrm{eff}}(x,y)=0.
\]
There are five such points. The collinear points \(L_1,L_2,L_3\) lie on the
\(x\)-axis and solve the scalar equation
\[
  x
  -
  (1-\mu)\frac{x+\mu}{|x+\mu|^3}
  -
  \mu\frac{x-1+\mu}{|x-1+\mu|^3}
  =
  0.
\]
The absolute values encode which side of each primary the point lies on. There
is one solution between the primaries, one to the right of \(P_2\), and one to
the left of \(P_1\).

The triangular points are explicit:
\[
  \boxed{
  L_4=\left(\frac12-\mu,\frac{\sqrt3}{2}\right),
  \qquad
  L_5=\left(\frac12-\mu,-\frac{\sqrt3}{2}\right).
  }
\]
At either point \(r_1=r_2=1\). Substituting \(L_4\), for example, gives
\[
\begin{aligned}
  \Omega_x
  &=
  x-(1-\mu)(x+\mu)-\mu(x-1+\mu)\\
  &=
  \left(\frac12-\mu\right)
  -(1-\mu)\frac12
  -\mu\left(-\frac12\right)
  =
  0,
\end{aligned}
\]
and similarly \(\Omega_y=0\). The gravitational and centrifugal forces balance
in the rotating frame.

\begin{remark}[Linear stability of \(L_4,L_5\)]
Linearizing about \(L_4\) or \(L_5\) gives stable oscillations when
\[
  \mu(1-\mu)<\frac{1}{27}.
\]
Here is the calculation. Write a perturbation as
\((\xi,\eta)\) around \(L_4\). At \(L_4\),
differentiate
\[
  \Omega_{\mathrm{eff}}
  =
  \frac12(x^2+y^2)
  +
  \frac{1-\mu}{r_1}
  +
  \frac{\mu}{r_2}
\]
twice. If \(X_1=x+\mu\), \(X_2=x-1+\mu\), and
\(m_1=1-\mu\), \(m_2=\mu\), then
\[
\begin{aligned}
  \Omega_{xx}
  &=
  1-\sum_{a=1}^2\frac{m_a}{r_a^3}
  +3\sum_{a=1}^2\frac{m_aX_a^2}{r_a^5},\\
  \Omega_{yy}
  &=
  1-\sum_{a=1}^2\frac{m_a}{r_a^3}
  +3\sum_{a=1}^2\frac{m_ay^2}{r_a^5},\\
  \Omega_{xy}
  &=
  3\sum_{a=1}^2\frac{m_aX_ay}{r_a^5}.
\end{aligned}
\]
At \(L_4\), \(r_1=r_2=1\), \(X_1=1/2\),
\(X_2=-1/2\), and \(y=\sqrt3/2\), so
\[
  \Omega_{xx}=\frac34,\qquad
  \Omega_{yy}=\frac94,\qquad
  \Omega_{xy}=\frac{3\sqrt3}{4}(1-2\mu).
\]
The linearized equations
\[
  \ddot\xi-2\dot\eta=\Omega_{xx}\xi+\Omega_{xy}\eta,\qquad
  \ddot\eta+2\dot\xi=\Omega_{xy}\xi+\Omega_{yy}\eta
\]
have characteristic polynomial
\[
  \lambda^4+\lambda^2+\frac{27}{4}\mu(1-\mu)=0.
\]
Put \(s=\lambda^2\). Then
\[
  s^2+s+\frac{27}{4}\mu(1-\mu)=0,
  \qquad
  s_\pm=
  \frac{-1\pm\sqrt{1-27\mu(1-\mu)}}{2}.
\]
All four eigenvalues are purely imaginary exactly when the two roots
\(s_\pm\) are real and negative. Since their sum is \(-1\) and their product
is positive for \(0<\mu<1\), this is equivalent to the discriminant condition
\[
  1-27\mu(1-\mu)\ge0.
\]
Equivalently, for \(\mu\leq 1/2\),
\[
  \mu<\mu_R
  =
  \frac12\left(1-\sqrt{\frac{23}{27}}\right)
  \approx 0.03852.
\]
The Sun-Jupiter value is about \(9.53\times10^{-4}\), far below this threshold.
Trojan motion near the triangular points is therefore stable in this
linearized nontrivial three-body setting.
The stability criterion was derived by Gascheau (1843) and Routh (1875); its
Solar System relevance was confirmed in 1906, when Max Wolf discovered the
first Trojan asteroid, 588 Achilles, near Jupiter's \(L_4\).
\end{remark}

\subsection{Connection To The Asteroid-Belt Model}
\label{subsec:cr3bp-to-asteroid-belt}

The CR3BP is autonomous in the rotating barycentric frame and has the Jacobi
constant. The ejection-probability demo uses a closely related heliocentric
form in which Jupiter is prescribed on a circular orbit and the asteroid is
integrated in Sun-centered coordinates. The two descriptions are related by a
time-dependent rotation and translation, but they emphasize different things:
\begin{itemize}
\item rotating barycentric CR3BP: Jacobi constant, zero-velocity curves, and
Lagrange points;
\item heliocentric asteroid ensemble: Kepler elements, resonance bins, and
finite-time ejection statistics.
\end{itemize}
Both are useful. The first gives the clean phase-space geometry of escape
channels. The second makes direct contact with semimajor axes, mean-motion
resonances, and the asteroid-belt simulations.

The CR3BP companion script computes \(L_1,\ldots,L_5\), plots representative
zero-velocity curves, and prints the drift of the Jacobi constant along the
numerical trajectory. Its time stepper is classical fourth-order Runge-Kutta:
this is transparent for a first classroom calculation, but it is not
symplectic and it does not preserve the Jacobi integral exactly. Long-time
transport claims should therefore be checked against the printed Jacobi drift
and, in research calculations, against a structure-aware integrator.

\section{Heliocentric Sun-Jupiter-Asteroid Equations}
\label{sec:restricted-sun-jupiter-asteroid}

Let \(r\in\R^2\) be the asteroid position relative to the Sun. Jupiter is
prescribed by
\[
  r_J(t)=a_J(\cos n_Jt,\sin n_Jt),
  \qquad
  n_J^2=\frac{G(M_\odot+M_J)}{a_J^3}.
\]
In a heliocentric non-inertial frame, the asteroid acceleration is
\[
  \boxed{
  \ddot r
  =
  -\frac{GM_\odot r}{\norm r^3}
  -
  GM_J
  \left(
    \frac{r-r_J}{\norm{r-r_J}^3}
    +
    \frac{r_J}{\norm{r_J}^3}
  \right)
  }.
\]
The first term is solar gravity. The first part of the Jupiter term is direct
gravitational attraction to Jupiter. The final term is the indirect term caused
by the acceleration of the heliocentric frame.

Equivalently, the asteroid obeys the explicitly time-dependent Hamiltonian
\[
  H_{\mathrm{hel}}(r,p,t)
  =
  \frac12\norm p^2
  -
  \frac{GM_\odot}{\norm r}
  -
  \frac{GM_J}{\norm{r-r_J(t)}}
  +
  GM_J\frac{r\cdot r_J(t)}{\norm{r_J(t)}^3}.
\]
The last term is the potential whose gradient gives the indirect acceleration.
Because \(r_J(t)\) is periodic, this is a periodically forced one-particle
Hamiltonian. It can be made autonomous by adjoining the forcing phase
\(\phi=n_Jt\) and its conjugate action \(I_\phi\). With
\[
  r_J(\phi)=a_J(\cos\phi,\sin\phi),
\]
the extended Hamiltonian is
\[
  K(r,p,\phi,I_\phi)
  =
  H_{\mathrm{hel}}(r,p,\phi)+n_JI_\phi,
\]
so Hamilton's equation gives \(\dot\phi=\partial K/\partial I_\phi=n_J\). The
resulting autonomous system has more than one degree of freedom and can exhibit
resonance and chaotic transport.

\begin{derivation}[Indirect term]
Let \(x\) be the asteroid position in an inertial frame and \(x_\odot\) the Sun
position. The heliocentric vector is \(r=x-x_\odot\). Then
\[
  \ddot r=\ddot x-\ddot x_\odot.
\]
The direct Jupiter acceleration of the asteroid is
\[
  -GM_J\frac{x-x_J}{\norm{x-x_J}^3}.
\]
The acceleration of the Sun due to Jupiter is
\[
  \ddot x_\odot=
  GM_J\frac{x_J-x_\odot}{\norm{x_J-x_\odot}^3}
  =
  GM_J\frac{r_J}{\norm{r_J}^3}.
\]
Subtracting \(\ddot x_\odot\) produces the indirect term
\[
  -GM_J\frac{r_J}{\norm{r_J}^3}.
\]
\end{derivation}

\begin{remark}[Why the indirect term matters]
Omitting the indirect term corresponds to holding the Sun fixed while Jupiter
pulls on the asteroid. That is sometimes useful as a simplified classroom
model, but it is not the heliocentric form of the restricted three-body
problem. Including the indirect term ensures that the coordinate origin
accelerates exactly as the Sun does in the prescribed Sun-Jupiter motion.
\end{remark}

\section{Numerical Ejection Probability}
\label{sec:numerical-ejection-probability}

The highlighted simulation samples \(N\) test particles with initial semimajor
axes \(a_0\in[a_{\min},a_{\max}]\), small eccentricities, random orbital phases,
and random periapse directions. Let \(T\) be the integration horizon.

\begin{definition}[Finite-time ejection probability]
For a specified model, sampling distribution, time step, and ejection criterion,
the finite-time ejection probability estimate is
\[
  \widehat P_{\mathrm{ej}}(T)
  =
  \frac{1}{N}
  \sum_{i=1}^{N}\mathbf 1\{\text{particle \(i\) is ejected before time \(T\)}\}.
\]
\end{definition}

\begin{remark}[Ejection criterion in the demo]
The Python demo counts a particle as ejected if it reaches a large heliocentric
radius or if its heliocentric two-body energy is positive after it has moved
outside Jupiter's orbital scale. In the default configuration this means
\[
  r>20\,\mathrm{AU}
  \quad\text{or}\quad
  E_{\odot}>0\ \text{and}\ r>a_J,
\]
with \(a_J=5.2044\,\mathrm{AU}\). The same run separately records close
approaches to Jupiter, using the default threshold \(0.03\,\mathrm{AU}\), and
collisions with the inner solar cutoff, using the default threshold
\(0.02\,\mathrm{AU}\).
\end{remark}

\(\widehat P_{\mathrm{ej}}(T)\) is not a property of the three-body problem in
isolation; it is a property of an experiment. Five inputs determine it:
\[
  \text{model}+\text{initial ensemble}+\text{time horizon}
  +\text{numerical method}+\text{loss criterion}.
\]
The same physical question with different choices yields different numerical
values.

The demo also labels each semimajor-axis bin by its nearest low-order interior
mean-motion resonance with Jupiter. The nominal resonance locations are
\[
  a_{p:q}=a_J\left(\frac{q}{p}\right)^{2/3},
  \qquad
  (p:q)\in\{3:1,5:2,7:3,2:1\}.
\]
These labels are a diagnostic overlay. A semimajor-axis bin is wider and cruder
than a resonance island, and the actual resonance width depends on eccentricity
and perturbation strength. The overlay asks whether large loss fractions cluster
near the same places where perturbation theory predicts slow resonant angles.

For a bin containing \(n_b\) sampled particles and \(e_b\) ejections, the
reported fraction is
\[
  \widehat p_b=\frac{e_b}{n_b}.
\]
If the particles in that bin are treated as independent Bernoulli trials, a
rough sampling uncertainty is
\[
  \operatorname{SE}(\widehat p_b)
  \approx
  \sqrt{\frac{\widehat p_b(1-\widehat p_b)}{n_b}}.
\]
This is only a Monte Carlo error bar for the chosen numerical experiment. It
does not measure modeling error, timestep error, or the long-time instability
of the idealized Hamiltonian system.

\section{Integrator Used In The Demo}
\label{sec:asteroid-integrator}

The demo uses a kick-drift-kick style update. If \(a(r,t)\) is the acceleration,
then
\[
\begin{aligned}
  v_{n+1/2} &= v_n+\frac{\Delta t}{2}a(r_n,t_n),\\
  r_{n+1} &= r_n+\Delta t\,v_{n+1/2},\\
  v_{n+1} &= v_{n+1/2}+\frac{\Delta t}{2}a(r_{n+1},t_{n+1}).
\end{aligned}
\]
As written in physical time with prescribed Jupiter forcing, this update is not
the symplectic map of the autonomous extended Hamiltonian \(K\). It is still
pedagogically preferable to an opaque black-box integrator because every force
evaluation is visible.

\section{What To Take Forward}
\label{sec:three-body-take-forward}

\begin{itemize}
\item The full three-body problem is Newtonian and Hamiltonian, but its
translation and rotation symmetries do not reduce it to a one-degree-of-freedom
radial equation.  The failure is structural, not a lack of clever coordinates:
Bruns proved in 1887 that there are no additional integrals algebraic in the
coordinates, momenta, and time, apart from the classical ones; Painleve extended
the algebraic scope to velocity variables. Poincare's 1890 theorem addressed a
different class, ruling out generic additional analytic integrals in the
perturbative setting. The obstruction is the same small-divisor mechanism seen
in Chapter~\ref{chap:perturbation-chaos}: a formal integral must solve
homological equations whose resonant denominators cannot be made to vanish
generically. These theorems are not identical statements with different names:
Bruns and Painleve limit algebraic first integrals of the full problem, while
Poincare explains why analytic integrability fails under generic perturbation
of an integrable model.
\item Jacobi coordinates remove center-of-mass motion and diagonalize the mass
metric.  They reveal two coupled relative vectors, which is the simplest
coordinate-level reason the two-body Kepler solution does not generalize.
\item Central configurations are exact special motions, not generic solutions.
They are valuable because they organize nearby dynamics and survive as
Lagrange-point structures in the restricted problem.
\item The circular restricted three-body problem supplies an autonomous
rotating-frame model with Lagrange points, zero-velocity curves, a canonical
Hamiltonian, and the Jacobi constant. These structures are the standard
geometric language for capture and escape channels.
\item The heliocentric Sun-Jupiter-asteroid model is a periodically forced
Kepler problem.  Its indirect term is part of the coordinate transformation,
not an optional correction.
\item Numerical ejection probabilities are finite-time, model-dependent
observables. They become meaningful only when the ensemble, integration time,
timestep, invariants, and loss criterion are stated.
\end{itemize}

\section{Chapter-End Laboratories And Exercises}
\label{sec:three-body-chapter-end-labs}

This laboratory is the finite-mass counterpart of the restricted
Sun-Jupiter-asteroid simulations: derive Newton's equations, choose controlled
initial data, and monitor invariants before interpreting instability.

\subsection{Equal-Mass Three-Body Laboratory}
\label{sec:lab-three-body}

For an equal-mass three-body laboratory, take three masses in the plane with
pair potential
\[
  V=-k\left(\frac1{r_{12}}+\frac1{r_{23}}+\frac1{r_{13}}\right),
\]
where
\[
  r_{ij}=\norm{r_i-r_j}.
\]
The Lagrangian is
\[
  L=\sum_{i=1}^3\frac12m\norm{\dot r_i}^2
  +
  k\left(\frac1{r_{12}}+\frac1{r_{23}}+\frac1{r_{13}}\right).
\]
The equations are
\[
  m\ddot r_i
  =
  -k\sum_{j\neq i}\frac{r_i-r_j}{\norm{r_i-r_j}^3}.
\]
The symmetric initial positions are
\[
  r_1(0)=(1,0),
  \qquad
  r_2(0)=\left(-\frac12,\frac{\sqrt3}{2}\right),
  \qquad
  r_3(0)=\left(-\frac12,-\frac{\sqrt3}{2}\right).
\]

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1.15, >=Latex, font=\small]
  \draw[->] (-2.0,0) -- (2.2,0) node[right] {\(x\)};
  \draw[->] (0,-1.7) -- (0,1.9) node[above] {\(y\)};
  \coordinate (r1) at (1.25,0);
  \coordinate (r2) at (-0.625,1.083);
  \coordinate (r3) at (-0.625,-1.083);
  \draw[gray!70, thick] (r1) -- (r2) -- (r3) -- cycle;
  \fill[blue!70!black] (r1) circle (2pt)
    node[above, fill=white, inner sep=1pt] {\(r_1\)};
  \fill[blue!70!black] (r2) circle (2pt)
    node[above left, fill=white, inner sep=1pt] {\(r_2\)};
  \fill[blue!70!black] (r3) circle (2pt)
    node[below left, fill=white, inner sep=1pt] {\(r_3\)};
  \draw[->, red!75!black, thick] (r1) -- ++(0,0.65);
  \draw[->, red!75!black, thick] (r2) -- ++(-0.56,-0.32);
  \draw[->, red!75!black, thick] (r3) -- ++(0.56,-0.32);
  \draw[dashed] (0,0) circle (1.25);
  \node[red!75!black, align=center, fill=white, inner sep=1.5pt] at (1.72,1.36)
    {tangent velocities\\for rotating triangle};
  \node[align=center, fill=white, inner sep=1.5pt] at (0,-1.95)
    {the perturbation lab changes the initial velocities\\and monitors
     energy, angular momentum, and separation};
\end{tikzpicture}
\caption{Equal-mass three-body laboratory. The unperturbed comparison solution
starts from an equilateral triangle with tangential velocities; perturbing this
initial data is a controlled way to watch the loss of symmetric motion.}
\label{fig:lab-three-body-initial-data}
\end{figure}
\FloatBarrier

For the rotating equilateral solution, velocities are tangent to the
circumcircle. A representative perturbed initial condition uses a speed
parameter \(v=0.8\) and perturbation \(\epsilon=0.05\), with initial velocities
such as
\[
  \dot x_1(0)=\epsilon,\qquad
  \dot x_2(0)=-\frac{\sqrt3}{2}v-\epsilon,\qquad
  \dot x_3(0)=\frac{\sqrt3}{2}v.
\]
The workflow matters more than the particular number \(0.8\): derive Newton's
equations from a Lagrangian, integrate the exact pairwise forces, monitor
energy and angular momentum, and compare unperturbed symmetric motion with
perturbed motion. This is the finite-mass cousin of the Sun-Jupiter-asteroid
ejection simulation.

\subsection{Questions For The Reader}
\label{sec:three-body-questions}

\begin{enumerate}
\item Starting from the inertial three-body Lagrangian, derive the
center-of-mass equation and explain why internal gravitational forces cancel in
pairs.
\item Verify the Jacobi-coordinate kinetic energy formula by substituting the
inverse transformation for \(q_1,q_2,q_3\). Which mass factors would be wrong
if the center of mass of bodies \(1\) and \(2\) were not used in \(\rho_2\)?
\item Use the homogeneity of \(U\) to rederive the Lagrange-Jacobi identity.
What does it imply for a bounded noncollisional orbit?
\item For the Sun-Jupiter value of \(\mu_{\mathrm{CR3BP}}\), where are
\(L_1,L_2,L_3\) relative to Jupiter?
\item For a chosen Jacobi constant, which zero-velocity necks are open?
\item Show explicitly that \(C_J=-2H_{\mathrm{rot}}\) after converting
canonical momenta to rotating-frame velocities.
\item How does \(\widehat P_{\mathrm{ej}}(T)\) change with \(T\)?
\item Which semimajor-axis bins have the largest ejection fractions?
\item What changes when \(M_J\) is scaled upward?
\item Which losses are close encounters, and which are clean ejections?
\item How sensitive are the results to \(\Delta t\)?
\item Are apparent resonance structures stable under a different random seed?
\end{enumerate}

\section{Associated Demos}
\label{sec:three-body-demos}

\begin{itemize}
\item \path{demos/python/circular_restricted_three_body.py}
\item \path{demos/python/asteroid_ejection_probability.py}
\item \path{demos/mathematica/CircularRestrictedThreeBody.wl}
\item \path{demos/mathematica/AsteroidEjectionProbability.wl}
\end{itemize}
Use the CR3BP demo first: it checks the Jacobi constant and draws the
zero-velocity geometry. Use the ejection-probability demo second: it turns the
same dynamical ideas into an ensemble statistic, so the meaningful outputs are
not only trajectories but also loss criteria, timestep sensitivity, resonance
bins, and finite-time uncertainty.
