\chapter{Elastic Bodies}
\label{chap:elastic-body}

\section{Elastic Solids As Material Continua}
\label{sec:elastic-material-continua}

Pull a rubber band and let go. It snaps back. Push a finger through honey and
the honey flows around the finger. The rubber band stores energy in shear; the
honey responds through a rate-dependent stress. That distinction separates an
elastic solid from a simple fluid.

Elasticity is the continuum setting closest to the finite-dimensional mechanics
developed earlier because every material point keeps a persistent label. This
lets the configuration-space logic of particles and rigid bodies pass into field
theory with minimal translation. The basic unknown is the material map
\[
  \varphi:B\times I\to\R^3,
  \qquad
  x=\varphi(X,t),
\]
where \(X\in B\) is the label of a material point and \(x\) is its current
spatial position. The central questions are local and material: how the
deformation gradient measures stretch, which strain fields can come from a
single-valued displacement, how stored energy produces stress, and how
boundary-value problems reduce the field equations to measurable stiffness
laws. Chapter~\ref{chap:deforming-body} keeps the material body but separates
overall rigid placement from internal shape. Chapter~\ref{chap:fluid} changes
the constitutive question: material labels can still be used in derivations, but
the state of a simple fluid is organized most effectively by Eulerian fields and
rate-dependent stress.

\paragraph{Roadmap.}
The chapter has one main narrative.  Deformation gradients measure local
stretching, compatibility distinguishes true displacement fields from
incompatible strain data, and a hyperelastic action gives elastodynamic
equations and stress.  The finite-strain theory is kept first because it makes
objectivity and stress measures unambiguous; the linear theory is then obtained
as a controlled small-displacement limit.

The guiding distinction is between intrinsic deformation and frame description.
Quantities such as \(C=F^TF\) and the Green strain do not change under a rigid
rotation of the observer. Stress measures such as \(P\) and \(\sigma\) live in
different configurations and are related by the deformation gradient.  Each
stress or strain tensor is introduced together with the space on which it acts
and the boundary term that defines its mechanical meaning.

Appendix~\ref{app:continuum-caveats} returns to these choices in a more
technical register, including weak forms, boundary terms, incompressibility as
a constraint, and the functional-analytic caveats behind existence and
regularity statements.

\begin{assumption}[Regular deformation map]
Unless stated otherwise, \(\varphi(\cdot,t)\) is a smooth orientation-preserving
embedding of the reference body into space:
\[
  J(X,t)=\det F(X,t)>0.
\]
The inequality \(J>0\) rules out local interpenetration and orientation
reversal. Global injectivity is stronger than \(J>0\). We assume it when contact
or self-intersection matters; otherwise the local stress derivations use only
nonsingularity of \(F\).
\end{assumption}

\begin{assumption}[Reference versus spatial densities]
The reference density \(\rho_0(X)\) is mass per reference volume. The spatial
density \(\rho(x,t)\) is mass per current volume. Conservation of mass for a
material volume gives
\[
  \rho(\varphi(X,t),t)J(X,t)=\rho_0(X).
\]
Thus \(J\) converts volume elements and \(\rho_0\) converts kinetic energy,
momentum balance, and stress work back to the fixed material domain \(B\).
\end{assumption}

\section{Deformation Gradient And Strain}
\label{sec:deformation-gradient-strain}

Elasticity begins by comparing nearby material points before and after
deformation. Local strain is measured by the derivative of the displacement
map: two deformations that differ by a rigid translation have the same local
strain. This is why the deformation gradient \(F\), and the metric-like tensor
\(C=F^TF\), are introduced before stresses and constitutive laws.

The deformation gradient is the local linearization of the deformation map:
\[
  F^i{}_A=\frac{\partial \varphi^i}{\partial X^A}.
\]
If \(\delta X\) is a small material vector, then its spatial image is
\[
  \delta x^i=F^i{}_A\delta X^A.
\]
The right Cauchy-Green tensor measures squared lengths after deformation:
\[
  |\delta x|^2
  =
  C_{AB}\delta X^A\delta X^B,
  \qquad
  C_{AB}=F^i{}_AF^i{}_B.
\]
The Green strain is
\[
  E_{AB}=\frac12(C_{AB}-\delta_{AB}).
\]
The names are part of the logic: Cauchy's finite-deformation tensors and
Green's strain organize deformation by comparing metrics, not by subtracting
points. Truesdell and Noll later made the objectivity requirements systematic
in modern continuum mechanics.
It vanishes for every rigid motion
\[
  \varphi(X)=a+QX,\qquad Q\in\SO(3),
\]
because \(F=Q\) and \(Q^TQ=I\). This invariance is why \(E\), not \(F-I\), is a
geometric strain measure for finite deformations. It is not the only such
measure. The Almansi strain, Biot strain, Hencky logarithmic strain, and the
Seth-Hill family are also objective finite-strain measures; they agree to first
order in small strain and differ in how they organize finite stretch.

For small displacements, the finite strain reduces to the symmetric gradient
of the displacement. To see this, write
\[
  \varphi(X,t)=X+\xi(X,t),
\]
one has
\[
  F^i{}_A=\delta^i_A+\partial_A\xi^i.
\]
Keeping only first-order displacement gradients,
\[
  E_{AB}
  =
  \frac12(\partial_A\xi_B+\partial_B\xi_A)
  +
  O(|\nabla\xi|^2).
\]
The infinitesimal strain is therefore
\[
  \boxed{
  e_{AB}=\frac12(\partial_A\xi_B+\partial_B\xi_A).
  }
\]

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1.0, >=Latex, font=\small]
  \begin{scope}
    \draw[->] (-0.2,0) -- (3.6,0) node[right] {\(X^1\)};
    \draw[->] (0,-0.2) -- (0,2.7) node[above] {\(X^2\)};
    \draw[fill=blue!7, thick]
      (0.55,0.45) .. controls (1.0,0.2) and (2.3,0.35) .. (2.85,0.75)
      .. controls (3.15,1.35) and (2.45,2.0) .. (1.65,2.15)
      .. controls (0.75,2.0) and (0.35,1.25) .. (0.55,0.45);
    \coordinate (X) at (1.4,1.15);
    \draw[black, fill=black] (X) circle (1.6pt) node[below left] {\(X\)};
    \draw[red!75!black, thick, ->] (X) -- ++(0.9,0.38);
    \node[red!75!black, fill=white, inner sep=1pt] at (2.12,1.74) {\(\delta X\)};
    \draw[red!75!black, ->] (1.96,1.60) -- (1.74,1.30);
    \node at (1.75,-0.55) {reference body \(B\)};
  \end{scope}
  \draw[thick, ->] (3.55,1.35) -- node[above] {\(\varphi(X,t)\)} (5.45,1.35);
  \begin{scope}[xshift=5.7cm]
    \draw[->] (-0.2,0) -- (3.7,0) node[right] {\(x^1\)};
    \draw[->] (0,-0.2) -- (0,2.7) node[above] {\(x^2\)};
    \draw[fill=green!7, thick]
      (0.55,0.55) .. controls (1.25,0.1) and (2.85,0.45) .. (3.05,1.0)
      .. controls (3.25,1.7) and (2.35,2.45) .. (1.25,2.18)
      .. controls (0.45,1.95) and (0.25,1.05) .. (0.55,0.55);
    \coordinate (x) at (1.55,1.15);
    \draw[black, fill=black] (x) circle (1.6pt);
    \node[fill=white, inner sep=1pt] at (2.75,0.42) {\(x=\varphi(X)\)};
    \draw[black!55, ->] (2.42,0.55) -- (1.67,1.05);
    \draw[red!75!black, thick, ->] (x) -- ++(1.05,0.62);
    \node[red!75!black, fill=white, inner sep=1pt] at (2.74,1.98)
      {\(\delta x=F\delta X\)};
    \draw[red!75!black, ->] (2.44,1.82) -- (2.10,1.47);
    \node at (1.8,-0.55) {current placement};
  \end{scope}
\end{tikzpicture}
\caption{The deformation gradient \(F=T_X\varphi\) is the tangent map of the
deformation. It sends an infinitesimal material vector \(\delta X\) to the
spatial vector \(\delta x=F\delta X\). Strain is obtained by comparing the
lengths and angles of these vectors, not by comparing absolute positions.}
\label{fig:deformation-gradient-tangent-map}
\end{figure}

\section{Polar Decomposition And Objectivity}
\label{sec:polar-decomposition-objectivity}

The deformation gradient contains both rigid rotation and intrinsic stretch.
For \(J=\det F>0\), polar decomposition gives
\[
  \boxed{
  F=R_{\mathrm p}U=VR_{\mathrm p},
  }
\]
where \(R_{\mathrm p}\in\SO(3)\), and \(U,V\) are symmetric positive-definite
stretch tensors. The right stretch \(U\) acts on material vectors before the
polar rotation; the left stretch \(V\) acts on spatial vectors after the
rotation. The right Cauchy-Green tensor removes the rigid rotation:
\[
  C=F^TF=U^2.
\]
Similarly,
\[
  B=FF^T=V^2.
\]

\begin{derivation}[Constructing the polar factors]
Since \(F\) is nonsingular, \(C=F^TF\) is symmetric positive definite. It has a
unique symmetric positive square root
\[
  U=C^{1/2}.
\]
The square root is constructed by the spectral theorem. Write
\[
  C=O\Lambda O^T,\qquad
  \Lambda=\operatorname{diag}(\lambda_1,\lambda_2,\lambda_3),
  \qquad \lambda_i>0,
\]
with \(O\) orthogonal. Define
\[
  C^{1/2}=O\operatorname{diag}(\sqrt{\lambda_1},\sqrt{\lambda_2},\sqrt{\lambda_3})O^T.
\]
This matrix is symmetric positive definite and squares to \(C\). If \(S\) is
another symmetric positive square root, then \(S\) commutes with \(C=S^2\) and
preserves the eigenspaces of \(C\); on each eigenspace positivity selects the
positive square root. Hence \(S=C^{1/2}\).
Define
\[
  R_{\mathrm p}=FU^{-1}.
\]
Then
\[
  R_{\mathrm p}^TR_{\mathrm p}
  =
  U^{-1}F^TFU^{-1}
  =
  U^{-1}U^2U^{-1}
  =
  I.
\]
Because \(\det F>0\) and \(\det U>0\), \(\det R_{\mathrm p}=1\). Thus
\(R_{\mathrm p}\in\SO(3)\). Finally, setting
\[
  V=R_{\mathrm p}UR_{\mathrm p}^T
\]
gives \(F=R_{\mathrm p}U=VR_{\mathrm p}\).
The construction also proves uniqueness: the positive square root \(U=C^{1/2}\)
is unique, so \(R_{\mathrm p}=FU^{-1}\) and \(V=R_{\mathrm p}UR_{\mathrm p}^T\)
are unique for every \(F\) with \(\det F>0\).
\end{derivation}

The eigenvalues of \(U\) are the principal stretches. They are intrinsic
measurements of local length change: a unit material vector in a principal
direction is stretched by \(\lambda_i>0\) before the rigid rotation
\(R_{\mathrm p}\) is applied. The polar rotation is not the same thing as
material spin; it is the rotation factor in the instantaneous deformation
gradient.

This distinction is the doorway to objectivity. A constitutive law for an
ordinary elastic solid may depend on stretch invariants such as those of
\(C\), and it must be invariant under a superposed rigid rotation of the
observer. Polar decomposition makes the separation visible: \(R_{\mathrm p}\)
tells how the local material frame is carried into space, while \(U\) and \(C\)
record the intrinsic stretching that can store elastic energy.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1.0, >=Latex, font=\small]
  \begin{scope}
    \draw[->] (-1.25,0) -- (1.45,0) node[right] {\(E_1\)};
    \draw[->] (0,-1.15) -- (0,1.35) node[above] {\(E_2\)};
    \draw[blue!70!black, thick] (0,0) circle (0.82);
    \draw[blue!70!black, ->] (0,0) -- (0.82,0);
    \draw[blue!70!black, ->] (0,0) -- (0,0.82);
    \node[align=center] at (0,-1.55) {material unit circle};
  \end{scope}
  \draw[->, thick] (1.45,0.2) -- node[above] {\(U\)} (2.85,0.2);
  \begin{scope}[xshift=4.2cm]
    \draw[->] (-1.35,0) -- (1.55,0) node[right] {\(E_1\)};
    \draw[->] (0,-1.15) -- (0,1.35) node[above] {\(E_2\)};
    \draw[green!50!black, thick] (0,0) ellipse (1.12 and 0.55);
    \draw[green!50!black, ->] (0,0) -- (1.12,0)
      node[midway, below] {\(\lambda_1\)};
    \draw[green!50!black, ->] (0,0) -- (0,0.55)
      node[midway, left] {\(\lambda_2\)};
    \node[align=center] at (0,-1.55) {stretch ellipse};
  \end{scope}
  \draw[->, thick] (5.70,0.2) -- node[above] {\(R_{\mathrm p}\)} (7.05,0.2);
  \begin{scope}[xshift=8.6cm, rotate=36]
    \draw[green!50!black, thick] (0,0) ellipse (1.12 and 0.55);
    \draw[green!50!black, ->] (0,0) -- (1.12,0);
    \draw[green!50!black, ->] (0,0) -- (0,0.55);
  \end{scope}
  \begin{scope}[xshift=8.6cm]
    \draw[->] (-1.35,0) -- (1.55,0) node[right] {\(e_1\)};
    \draw[->] (0,-1.15) -- (0,1.35) node[above] {\(e_2\)};
    \draw[red!70!black, very thick, ->] (0.78,-0.05) arc (-5:38:0.78);
    \node[red!70!black, fill=white, inner sep=1pt] at (1.42,0.98) {\(R_{\mathrm p}\)};
    \draw[red!70!black, ->] (1.22,0.82) -- (0.82,0.30);
    \node[align=center] at (0,-1.55) {image under \(F=R_{\mathrm p}U\)};
  \end{scope}
\end{tikzpicture}
\caption{Polar decomposition separates stretch from rigid rotation. In a
principal two-dimensional slice, \(U\) sends the material unit circle to an
ellipse with semiaxes equal to principal stretches; \(R_{\mathrm p}\) then
rotates that stretched figure without changing its lengths.}
\label{fig:polar-decomposition-stretch-rotation}
\end{figure}

This decomposition clarifies frame indifference. If an observer superposes a
rigid rotation \(Q\in\SO(3)\), the deformation gradient becomes \(QF\), but
\[
  (QF)^T(QF)=F^TF=C.
\]
Thus, for an ordinary hyperelastic solid, frame indifference is equivalent to
the statement that the stored energy factors through \(C\):
\[
  W(F,X)=\widehat W(C,X).
\]
The qualifier ``ordinary'' matters: if the material carries additional internal
directors or micro-rotations, as in Cosserat or liquid-crystal models, those
variables must be transformed along with \(F\), and frame indifference is a
statement about all of them together.
The converse is immediate: if \(W(F,X)=\widehat W(F^TF,X)\), then
\[
  W(QF,X)=\widehat W(F^TQ^TQF,X)=\widehat W(F^TF,X)=W(F,X).
\]
For orientation-preserving deformations with \(J>0\), polar decomposition gives
the other direction: \(F=R_{\mathrm p}U\), so frame indifference removes
\(R_{\mathrm p}\) and leaves dependence only on \(U^2=C\).
For an isotropic material, the material has no preferred direction in the
reference body, so \(\widehat W\) may depend on \(C\) only through its principal
invariants:
\[
  I_1=\tr C,\qquad
  I_2=\frac12\left((\tr C)^2-\tr(C^2)\right),\qquad
  I_3=\det C=J^2.
\]
The distinction is important. Frame indifference is invariance under rotations
of the spatial observer. Isotropy is invariance under rotations of the
material reference frame. A crystal can be frame indifferent but not isotropic.

\begin{example}[Simple shear is not a rigid rotation]
For the simple shear
\[
  \varphi(X^1,X^2,X^3)=(X^1+\gamma X^2,X^2,X^3),
\]
the deformation gradient is
\[
  F=
  \begin{pmatrix}
  1&\gamma&0\\
  0&1&0\\
  0&0&1
  \end{pmatrix}.
\]
Although \(\det F=1\), the body is not undeformed:
\[
  C=F^TF=
  \begin{pmatrix}
  1&\gamma&0\\
  \gamma&1+\gamma^2&0\\
  0&0&1
  \end{pmatrix},
  \qquad
  E_{12}=\frac{\gamma}{2}.
\]
The volume is preserved, but right angles in the reference square are changed.
This is the basic solid-mechanics distinction between volume change and shear
strain.
\end{example}

\section{Compatibility}
\label{sec:elastic-compatibility}

Not every matrix field is a deformation gradient. If
\[
  F^i{}_A=\partial_A\varphi^i
\]
for a smooth deformation, then mixed partial derivatives commute:
\[
  \boxed{
  \partial_AF^i{}_B-\partial_BF^i{}_A=0.
  }
\]
On a simply connected body, this curl-free condition is also enough to recover
\(\varphi\) from \(F\), up to a rigid translation.

For small strain, the compatibility question is subtler because \(e\) is only
the symmetric part of \(\nabla\xi\). Given a symmetric tensor field \(e_{ij}\),
we ask whether there exists a displacement \(\xi\) such that
\[
  e_{ij}=\frac12(\partial_i\xi_j+\partial_j\xi_i).
\]
The answer is controlled by the Saint-Venant tensor:
\[
  \boxed{
  \epsilon_{ik\ell}\epsilon_{jmn}
  \partial_k\partial_m e_{\ell n}=0.
  }
\]
Saint-Venant formulated the compatibility condition in 1864; Beltrami later
recast it in the double-curl form in 1886. Its geometric content is that
incompatibility records defects, residual stress, or a non-Euclidean reference
geometry inside the strain field.
This is often written
\[
  \operatorname{curl}\operatorname{curl} e=0.
\]
When the condition fails, the local strain field cannot be assembled into a
single-valued displacement without cuts, defects, residual stress, or a
non-Euclidean reference geometry.

\begin{derivation}[Saint-Venant compatibility as linearized curvature]
A small strain field defines the metric
\[
  g_{ij}=\delta_{ij}+2e_{ij}.
\]
To first order in \(e\), the Christoffel symbols are
\[
  \Gamma^i{}_{jk}
  =
  \partial_j e_{ik}+\partial_k e_{ij}-\partial_i e_{jk},
\]
where indices are raised and lowered with \(\delta_{ij}\). The linearized
Riemann tensor is therefore
\[
  R_{i\ell jm}
  =
  \partial_j\Gamma_{i\ell m}-\partial_m\Gamma_{i\ell j}
  =
  \partial_j\partial_\ell e_{im}
  +\partial_m\partial_i e_{\ell j}
  -\partial_j\partial_i e_{\ell m}
  -\partial_m\partial_\ell e_{ij}.
\]
If \(e_{ij}=\frac12(\partial_i\xi_j+\partial_j\xi_i)\), this curvature
vanishes because mixed derivatives commute. In three dimensions, the
Saint-Venant tensor is the double dual of this linearized curvature:
\[
  (\operatorname{curl}\operatorname{curl}e)_{ij}
  =
  \epsilon_{ik\ell}\epsilon_{jmn}\partial_k\partial_m e_{\ell n}.
\]
Thus the displayed compatibility condition is exactly the linearized flatness
condition.
\end{derivation}

On a simply connected domain the condition is also sufficient. If
\(\operatorname{curl}\operatorname{curl}e=0\), the linearized curvature of
\(g=\delta+2e\) vanishes. Locally there are vector fields \(\xi\) solving
\[
  \partial_i\xi_j+\partial_j\xi_i=2e_{ij}.
\]
The obstruction to patching these local solutions is a rigid motion on each
overlap, because the difference of two solutions has zero symmetric gradient.
On a simply connected body these rigid-motion cocycles can be chosen
consistently, giving a single-valued displacement \(\xi\) unique up to a global
rigid motion. This is the sufficiency part of Saint-Venant compatibility; on a
multiply connected body, periods around noncontractible loops must also vanish.

Here is a more constructive way to read the last sentence. Choose a base point
\(X_0\), prescribe \(\xi(X_0)\) and the antisymmetric part
\(\omega_{ij}(X_0)=-\omega_{ji}(X_0)\), and write
\[
  \partial_i\xi_j=e_{ij}+\omega_{ij}.
\]
Differentiating and eliminating \(\xi\) gives first-order equations for
\(\omega\). The integrability conditions for this first-order system are
precisely \(\operatorname{curl}\operatorname{curl}e=0\). Once those equations
are satisfied, \((\xi,\omega)\) may be transported along any path from \(X_0\).
Path independence on a simply connected body follows from the same flatness
condition. Equivalently, the displacement can be reconstructed by the
Volterra-Cesaro path integral
\[
  \xi_i(X)=\xi_i(X_0)
  +
  \int_{X_0}^{X}
  \bigl(e_{ij}+\omega_{ij}\bigr)\,\dd X^j,
\]
where \(\omega_{ij}\) is transported along the same path by the first-order
system implied by compatibility. Vanishing linearized curvature makes the
integral path independent on a simply connected body. Changing the initial
\(\xi(X_0)\) adds a translation, and changing \(\omega(X_0)\) adds an
infinitesimal rotation. Thus the only freedom left by the compatible strain is
the expected rigid motion.

\section{Hyperelastic Action And Piola Stress}
\label{sec:hyperelastic-action}

A hyperelastic material has stored energy density \(W(F,X)\) per reference
volume. Let \(\rho_0(X)\) be reference mass density. The action is
\[
  S[\varphi]
  =
  \int_{t_0}^{t_1}\int_B
  \left(
    \frac12\rho_0|\partial_t\varphi|^2-W(F,X)
  \right)\dd^3X\,\dd t.
\]

\begin{assumption}[Hyperelastic material]
The stress is derived from stored energy:
\[
  P^A{}_i=\frac{\partial W}{\partial F^i{}_A}.
\]
This assumption excludes hypoelastic, Cauchy-elastic but non-potential,
viscous, plastic, fracture, and history-dependent materials unless additional
internal variables or dissipation laws are added.
\end{assumption}

\begin{derivation}[Elastodynamic equation]
Let \(\delta\varphi=\eta\), with \(\eta=0\) at the time endpoints. Since
\[
  \delta F^i{}_A=\partial_A\eta^i,
\]
the first variation is
\[
\begin{aligned}
  \delta S
  &=
  \int_{t_0}^{t_1}\int_B
  \left(
    \rho_0\dot\varphi_i\dot\eta_i
    -
    P^A{}_i\partial_A\eta^i
  \right)\dd^3X\,\dd t\\
  &=
  \int_{t_0}^{t_1}\int_B
  \left(
    -\rho_0\ddot\varphi_i
    +
    \partial_A P^A{}_i
  \right)\eta^i\dd^3X\,\dd t
  -
  \int_{t_0}^{t_1}\int_{\partial B}
  P^A{}_iN_A\eta^i\,\dd S\,\dd t.
\end{aligned}
\]
For arbitrary interior variations,
\[
  \boxed{
  \rho_0\ddot\varphi_i=\partial_A P^A{}_i.
  }
\]
This is the material form of momentum balance with no body force. A body force
per unit mass \(b_i\) would add \(+\rho_0b_i\) to the right-hand side. If an
external boundary traction \(t_i\) per reference area is included through
virtual work, the natural boundary condition is
\[
  P^A{}_iN_A=t_i.
\]
\end{derivation}

The derivation also explains why \(P\), not \(\sigma\), is the stress that
appears naturally in a reference-domain action: variations are differentiated
with respect to material coordinates \(X^A\), so the boundary term is a force
per reference area.

\section{Spatial Stress And Frame Indifference}
\label{sec:spatial-stress-frame-indifference}

The first Piola-Kirchhoff stress \(P\) measures force per reference area. The
Cauchy stress \(\sigma\) measures force per current spatial area. The two
stress tensors represent the same physical traction, but on different surface
elements.

Two additional stress measures are often useful in finite strain and will
reappear in Appendix~\ref{app:continuum-caveats}. If the stored energy is
written as \(W(F)=\widehat W(C)\), with \(C=F^TF\), the second
Piola-Kirchhoff stress is the material tensor
\[
  S=2\frac{\partial \widehat W}{\partial C},
  \qquad
  P=F S .
\]
It is work-conjugate to the Green strain \(E=(C-I)/2\), because
\[
  P:\dot F=S:\dot E .
\]
The Kirchhoff stress is the spatial tensor
\[
  \tau=J\sigma=P F^T .
\]
Thus \(\sigma\) measures current traction per current area, \(P\) converts
reference surface elements to spatial force, \(S\) pulls both slots back to the
material description, and \(\tau\) is the volume-weighted spatial stress. These
are not four different forces; they are four representations of the same stress
power and traction balance.

Throughout this section, capital indices and normals are material/reference
objects, while lower-case indices and normals are spatial/current objects. In
linearized elasticity the two descriptions coincide to leading order, but the
distinction is essential before linearization.
Cauchy's 1822 tetrahedron argument is the classical reason a traction law
linear in the surface normal must be represented by a stress tensor.

\begin{proposition}[Cauchy's traction theorem]
At a fixed point and time, suppose the traction \(t(n)\) on a plane with unit
normal \(n\) is continuous in \(n\), local, and satisfies force balance on
shrinking material volumes. Then there is a unique tensor \(\sigma\) such that
\[
  t(n)=\sigma n.
\]
\end{proposition}

\begin{proof}
Use Cauchy's tetrahedron. Let the oblique face have area \(A\) and outward
normal \(n\), and let the three coordinate faces have outward normals
\(-e_i\) and areas \(A_i=n_iA\), first for \(n_i\ge0\). Force balance on the
tetrahedron gives
\[
  t(n)A+\sum_i t(-e_i)A_i+O(V)=0,
\]
where \(V\) is the tetrahedron volume. Shrink the tetrahedron while keeping its
shape fixed. Traction terms scale like area and the inertial and body-force
terms scale like volume, so \(O(V)/A\to0\). More explicitly, locality and
continuity imply that the tractions on the shrinking faces differ from their
values at the limiting point by \(o(1)\) uniformly for the finitely many face
normals used in the tetrahedron; after multiplying by areas this gives an
\(o(A)\) correction. The volume force is \(O(h^3)\) for a tetrahedron of size
\(h\), while \(A=O(h^2)\). Dividing by \(A\) and taking \(h\to0\) removes all
non-surface terms. Hence
\[
  t(n)=-\sum_i n_i t(-e_i).
\]
Applying the same argument to a thin pillbox gives \(t(-e_i)=-t(e_i)\).
Therefore \(t(n)=\sum_i n_i t(e_i)\) for normals in the chosen octant. Repeating
the tetrahedron in each octant gives the same formula whenever the signs of the
components of \(n\) are fixed. Continuity of \(t(n)\) on the unit sphere then
extends the formula across the coordinate great circles where one component
vanishes. Define \(\sigma\) by taking its \(i\)-th column to be \(t(e_i)\).
Then \(t(n)=\sigma n\) for every unit normal \(n\), and uniqueness follows by
evaluating the formula on the basis normals \(e_i\).
\end{proof}

\begin{derivation}[Piola transform from traction balance]
Let a material surface element have reference unit normal \(N\) and reference
area \(\dd S\). Its spatial image has unit normal \(n\) and area \(\dd a\).
Nanson's formula states
\[
  n_i\,\dd a
  =
  J(F^{-T})_{Ai}N_A\,\dd S.
\]
The force across the same cut can be written in material variables as
\[
  \dd f_i=P^A{}_iN_A\,\dd S
\]
or in spatial variables as
\[
  \dd f_i=\sigma_{ij}n_j\,\dd a.
\]
Substituting Nanson's formula gives
\[
  P^A{}_iN_A\,\dd S
  =
  \sigma_{ij}J(F^{-T})_{Aj}N_A\,\dd S.
\]
Since \(N_A\) is arbitrary,
\[
  P^A{}_i
  =
  J\sigma_{ij}(F^{-T})_{Aj}.
\]
Multiplying by \(F^k{}_A\) and using
\[
  (F^{-T})_{Aj}F^k{}_A=\delta^k{}_j
\]
gives
\[
  \boxed{
  \sigma_{ij}
  =
  \frac1J P^A{}_iF^j{}_A.
  }
\]
\end{derivation}

A stored energy is frame indifferent if a superposed rigid rotation does not
change it:
\[
  W(QF,X)=W(F,X)\qquad\text{for all }Q\in\SO(3).
\]
Let \(Q(\eps)=I+\eps A+O(\eps^2)\), with \(A^T=-A\). Differentiating
frame indifference at \(\eps=0\) gives
\[
  0=
  P^A{}_iA^i{}_jF^j{}_A.
\]
Since \(A\) is any skew matrix, the tensor \(PF^T\) must be symmetric:
\[
  P^A{}_iF^j{}_A=P^A{}_jF^i{}_A.
\]
Equivalently, if \(B=PF^T-(PF^T)^T\), then in three dimensions
\(B_{ij}=\epsilon_{ijk}b_k\) for some axial vector \(b\). The identity
\(B:A=0\) for every skew \(A\) forces \(b=0\), hence \(B=0\).
Therefore \(\sigma=\sigma^T\). Thus symmetry of stress follows from rotational
invariance of stored energy, matching angular-momentum balance.

When \(W(F,X)=\widehat W(C,X)\), this symmetry can also be seen directly. Since
\[
  \delta C=F^T\delta F+\delta F^TF,
\]
the variation of \(W\) is
\[
  \delta W
  =
  \frac{\partial\widehat W}{\partial C}:\delta C
  =
  2\left(F\frac{\partial\widehat W}{\partial C}\right):\delta F,
\]
where \(\partial\widehat W/\partial C\) is symmetric. Comparing with
\(\delta W=P:\delta F\) gives
\[
  P=2F\frac{\partial \widehat W}{\partial C}.
\]
Because \(\partial\widehat W/\partial C\) is symmetric,
\[
  PF^T=2F\frac{\partial\widehat W}{\partial C}F^T
\]
is symmetric, and therefore so is \(\sigma=J^{-1}PF^T\).

\section{Linear Isotropic Elasticity}
\label{sec:linear-isotropic-elasticity}

For small strain in an isotropic homogeneous solid, the quadratic energy density
is
\[
  W(e)=\frac{\lambda}{2}(\tr e)^2+\mu e_{ij}e_{ij},
\]
where \(\lambda,\mu\) are Lame moduli. In this linearized section, the
reference and spatial Euclidean frames have been identified to first order, so
lower-case indices \(i,j\) are used for components of the displacement gradient
and strain. The stress is
\[
  \sigma_{ij}
  =
  \frac{\partial W}{\partial e_{ij}}
  =
  \lambda(\tr e)\delta_{ij}+2\mu e_{ij}.
\]
In a crystalline solid, one microscopic route to such an energy is the
Cauchy-Born assumption: atoms in a unit cell are assumed to follow the
macroscopic deformation gradient \(F\) affinely, so the lattice energy becomes a
function \(W(F)\). The assumption is useful but not automatic; defects,
multi-lattices, phase transitions, and short-wavelength instabilities can make
the continuum energy depart from the affine lattice prediction.
The linearized equation of motion is
\[
  \rho_0\partial_t^2\xi_i=\partial_j\sigma_{ij}+\rho_0 b_i.
\]
Substituting the stress-strain relation,
\[
\begin{aligned}
  \partial_j\sigma_{ij}
  &=
  \lambda\partial_i(\partial_k\xi_k)
  +
  \mu\partial_j(\partial_i\xi_j+\partial_j\xi_i)\\
  &=
  (\lambda+\mu)\partial_i(\nabla\cdot\xi)
  +
  \mu\Delta\xi_i.
\end{aligned}
\]
Hence
\[
  \boxed{
  \rho_0\partial_t^2\xi
  =
  (\lambda+\mu)\nabla(\nabla\cdot\xi)
  +
  \mu\Delta\xi
  +
  \rho_0 b.
  }
\]
Using
\[
  \Delta\xi=\nabla(\nabla\cdot\xi)-\nabla\times\nabla\times\xi,
\]
one sees two wave families:
\[
  c_P=\sqrt{\frac{\lambda+2\mu}{\rho_0}},
  \qquad
  c_S=\sqrt{\frac{\mu}{\rho_0}}.
\]
Longitudinal waves change volume; shear waves change shape at nearly fixed
volume.

\begin{derivation}[Plane-wave split into \(P\) and \(S\) modes]
Set \(b=0\) and try a plane wave
\[
  \xi(X,t)=a\,e^{i(k\cdot X-\omega t)},
\]
where \(a\) is a constant polarization vector and \(k\) is the wave vector.
Then
\[
  \nabla(\nabla\cdot\xi)=-k(k\cdot a)e^{i(k\cdot X-\omega t)},
  \qquad
  \Delta\xi=-|k|^2a\,e^{i(k\cdot X-\omega t)}.
\]
Substituting into the linear elastodynamic equation gives
\[
  \rho_0\omega^2 a
  =
  (\lambda+\mu)k(k\cdot a)+\mu |k|^2a.
\]
If \(a\parallel k\), then \(k(k\cdot a)=|k|^2a\), so
\[
  \omega^2=c_P^2|k|^2,
  \qquad
  c_P^2=\frac{\lambda+2\mu}{\rho_0}.
\]
If \(a\perp k\), then \(k\cdot a=0\), so
\[
  \omega^2=c_S^2|k|^2,
  \qquad
  c_S^2=\frac{\mu}{\rho_0}.
\]
Thus the eigenvectors of the acoustic tensor split into one longitudinal
polarization and two transverse polarizations in an isotropic solid.
\end{derivation}

Wave-speed positivity is strictly weaker than energy convexity. Propagating
plane waves require \(c_P^2>0\) and \(c_S^2>0\), namely
\(\mu>0\) and \(\lambda+2\mu>0\). Positive definiteness of the quadratic
elastic energy requires the stronger-looking but different conditions
\(\mu>0\) and \(K_{\mathrm{bulk}}=\lambda+\frac23\mu>0\). The distinction
foreshadows the hierarchy in finite elasticity: strong ellipticity tests the
second variation on rank-one perturbations \(a\otimes n\), rank-one convexity
and quasiconvexity are stronger global conditions, and polyconvexity is a
checkable sufficient condition often used in existence theory. These are not
cosmetic terms: a stored energy can be frame indifferent and support real
small-amplitude waves while still failing to be globally convex as a function
of all entries of \(F\).
The Lame constants entered elasticity through nineteenth-century work on
isotropic solids; the \(P\)- and \(S\)-wave names are seismological terminology
for the same longitudinal/transverse split. Oldham (1906) used the \(S\)-wave
shadow zone to infer that Earth has a fluid outer core.

\section{Elastic Constants And Classical Boundary-Value Problems}
\label{sec:elastic-constants-boundary-problems}

Stiffness laws used in the lab,
\[
  F=\frac{E_{\mathrm Y}A_{\mathrm{cs}}}{L}\Delta L,
  \qquad
  T=\mu J_t\,\Theta',
  \qquad
  E_{\mathrm Y}I_{\mathrm{area}}w''''=q,
\]
are reductions of the same isotropic equations. This section derives the
standard bar, shaft, and beam formulas and tracks which modulus each experiment
measures. The notation \(E_{\mathrm Y}\), \(\nu_{\mathrm P}\), \(\lambda\),
\(\mu\), and \(K_{\mathrm{bulk}}\) is therefore not interchangeable; each symbol
names a different response.

Saint-Venant's principle gives the modelling justification behind many of the
one-dimensional reductions below. Details of how a load is applied at an end
are expected to decay over distances comparable to the cross-section size, so a
slender body can have a robust interior bar, shaft, or beam solution even when
the end fixtures are complicated. This is an asymptotic modelling principle,
not an exact theorem for every geometry.

\subsection{Moduli, Positivity, And Nearly Incompressible Solids}
\label{subsec:elastic-moduli-positivity}

The quadratic energy
\[
  W(e)=\frac{\lambda}{2}(\tr e)^2+\mu e:e
\]
becomes transparent after splitting strain into volume-changing and
shape-changing parts. Define
\[
  e_0=e-\frac13(\tr e)I,
  \qquad
  K_{\mathrm{bulk}}=\lambda+\frac23\mu .
\]
Since
\[
  e:e=e_0:e_0+\frac13(\tr e)^2,
\]
the energy becomes
\[
  \boxed{
  W(e)=\mu\,e_0:e_0+\frac12K_{\mathrm{bulk}}(\tr e)^2.
  }
\]
The shear modulus \(\mu\) penalizes distortion at fixed volume. The bulk
modulus \(K_{\mathrm{bulk}}\) penalizes volume change. For a stable isotropic
linear elastic solid one requires
\[
  \mu>0,
  \qquad
  K_{\mathrm{bulk}}>0.
\]
These inequalities have two related meanings in the linear theory. They make
the quadratic energy positive on pure shear strains and pure volumetric
strains, after rigid translations and rotations have been removed. They also
make the acoustic wave speeds real, because the shear and longitudinal wave
speeds below contain \(\mu\) and
\(\lambda+2\mu=K_{\mathrm{bulk}}+\frac43\mu\). For finite-strain elasticity,
however, stability is subtler: strong ellipticity, rank-one convexity,
quasiconvexity, and polyconvexity are distinct conditions. The linear isotropic
model collapses these distinctions to the simple positive-definiteness test
above, which is why the small-strain theory is so transparent.

Engineers often use Young's modulus \(E_{\mathrm Y}\) and Poisson's ratio
\(\nu_{\mathrm P}\) instead of \(\lambda,\mu\). The conversion is
\[
  \boxed{
  E_{\mathrm Y}
  =
  \mu\frac{3\lambda+2\mu}{\lambda+\mu},
  \qquad
  \nu_{\mathrm P}
  =
  \frac{\lambda}{2(\lambda+\mu)}.
  }
\]
Equivalently,
\[
  \boxed{
  \mu=\frac{E_{\mathrm Y}}{2(1+\nu_{\mathrm P})},
  \qquad
  \lambda=
  \frac{E_{\mathrm Y}\nu_{\mathrm P}}
  {(1+\nu_{\mathrm P})(1-2\nu_{\mathrm P})},
  \qquad
  K_{\mathrm{bulk}}=
  \frac{E_{\mathrm Y}}{3(1-2\nu_{\mathrm P})}.
  }
\]
Positive energy is equivalent to
\[
  E_{\mathrm Y}>0,
  \qquad
  -1<\nu_{\mathrm P}<\frac12.
\]
Most ordinary solids have \(0<\nu_{\mathrm P}<1/2\): they contract laterally
when stretched. The limit \(\nu_{\mathrm P}\to1/2\) is the nearly
incompressible limit. In that limit \(\lambda\) and \(K_{\mathrm{bulk}}\) grow
large while \(\mu\) can remain finite, and the equations begin to resemble
incompressible fluid mechanics: pressure-like stresses enforce a volume
constraint.

\subsection{Static Navier-Cauchy Equation And Boundary Data}
\label{subsec:static-navier-cauchy}

In static linear elasticity, inertia is absent and the displacement
\(\xi(X)\) satisfies
\[
  \boxed{
  (\lambda+\mu)\nabla(\nabla\cdot\xi)+\mu\Delta\xi+\rho_0 b=0.
  }
\]
Equivalently,
\[
  \partial_j\sigma_{ij}+\rho_0b_i=0,
  \qquad
  \sigma_{ij}=\lambda(\partial_k\xi_k)\delta_{ij}
  +\mu(\partial_i\xi_j+\partial_j\xi_i).
\]
This is the static Navier-Cauchy equation, due to Navier (1821) in a molecular
formulation and Cauchy (1822) in the stress-tensor formulation.
Boundary conditions are part of the problem statement. On a displacement
boundary \(\partial B_D\), one prescribes
\[
  \xi=\xi_D.
\]
On a traction boundary \(\partial B_N\), one prescribes
\[
  \sigma n=t_N,
\]
where \(n\) is the outward unit normal to the reference boundary, to linear
order. The pure traction problem determines displacement only up to rigid
translations and rotations; these rigid modes must be fixed by constraints or
by imposing zero mean displacement and zero mean rotation.

The weak form follows from virtual work. For any test displacement \(\eta\)
vanishing on \(\partial B_D\),
\[
  \int_B \sigma(\xi):e(\eta)\,\dd^3X
  =
  \int_B \rho_0 b\cdot\eta\,\dd^3X
  +
  \int_{\partial B_N} t_N\cdot\eta\,\dd S.
\]
This is the elastic analogue of deriving Cauchy's equation from balance laws:
internal stress work equals external virtual work.

The weak form also gives uniqueness in the standard coercive case. If the body
has a nonempty displacement boundary, or if rigid translations and rotations
are otherwise fixed, the quadratic elastic energy is strictly convex on the
admissible displacement space. Korn's 1909 inequality is the missing analytic
bridge:
\[
  \|\xi\|_{H^1(B)}
  \le C\left(\|e(\xi)\|_{L^2(B)}+\|\xi\|_{L^2(B)}\right),
\]
which is often called Korn's second inequality. If \(\mathcal R\) denotes the
finite-dimensional space of infinitesimal rigid motions
\(r(X)=a+\Omega X\), \(\Omega^T=-\Omega\), the sharper rigid-motion form is
\[
  \inf_{r\in\mathcal R}\|\xi-r\|_{H^1(B)}
  \le C\|e(\xi)\|_{L^2(B)}.
\]
After fixing the rigid part by displacement boundary conditions, zero mean
translation and rotation, or another side condition, this becomes
\(\|\xi\|_{H^1}\le C\|e(\xi)\|_{L^2}\). The inequality is an
integration-by-parts statement that the skew part of \(\nabla\xi\) is controlled
by the symmetric part once rigid rotations are excluded; modern proofs are
standard in the Korn-Friedrichs theory of elasticity. With Korn's inequality,
the bilinear elastic energy is coercive, and the Lax-Milgram theorem gives a
unique weak solution for compatible loads. In a pure traction problem,
uniqueness holds only modulo rigid motions, which must be removed by side
constraints.

\subsection{Uniaxial Extension}
\label{subsec:uniaxial-extension}

Consider a slender bar of length \(L\), cross-sectional area \(A_{\mathrm{cs}}\),
and Young modulus \(E_{\mathrm Y}\). A tensile force \(F_{\mathrm{end}}\) is
applied at the end, and lateral surfaces are traction-free. The axial stress is
\[
  \sigma_{11}=\frac{F_{\mathrm{end}}}{A_{\mathrm{cs}}},
  \qquad
  \sigma_{22}=\sigma_{33}=0.
\]
Hooke's law in the \(E_{\mathrm Y},\nu_{\mathrm P}\) form gives
\[
  e_{11}=\frac{\sigma_{11}}{E_{\mathrm Y}},
  \qquad
  e_{22}=e_{33}=-\nu_{\mathrm P}\frac{\sigma_{11}}{E_{\mathrm Y}}.
\]
Thus the axial elongation is
\[
  \boxed{
  \Delta L
  =
  \frac{F_{\mathrm{end}}L}{E_{\mathrm Y}A_{\mathrm{cs}}}.
  }
\]
The negative lateral strain is the definition of Poisson's ratio:
\[
  \nu_{\mathrm P}
  =
  -\frac{\text{lateral strain}}{\text{axial strain}}.
\]
This formula identifies the experimental meaning of Young's modulus: it is
stress divided by axial strain in a uniaxial test with free lateral
contraction.
The appearance of Poisson's ratio is also a historical caution: early
``rari-constant'' elasticity in the Cauchy-Poisson tradition overconstrained
the moduli. Modern isotropic linear elasticity needs two independent elastic
constants.

\subsection{Torsion Of A Circular Shaft}
\label{subsec:circular-shaft-torsion}

Let a circular shaft of radius \(a\) and length \(L\) be twisted by torque
\(T_{\mathrm{end}}\). In the Saint-Venant torsion solution, cross-sections
rotate by an angle \(\Theta(z)\) about the shaft axis. For a circular cross
section, there is no warping correction, and the twist per length is constant:
\[
  \Theta'(z)=\frac{T_{\mathrm{end}}}{\mu J_{\mathrm{polar}}},
  \qquad
  J_{\mathrm{polar}}=\frac{\pi a^4}{2}.
\]
The total twist angle is
\[
  \boxed{
  \Theta(L)-\Theta(0)=
  \frac{T_{\mathrm{end}}L}{\mu J_{\mathrm{polar}}}.
  }
\]
The shear stress grows linearly with radius:
\[
  \tau(r)=\frac{T_{\mathrm{end}}r}{J_{\mathrm{polar}}},
  \qquad
  0\leq r\leq a.
\]
The maximum shear stress occurs at the boundary \(r=a\). This is the torsional
counterpart of the axial bar formula: stiffness is the elastic modulus times a
geometric moment of the cross-section.
Circular shafts are special because warping vanishes. Saint-Venant's general
torsion theory allows non-circular sections, but then the cross-section warps and
the simple polar-moment formula must be replaced by a torsion function.
For a non-circular cross-section \(A\), take the shaft axis to be \(z\) and let
\(\tau=\Theta'(z)\) be the twist per length. A Saint-Venant ansatz is
\[
  u_x=-\tau zy,\qquad
  u_y=\tau zx,\qquad
  u_z=\tau\chi(x,y),
\]
where \(\chi\) is the warping function. The only nonzero shear stresses are
\[
  \sigma_{xz}=\mu\tau(\partial_x\chi-y),
  \qquad
  \sigma_{yz}=\mu\tau(\partial_y\chi+x).
\]
Equilibrium \(\partial_x\sigma_{xz}+\partial_y\sigma_{yz}=0\) gives
\[
  \Delta\chi=0\quad\text{in }A,
\]
and the traction-free lateral boundary condition
\((\sigma_{xz},\sigma_{yz})\cdot n=0\) gives
\[
  \frac{\partial\chi}{\partial n}=yn_x-xn_y
  \quad\text{on }\partial A.
\]
The torque is \(T_{\mathrm{end}}=\mu\tau J_t\), where the torsional rigidity is
computed from the shear-energy integral
\[
  J_t=
  \int_A
  \left[(\partial_x\chi-y)^2+(\partial_y\chi+x)^2\right]\,\dd A .
\]
For a circular section, \(\chi=0\) satisfies the boundary condition and
\(J_t=J_{\mathrm{polar}}\). For a non-circular section, warping changes the
rigidity and the stress distribution.

\subsection{Euler-Bernoulli Beam Bending}
\label{subsec:euler-bernoulli-beam}

The model is due to Daniel Bernoulli's elastic-energy viewpoint, around 1742,
and Euler's 1744 \emph{De curvis elasticis}. Its assumptions are deliberate: later
Timoshenko theory restores shear deformation and rotary inertia when the beam
is not sufficiently slender.
As an asymptotic reduction from three-dimensional elasticity, it assumes a
slenderness parameter \(h/L\ll1\), where \(h\) is a cross-section diameter.
Keeping the leading bending energy gives strains of order \(h\,w''\); transverse
shear strains and rotary inertia enter at higher order in \(h/L\). This is why
the model is excellent for long thin beams and unreliable for short thick ones.

For a slender beam, let \(x\in[0,L]\) be arclength along the undeformed neutral
axis and let \(w(x)\) be a small transverse deflection. The Euler-Bernoulli
assumptions are:
\begin{enumerate}
\item slopes are small, so curvature is \(w''(x)\) to leading order;
\item plane cross-sections remain plane and normal to the neutral axis;
\item shear deformation and rotary inertia are neglected.
\end{enumerate}
Under these assumptions the axial displacement at height \(y\) from the neutral
axis is \(u_x(x,y)=-y\,w'(x)\) to leading order. Hence the dominant axial strain
is
\[
  e_{xx}=\partial_xu_x=-y\,w''(x).
\]
The three-dimensional elastic energy density
\(\frac12 E_{\mathrm Y}e_{xx}^2\) therefore reduces to
\[
  \frac12 E_{\mathrm Y}(w'')^2\int_A y^2\,\dd A
  =
  \frac12 E_{\mathrm Y}I_{\mathrm{area}}(w'')^2,
\]
where \(I_{\mathrm{area}}=\int_A y^2\,\dd A\). This is the leading
three-dimensional origin of the one-dimensional bending energy.
The bending energy is
\[
  U_{\mathrm{bend}}
  =
  \frac12\int_0^L E_{\mathrm Y}I_{\mathrm{area}}(w'')^2\,\dd x,
\]
where \(I_{\mathrm{area}}\) is the second moment of area of the cross-section.
With transverse distributed load \(q(x)\), the potential energy is
\[
  \Pi[w]
  =
  \frac12\int_0^L E_{\mathrm Y}I_{\mathrm{area}}(w'')^2\,\dd x
  -
  \int_0^L q(x)w(x)\,\dd x.
\]
Varying \(w\) gives
\[
\begin{aligned}
  \delta\Pi
  &=
  \int_0^L E_{\mathrm Y}I_{\mathrm{area}}w''\eta''\,\dd x
  -
  \int_0^L q\eta\,\dd x\\
  &=
  \int_0^L
  \left(E_{\mathrm Y}I_{\mathrm{area}}w''''-q\right)\eta\,\dd x
  +
  \left[
    E_{\mathrm Y}I_{\mathrm{area}}w''\eta'
    -
    E_{\mathrm Y}I_{\mathrm{area}}w'''\eta
  \right]_{0}^{L}.
\end{aligned}
\]
The beam equation is therefore
\[
  \boxed{
  E_{\mathrm Y}I_{\mathrm{area}}w''''=q.
  }
\]
The natural boundary quantities are the bending moment
\[
  M=E_{\mathrm Y}I_{\mathrm{area}}w''
\]
and shear force
\[
  V_{\mathrm{shear}}=-E_{\mathrm Y}I_{\mathrm{area}}w'''.
\]

For a cantilever clamped at \(x=0\) and loaded by a downward end force
\(P_{\mathrm{end}}\) at \(x=L\), the boundary conditions are
\[
  w(0)=0,\qquad w'(0)=0,\qquad
  M(L)=0,\qquad V_{\mathrm{shear}}(L)=P_{\mathrm{end}}.
\]
With \(q=0\) in the interior, the solution is
\[
  \boxed{
  w(x)=
  \frac{P_{\mathrm{end}}x^2(3L-x)}
  {6E_{\mathrm Y}I_{\mathrm{area}}}.
  }
\]
The tip deflection is
\[
  \boxed{
  w(L)=\frac{P_{\mathrm{end}}L^3}
  {3E_{\mathrm Y}I_{\mathrm{area}}}.
  }
\]
For a simply supported beam under uniform load \(q_0\), the deflection is
\[
  \boxed{
  w(x)=
  \frac{q_0 x(L^3-2Lx^2+x^3)}
  {24E_{\mathrm Y}I_{\mathrm{area}}}.
  }
\]
It satisfies \(w(0)=w(L)=0\), zero bending moment at both supports, and
\(E_{\mathrm Y}I_{\mathrm{area}}w''''=q_0\).

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1.0, >=Latex, font=\small]
  \begin{scope}
    \draw[fill=blue!8, thick] (0,0.65) rectangle (3.5,1.15);
    \draw[thick] (0,0.55) -- (0,1.25);
    \draw[thick, ->] (3.5,0.9) -- (4.3,0.9)
      node[right, fill=white, inner sep=1pt] {\(F_{\mathrm{end}}\)};
    \draw[<->] (0,0.25) -- node[below, fill=white, inner sep=1pt] {\(L\)} (3.5,0.25);
    \node[align=center, fill=white, inner sep=1.5pt] at (1.75,1.72)
      {uniaxial bar\\\(\Delta L=F_{\mathrm{end}}L/(E_{\mathrm Y}A_{\mathrm{cs}})\)};
  \end{scope}
  \begin{scope}[xshift=5.7cm]
    \draw[line width=1.0pt] (0,1.28) -- (0,0.15);
    \foreach \y in {0.20,0.42,0.64,0.86,1.08,1.30}
      \draw[line width=0.8pt] (-0.42,\y-0.22) -- (0,\y);
    \draw[blue!70!black, thick, domain=0:3.6, samples=100]
      plot (\x,{1.05-0.055*\x*\x*(3.6-\x/3)});
    \draw[red!75!black, thick, ->] (3.92,0.66) -- (3.92,-0.08);
    \node[red!75!black, anchor=west, fill=white, inner sep=1.5pt]
      at (4.03,0.29) {\(P_{\mathrm{end}}\)};
    \draw[<->] (0,0.02) -- node[below, fill=white, inner sep=1pt] {\(L\)} (3.6,0.02);
    \draw[densely dashed] (3.6,1.05) -- (3.6,0.34);
    \draw[<->, purple!75!black] (3.35,1.05) -- (3.35,0.34)
      node[midway, left, fill=white, inner sep=1pt] {\(w(L)\)};
    \node[align=center, fill=white, inner sep=1.5pt] at (1.85,1.72)
      {cantilever beam\\\(w(L)=P_{\mathrm{end}}L^3/(3E_{\mathrm Y}I_{\mathrm{area}})\)};
  \end{scope}
\end{tikzpicture}
\caption{Two classical reductions of three-dimensional linear elasticity. A
slender bar under axial load measures Young's modulus. A slender beam under
transverse load converts elastic energy in curvature into the fourth-order
Euler-Bernoulli equation.}
\label{fig:bar-and-beam-elasticity}
\end{figure}

\section{What To Take Forward}
\label{sec:elastic-body-take-forward}

An elastic body is described by a deformation map \(\varphi(X,t)\) from a
material reference body \(B\) into physical space. The deformation gradient
\(F^i{}_A=\partial\varphi^i/\partial X^A\) measures local stretch and rotation,
while objective strain tensors remove the arbitrary rigid rotation.
Compatibility says that not every field of strain is the strain of a
single-valued displacement; it must arise from a deformation with commuting
mixed derivatives, or equivalently satisfy the Saint-Venant conditions in the
linear theory.

The variational derivation is parallel to finite-dimensional Lagrangian
mechanics but with fields. A hyperelastic stored energy \(W(F,X)\) defines the
first Piola-Kirchhoff stress \(P^i{}_A=\partial W/\partial F^i{}_A\), and
stationarity of the action gives material momentum balance
\(\rho_0\ddot\varphi^i=\partial_A P^i{}_A+\rho_0 b^i\) together with the
correct traction boundary term. Frame indifference forces the stored energy to
depend on deformation through rotationally invariant combinations such as
\(C=F^T F\), and it implies symmetry of the spatial Cauchy stress.

Linear isotropic elasticity is the small-strain limit. Its two Lam\'e constants
encode the quadratic energy, the Navier-Cauchy equation describes displacement
waves and static equilibrium, and the classical bar, torsion, and beam problems
show how continuum equations reduce to experimentally measurable stiffness
laws. The same material description also has a frame-theoretic extension:
once each material point is allowed its own local orientation, elastic
kinematics becomes the material side of a gauge theory of deforming bodies.

\section{Chapter-End Laboratories And Exercises}
\label{sec:elastic-body-chapter-end-labs}

The chapter-end work is organized as an audit of the elastic modelling chain:
objectivity, variational stress, linearization, and classical reductions. The
questions are short, but each one points to a place where continuum mechanics
often becomes confusing if one forgets which configuration a tensor belongs to.

\begin{enumerate}
\item Starting from \(C=F^TF\), show explicitly that superposing a rigid spatial
rotation on \(\varphi\) leaves \(C\) unchanged. Which strain tensors would fail
this objectivity test?
\item In the hyperelastic action, identify the boundary term whose coefficient
is the first Piola-Kirchhoff traction. How does that term transform into a
spatial Cauchy traction?
\item Derive the Navier-Cauchy equation from the quadratic isotropic energy and
identify where the two Lame constants enter longitudinal and transverse waves.
\item For the cantilever beam formula, check the four boundary conditions at the
clamped and loaded ends. Which condition represents the applied point load?
\item Compare the bar and beam reductions in Figure~\ref{fig:bar-and-beam-elasticity}.
Which geometric small parameter justifies replacing a three-dimensional elastic
body by a one-dimensional model?
\end{enumerate}

\subsection{Boundary-Value Laboratory: Bar Versus Beam}
\label{sec:elastic-bar-beam-lab}

Use Figure~\ref{fig:bar-and-beam-elasticity} as a modelling checklist. The bar
problem assumes that the dominant strain is axial and approximately uniform
over each cross-section, giving the stiffness law \(F=EA\,\Delta L/L\). The
beam problem assumes that cross-sections remain approximately plane and that
curvature, not axial stretch, carries the leading elastic energy, giving
\[
  EI\,w''''(x)=q(x).
\]
Both are one-dimensional reductions of the same three-dimensional elastic
theory. The useful comparison is therefore not ``bar formula versus beam
formula,'' but which geometric small parameter justifies throwing away the
other strain components. A good solution should state the slenderness
assumption, the boundary data, the stress or moment being measured, and the
energy term that survives the reduction.

\section{Associated Demos}
\label{sec:elastic-body-demos}

The associated elastic-boundary-value demos are:
\begin{itemize}
\item \path{demos/python/linear_elasticity.py}
\item \path{demos/mathematica/LinearElasticity.wl}
\end{itemize}
They solve elementary elastic boundary-value problems and visualize
displacement, strain, and stress fields. They are designed to be read beside
this chapter's derivations of the Piola stress, Cauchy stress, and linear
elastic equations.
