\chapter{Gauge Theory Of Deforming Bodies}
\label{chap:deforming-body}

A cat falls. It rights itself in mid-air without any external torque. Its
angular momentum can remain zero while its orientation changes, because the cat
cycles through internal shapes. Related geometric-phase mechanisms appear in
low-Reynolds swimming, astronaut reorientation, and the rotor laboratory below:
closed loops in shape space produce rigid motions in physical space, although
the constraints and shape variables differ from one system to another.

Frank Wilczek's gauge-theoretic viewpoint, developed with Shapere in the
context of swimming and self-propulsion,\footnote{A. Shapere and F. Wilczek,
``Gauge kinematics of
deformable bodies,'' \emph{American Journal of Physics} 57, 514--518 (1989),
\href{https://doi.org/10.1119/1.15986}{doi:10.1119/1.15986}. Their closely
related low-Reynolds-number swimming formulation appears in
\href{https://doi.org/10.1017/S002211208900025X}{doi:10.1017/S002211208900025X}.}
treats the decomposition into overall rigid motion and internal deformation as
a choice of gauge.  The connection tells us which rigid motion accompanies an
infinitesimal change of shape under a specified momentum constraint, and its
curvature and holonomy describe the observable geometric phase.

The central phenomenon is geometric phase.  A body can execute a closed loop in
shape space and return to the same internal shape while acquiring a net rotation
or translation in physical space.  This is the mechanical analogue of parallel
transport in a gauge field.  The curvature of the connection measures the
leading displacement produced by small loops, while the full holonomy records
the finite motion produced by large cycles.

\paragraph{Roadmap.}
The chapter starts with the finite-dimensional Shapere-Wilczek split between
rigid placement \(g\) and shape \(s\). It then derives the mechanical
connection from zero momentum and explains how this connection transforms under
changes of body frame. The two-rotor model is a minimal laboratory for
noncommuting connection components. The last third returns to continuum
mechanics: local frames, Cosserat rods, curvature, torsion, and defects show
how the same gauge language becomes a field theory over a material body.

\paragraph{Reading rule.}
There are two meanings of ``gauge'' in this chapter. One is finite-dimensional:
choosing body axes over shape space. The other is local and continuum: choosing
a frame at each material point. Both are frame choices, and both have
curvatures, but they live over different base spaces. Whenever a symbol
\(\mathcal A\), \(A\), \(\omega\), or \(R\) appears, first identify whether the
base variable is shape \(s\), time \(t\), or material position \(X\).

Appendix~\ref{app:gauge-taxonomy} systematizes this distinction after the
examples have been developed. It separates mechanical connections,
Shapere-Wilczek swimming connections, Cosserat frame fields, defect geometry,
and fluid relabeling symmetry so that the same word ``gauge'' does not blur
different mechanisms.

\section{Gauge Kinematics Of Deformable Bodies}
\label{sec:gauge-kinematics-deformable-bodies}

Elasticity describes deformation relative to a reference body. Gauge kinematics
adds another question: how is the same changing shape placed and oriented in
space? A choice of body axes at each shape gives a local description, while
holonomy around a closed loop records the observable rigid motion produced by
the shape cycle.

The Wilczek-Shapere point of view begins with a finite deformable body moving in
space. Separate its configuration into:
\[
  g\in G,\qquad s\in\mathcal S,
\]
where \(G\) is the rigid-motion group, such as \(\SE(3)\) or \(\SO(3)\), and
\(\mathcal S\) is shape space. A point of the full configuration space is a
rigid placement \(g\) together with an internal shape \(s\).

The subtlety is that the split into \(g\) and \(s\) is not unique. To say what
the body's orientation is at a given shape, one must choose body axes for that
shape. A shape-dependent change of axes
\[
  h:\mathcal S\to G
\]
changes the representative \(g\). At this stage this is best described as the
analogue of a gauge choice: the full gauge-theoretic content appears once the
connection, curvature, and holonomy transformation laws are introduced below.

Geometrically, the full configuration space is locally a product
\[
  Q_{\mathrm{full}}\simeq G\times\mathcal S,
\]
but no preferred product splitting is supplied by bare kinematics. A kinetic
energy metric on \(Q_{\mathrm{full}}\), together with a momentum constraint,
does pick out a distinguished splitting; that distinguished splitting is the
mechanical connection. The projection
\[
  Q_{\mathrm{full}}\to\mathcal S
\]
forgets the rigid placement and remembers only shape. A connection is a rule
for deciding which infinitesimal motions in \(Q_{\mathrm{full}}\) count as
``pure shape change'' and which contain rigid motion. In mechanics this rule is
not arbitrary: the mechanical connection is selected by the kinetic energy and
the momentum constraint.

The observable content is therefore not the connection coefficient by itself.
A different convention for body axes changes the coefficient, just as a
different electromagnetic gauge changes the vector potential. What survives is
the reconstructed rigid motion, especially the holonomy around a closed shape
cycle and the curvature that controls small cycles.

\subsection{Principal-Bundle Form Of The Split}
\label{subsec:deforming-body-principal-bundle}

The geometric content of the split is a principal bundle.  Let
\(Q_{\mathrm{full}}\)
be the full configuration space of the deforming body, after any constraints
such as fixed center of mass have been imposed.  The rigid-motion group \(G\)
acts on \(Q_{\mathrm{full}}\) by changing the overall placement of the same
internal shape.  The shape space is the quotient
\[
  \mathcal S=Q_{\mathrm{full}}/G,
  \qquad
  \pi:Q_{\mathrm{full}}\to\mathcal S .
\]
The point \(s=\pi(q)\) remembers the shape and forgets the rigid placement.

\begin{assumption}[Regular shape bundle]
In this chapter the \(G\)-action is assumed to be free and proper on the
configurations under discussion, so that \(\mathcal S\) is a smooth manifold
and \(Q_{\mathrm{full}}\to\mathcal S\) is locally a product.  Shapes with
extra symmetry, such as a perfectly spherical body, have stabilizers and require
minor orbifold or stratified-space modifications.  The formulas below describe
the generic regular stratum.
\end{assumption}

A local choice of body frame is a local section
\[
  \sigma:\mathcal U\subset\mathcal S\to Q_{\mathrm{full}}.
\]
Every nearby configuration can then be written as
\[
  q=g\,\sigma(s),
  \qquad g\in G,\quad s\in\mathcal U.
\]
This expression is not unique because the section is not unique.  If the
internal representative is changed by a shape-dependent rigid motion
\[
  \sigma'(s)=h(s)^{-1}\sigma(s),
  \qquad h:\mathcal U\to G,
\]
then the same physical configuration is represented by
\[
  q=g\,\sigma(s)=g'\sigma'(s),
  \qquad
  g'=g\,h(s).
\]
Thus a change of section changes the coordinate \(g\) even though the physical
configuration \(q\) has not changed.  This is the precise analogue of a gauge
transformation.

A connection is a \(G\)-equivariant rule that splits a tangent vector
\(\dot q\in T_qQ_{\mathrm{full}}\) into a vertical rigid part and a horizontal
shape-changing part.  In the local coordinates \((g,s)\), write the body-frame
group velocity as
\[
  \xi=g^{-1}\dot g\in\mathfrak g .
\]
The horizontal condition has the form
\[
  \xi+\mathcal A_\alpha(s)\dot s^\alpha=0.
\]
Therefore a horizontal lift of a shape path obeys the reconstruction equation
\[
  g^{-1}\dot g=-\mathcal A_\alpha(s)\dot s^\alpha .
\]
The local coefficient \(\mathcal A=\mathcal A_\alpha ds^\alpha\) is the gauge
potential.  Under the change of section above it transforms as
\[
  \mathcal A'
  =
  \Ad_{h^{-1}}\mathcal A-h^{-1}\dd h,
\]
which is exactly the inhomogeneous transformation law expected of a connection.
A connection supplies a chosen horizontal subspace, so the phrase ``pure shape
change'' has a specified meaning. The mechanical connection is the
particular connection selected by the kinetic-energy metric and the momentum
value.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1.0, >=Latex, font=\small]
  \draw[->] (-0.2,0) -- (6.5,0) node[right] {\(s^1\)};
  \draw[->] (0,-0.2) -- (0,4.1) node[above] {\(s^2\)};
  \draw[thick, blue, ->] (1.2,1.2) .. controls (2.8,3.4) and (5.0,3.1) .. (5.0,1.4);
  \draw[thick, blue, ->] (5.0,1.4) .. controls (3.9,0.4) and (2.2,0.3) .. (1.2,1.2);
  \node[blue, fill=white, inner sep=2pt] at (3.05,3.70) {shape loop};
  \begin{scope}[shift={(1.2,1.2)}, scale=0.46]
    \draw[fill=gray!20, thick] (-0.8,0) ellipse (0.45 and 0.25);
    \draw[fill=gray!20, thick] (0.8,0) ellipse (0.45 and 0.25);
    \draw[thick] (-0.35,0) -- (0.35,0);
  \end{scope}
  \begin{scope}[shift={(5.0,1.4)}, scale=0.46, rotate=-45]
    \draw[fill=gray!20, thick] (-1.0,0) ellipse (0.55 and 0.18);
    \draw[fill=gray!20, thick] (1.0,0) ellipse (0.55 and 0.18);
    \draw[thick] (-0.45,0.05) -- (0.45,-0.05);
  \end{scope}
  \begin{scope}[shift={(5.35,3.05)}, scale=0.46, rotate=70]
    \draw[fill=gray!20, thick] (-0.8,0) ellipse (0.45 and 0.25);
    \draw[fill=gray!20, thick] (0.8,0) ellipse (0.45 and 0.25);
    \draw[thick] (-0.35,0.05) -- (0.35,-0.05);
    \draw[red!80!black, thick, ->] (0.0,0.65) arc (90:300:0.45);
  \end{scope}
  \draw[red!80!black, thick, ->] (4.85,2.72) -- (5.18,2.92);
  \node[align=center, fill=white, inner sep=2pt] at (2.55,0.38)
    {same shape,\\rotated rigid frame};
  \node[red!80!black, align=center, fill=white, inner sep=2pt] at (5.62,2.28)
    {net rigid\\rotation};
\end{tikzpicture}
\caption{Gauge kinematics: a closed loop in shape space can produce a net rigid
motion. The start and end shape can be the same while the rigid placement has
changed; the effect is the holonomy of a connection.}
\label{fig:gauge-shape-cycle-holonomy}
\end{figure}

\subsection{Mechanical Connection From Zero Momentum}
\label{subsec:mechanical-connection-zero-momentum}

The mechanical connection is obtained by solving a constraint, not by adding a
new force law. If the body has no total rigid momentum, then any internal shape
velocity that would otherwise create angular or linear momentum must be
cancelled by an accompanying rigid velocity. The connection is the linear rule
that performs this cancellation at each shape.

The zero-momentum assumption is a modelling choice, not a mathematical
necessity. It is appropriate for problems such as a freely falling cat or an
isolated rotor device initialized with no angular momentum. With nonzero
momentum, the same kinetic metric still defines a connection, but the
reconstruction equation contains an additional momentum-driven drift. The
zero-momentum case isolates geometric phase because the
entire rigid motion is produced by shape change.

Let the body velocity in the moving frame be
\[
  \xi=g^{-1}\dot g\in\mathfrak g,
\]
where \(\mathfrak g\) is the Lie algebra of \(G\). Suppose the kinetic energy
has the block form
\[
  T
  =
  \frac12
  \begin{pmatrix}
  \xi\\ \dot s
  \end{pmatrix}^T
  \begin{pmatrix}
  M(s) & A(s)\\
  A(s)^T & m(s)
  \end{pmatrix}
  \begin{pmatrix}
  \xi\\ \dot s
  \end{pmatrix}.
\]
This is the quadratic form of kinetic energy written in the variables
\((\xi,\dot s)\), not a linearization of the equations of motion. The block
\(M(s)\) is the locked inertia operator:
it measures the rigid momentum produced by a rigid velocity when the shape is
held fixed, as if the internal shape coordinates were locked at that instant.
The block \(m(s)\) is the shape-space metric for pure deformation
in the chosen frame.  The off-diagonal block \(A(s)\) is the crucial one: it
measures the rigid momentum generated by changing shape while the chosen body
frame is held fixed.

\begin{assumption}[Regular locked inertia]
The zero-momentum reconstruction formula below assumes that \(M(s)\) is
invertible on the rigid motions being reconstructed.  If a shape has an exact
symmetry that makes some rigid velocity physically invisible, then \(M(s)\)
has a null direction and the regular principal-bundle description must be
restricted or quotient further.
\end{assumption}

The momentum conjugate to rigid motion is
\[
  \Pi=\frac{\partial T}{\partial \xi}=M(s)\xi+A(s)\dot s.
\]
The symbol \(\Pi\) is the total rigid-motion momentum written in body
variables: for \(G=\SE(3)\) it packages total linear and angular momentum.
If the total angular and linear momentum vanish,
\[
  \Pi=0,
\]
as they do for an isolated body prepared with zero net external impulse, then
\(\Pi\) remains zero by conservation and
\[
  \xi=-M(s)^{-1}A(s)\dot s.
\]
Writing \(\dot s=\dot s^\alpha\partial_\alpha\), define the connection
\[
  \boxed{
  \mathcal A_\alpha(s)=M(s)^{-1}A_\alpha(s).
  }
\]
The rigid velocity is then
\[
  \boxed{
  g^{-1}\dot g=-\mathcal A_\alpha(s)\dot s^\alpha.
  }
\]
This is the direct mechanical derivation of the gauge field: it is the rule that tells a
zero-momentum deforming body how much rigid motion is forced by an infinitesimal
shape change.

The minus sign is part of the reconstruction convention. A positive connection
component means that the chosen shape change would create rigid momentum in the
body frame; the actual zero-momentum motion supplies the opposite rigid
velocity to cancel it. This is the mechanical meaning behind the gauge-theory
formula.

\subsection{Explicit Inertia-Operator Form}
\label{subsec:explicit-gauge-equation}

A concrete \(\SO(3)\) version of the same construction is useful because it
turns the abstract mechanical connection into a linear algebra problem. Let
\[
  A=R^{-1}\dot R\in\mathfrak{so}(3)
\]
be the angular velocity matrix of the body frame. Let \(Q=Q^T\) encode the
instantaneous shape through the second moment tensor
\[
  Q_{ij}=\int_B\rho(X)X_iX_j\,\dd^3X
\]
in the chosen internal frame. The corresponding inertia tensor is
\[
  \widetilde I_{ij}
  =
  \delta_{ij}\sum_{k=1}^3Q_{kk}-Q_{ij}.
\]
Let \(Y(X,s(t))\) be the position of a material point in the chosen internal
frame, with the center of mass at the origin. The spatial position is
\(x=RY\). In the body frame,
\[
  R^{-1}\dot x=A Y+\dot Y.
\]
The first term is the rigid velocity required by the frame motion; the second
term is internal deformation. The body angular momentum is therefore
\[
  M_{\mathrm{tot}}
  =
  \int_B\rho\,Y\times(AY+\dot Y)\,\dd^3X.
\]
With the hat convention used for the rigid body,
\(\widehat v\,w=v\times w\), or
\[
  (\widehat v)_{ij}=-\epsilon_{ijk}v_k,
\]
the rigid part obeys the matrix identity
\[
  \widehat{\widetilde I\Omega}=AQ+QA,
  \qquad A=\widehat\Omega.
\]
The internal part is
\[
  J_{\mathrm{int}}
  =
  \int_B\rho\,Y\times\dot Y\,\dd^3X.
\]
For zero total angular momentum,
\[
  \widetilde I\Omega+J_{\mathrm{int}}=0.
\]
Define the deformation-generated momentum that the rigid motion must cancel by
\[
  \widetilde J=-J_{\mathrm{int}}.
\]
Then the zero-momentum condition takes the linear matrix form
\[
  \boxed{
  AQ+QA=\widehat{\widetilde J}.
  }
\]
Solving this equation is the explicit finite-dimensional gauge connection:
the angular velocity \(A\) is determined by the shape velocity through
\(\widetilde J\).

\begin{derivation}[Solving the finite-dimensional gauge equation]
Write
\[
  A=\widehat\Omega,\qquad A_{ij}=-\epsilon_{ijk}\Omega_k.
\]
To verify the identity \(AQ+QA=\widehat{\widetilde I\Omega}\), diagonalize the
symmetric tensor \(Q\) at the instant under consideration:
\[
  Q=\diag(q_1,q_2,q_3).
\]
Then
\[
  (AQ+QA)_{12}=-(q_1+q_2)\Omega_3,\quad
  (AQ+QA)_{23}=-(q_2+q_3)\Omega_1,\quad
  (AQ+QA)_{31}=-(q_3+q_1)\Omega_2.
\]
But the inertia tensor has principal moments
\[
  \widetilde I_1=q_2+q_3,\qquad
  \widetilde I_2=q_3+q_1,\qquad
  \widetilde I_3=q_1+q_2,
\]
so these entries are exactly those of
\(\widehat{\widetilde I\Omega}\). Since both sides transform tensorially under
orthogonal changes of the internal frame, the identity holds in any frame.
At zero total angular momentum, the rigid angular momentum must equal
\(\widetilde J\), the negative of the internal contribution. The linear map
from angular velocity \(\Omega\) to rigid angular momentum is precisely the
inertia tensor:
\[
  \widetilde J_i=\widetilde I_{ij}\Omega_j.
\]
Equivalently,
\[
  \Omega_i=(\widetilde I^{-1})_{ij}\widetilde J_j.
\]
Therefore
\[
  \boxed{
  A_{ij}
  =
  -\epsilon_{ijk}
  (\widetilde I^{-1})_{k\ell}\widetilde J_\ell.
  }
\]
This is the concrete \(\SO(3)\) deforming-body gauge formula. In the abstract
notation above, \(M(s)\) is the inertia operator
\(\widetilde I(s)\), while the cross term \(A(s)\dot s\) is the internally
generated angular momentum \(\widetilde J(s,\dot s)\). The helicopter rotor
laboratory in Section~\ref{sec:lab-helicopter-gauge} is the finite-dimensional
special case where the shape variables are two rotor angles and this abstract
inertia-operator equation reduces to a pair of scalar connection components.
\end{derivation}

\subsection{Gauge Transformation And Holonomy}
\label{subsec:gauge-holonomy}

Under a shape-dependent change of body frame \(h(s)\), the connection transforms
as
\[
  \boxed{
  \mathcal A'
  =
  \Ad_{h^{-1}}\mathcal A-h^{-1}\dd h.
  }
\]
The sign of the inhomogeneous term follows from the reconstruction convention
\(g^{-1}\dot g=-\mathcal A(\dot s)\). With the opposite convention, the same
geometric connection is written with the more familiar plus sign. The base space
is shape space and the structure group is the group of rigid motions.

For a path \(s(t)\), the rigid motion is obtained by integrating
\[
  g^{-1}\dot g=-\mathcal A_\alpha(s)\dot s^\alpha.
\]
For a closed shape cycle \(\Gamma\), the net displacement is
\[
  g(T)
  =
  g(0)\,
  \mathcal P\exp\left(
    -\oint_\Gamma \mathcal A
  \right),
\]
where \(\mathcal P\) denotes path ordering. The result can be nontrivial even
though the shape returns to its starting point.
The path-ordered exponential is the limit of multiplying many small
reconstruction steps. If \(0=t_0<t_1<\cdots<t_N=T\), then
\[
  g(t_{j+1})
  \approx
  g(t_j)\exp\left(-\mathcal A_{\alpha}(s(t_j))
  \dot s^\alpha(t_j)\Delta t_j\right),
\]
and the factors must remain in time order because elements of \(G\) need not
commute. Taking the mesh size to zero gives the displayed
\(\mathcal P\exp\) formula.

With the reconstruction convention \(g^{-1}\dot g=-\mathcal A(\dot s)\), the
finite-dimensional curvature used for shape holonomy is
\[
  \boxed{
  \mathcal F
  =
  \dd\mathcal A-\mathcal A\wedge\mathcal A,
  }
\]
or in components,
\[
  \mathcal F_{\alpha\beta}
  =
  \partial_\alpha\mathcal A_\beta
  -
  \partial_\beta\mathcal A_\alpha
  -
  [\mathcal A_\alpha,\mathcal A_\beta].
\]
This sign is tied to the minus sign in the reconstruction equation. The local
frame connection used later for Cosserat and defect geometry follows the
standard Maurer-Cartan convention \(F=\dd A+A\wedge A\).

\begin{derivation}[Small-loop holonomy]
Take shape coordinates \((s^1,s^2)\) and traverse a tiny rectangle based at
\(s_0\): first \(+\Delta s^1\), then \(+\Delta s^2\), then
\(-\Delta s^1\), then \(-\Delta s^2\). Along each side the reconstruction
equation
\[
  g^{-1}\dot g=-\mathcal A_\alpha\dot s^\alpha
\]
produces an infinitesimal group exponential. If \(\mathcal A\) were constant,
the product would be the group commutator
\[
  e^{-\mathcal A_1\Delta s^1}
  e^{-\mathcal A_2\Delta s^2}
  e^{\mathcal A_1\Delta s^1}
  e^{\mathcal A_2\Delta s^2}
  =
  \exp\left(
    [\mathcal A_1,\mathcal A_2]\Delta s^1\Delta s^2
    +O(|\Delta s|^3)
  \right),
\]
with the sign following the reconstruction convention. If \(\mathcal A\) varies
over the rectangle, the two \(s^1\)-edges differ by
\(\partial_2\mathcal A_1\Delta s^2\), and the two \(s^2\)-edges differ by
\(\partial_1\mathcal A_2\Delta s^1\). Combining the derivative terms with the
commutator gives
\[
  g_{\mathrm{loop}}
  =
  \exp\left(
    -\mathcal F_{12}\Delta s^1\Delta s^2
    +O(|\Delta s|^3)
  \right).
\]
This is the finite-dimensional version of the usual curvature-as-holonomy
calculation for a gauge connection.
\end{derivation}

For a small loop enclosing area element
\(\Delta S^{\alpha\beta}\),
\[
  \Delta g
  \approx
  \exp\left(
    -\mathcal F_{\alpha\beta}\Delta S^{\alpha\beta}
  \right).
\]

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1.0, >=Latex, font=\small]
  \draw[->] (-0.2,0) -- (6.9,0) node[right] {\(s^1\)};
  \draw[->] (0,-0.2) -- (0,3.75) node[above] {\(s^2\)};
  \coordinate (a) at (1.35,1.25);
  \coordinate (b) at (3.65,1.25);
  \coordinate (c) at (3.65,2.55);
  \coordinate (d) at (1.35,2.55);
  \fill[blue!10] (a) -- (b) -- (c) -- (d) -- cycle;
  \draw[blue!70!black, thick, ->] (a) -- (b);
  \node[blue!70!black, fill=white, inner sep=1pt, font=\large]
    at (2.50,1.53)
    {\(\delta s^1\)};
  \draw[blue!70!black, thick, ->] (b) -- node[right] {\(\delta s^2\)} (c);
  \draw[blue!70!black, thick, ->] (c) -- (d);
  \draw[blue!70!black, thick, ->] (d) -- (a);
  \draw[red!75!black, thick, ->] (2.35,2.15) arc (210:-80:0.42);
  \begin{scope}[shift={(0.22,0.18)}]
    \draw[red!75!black, dashed, thick]
      (1.35,1.25) -- (3.65,1.25) -- (3.65,2.55) -- (1.35,2.55) -- cycle;
  \end{scope}
  \draw[red!75!black, ->] (4.55,2.78) -- (2.95,2.22);
  \node[red!75!black, align=left, fill=white, inner sep=2pt] at (5.72,3.02)
    {net rigid motion\\\(-\mathcal F_{12}\delta s^1\delta s^2\)};
  \node[align=center, fill=white, inner sep=2pt] at (2.5,0.34)
    {same infinitesimal shape loop,\\different final placement};
\end{tikzpicture}
\caption{Curvature is the local density of holonomy. Traversing a small
rectangle in shape space returns to the same shape but can leave a residual
rigid displacement proportional to the curvature through the enclosed area.}
\label{fig:gauge-curvature-small-loop}
\end{figure}

Thus curvature is the local density of geometric phase. Berry phase, Hannay
angle, and falling-cat reorientation are all holonomies of connections over
parameter or shape spaces, but the constructions differ. Berry (1984)
introduced the quantum \(U(1)\) geometric phase; Hannay (1985) gave the
classical angle holonomy on invariant tori; Shapere and Wilczek (1989)
developed the gauge-kinematic program for deformable bodies; and Montgomery
(1993) gave the gauge-theoretic treatment of the falling cat. The common name
is geometric phase; the underlying mechanisms differ as described above.

Flat connections can still produce frame changes along open paths, but a small
closed loop detects curvature. The curvature is the local observable of the
gauge description. Different choices of body frame change \(\mathcal A\), while
the holonomy around a loop changes by conjugation and therefore represents the
same physical displacement.

\section{A Two-Rotor Deforming-Body Model}
\label{sec:two-rotor-deforming-body-model}

A minimal finite-dimensional example is a rigid carrier with two internal
rotors. The rotor angles are shape variables
\[
  s=(\alpha,\beta)\in S^1\times S^1.
\]
This is the compact version of the helicopter-rotor laboratory in
Section~\ref{sec:lab-helicopter-gauge}: here we strip the device down to its
two noncommuting shape directions so the curvature calculation can be seen in
one page.
Let the carrier have isotropic inertia \(I_0\mathbf 1\). Let one rotor spin
about the body \(z\)-axis with inertia \(I_1\), and another spin about the
body \(y\)-axis with inertia \(I_2\). If \(\Omega\) is the carrier's body
angular velocity, the rotational kinetic energy is
\[
  T=
  \frac12I_0\norm{\Omega}^2
  +
  \frac12I_1(\Omega_z+\dot\alpha)^2
  +
  \frac12I_2(\Omega_y+\dot\beta)^2.
\]
The body angular momentum conjugate to \(\Omega\) is
\[
  \Pi
  =
  I_0\Omega
  +
  I_1(\Omega_z+\dot\alpha)e_z
  +
  I_2(\Omega_y+\dot\beta)e_y.
\]
Imposing zero total angular momentum gives
\[
  \Omega
  =
  -a\,\dot\alpha\,e_z
  -
  b\,\dot\beta\,e_y,
  \qquad
  a=\frac{I_1}{I_0+I_1},
  \qquad
  b=\frac{I_2}{I_0+I_2}.
\]
Thus
\[
  \mathcal A_\alpha=a\,e_z,
  \qquad
  \mathcal A_\beta=b\,e_y.
\]
The connection coefficients are constant, so the derivative part of curvature
vanishes. But \(\SO(3)\) is nonabelian:
\[
  [e_z,e_y]=e_z\times e_y=-e_x.
\]
Therefore
\[
  \mathcal F_{\alpha\beta}
  =
  -[\mathcal A_\alpha,\mathcal A_\beta]
  =
  ab\,e_x.
\]
A small rectangular stroke with sides \(\Delta\alpha,\Delta\beta\), traversed
first in the \(\alpha\) direction and then in the \(\beta\) direction, gives
\[
\begin{aligned}
  g_{\mathrm{loop}}
  &=
  \exp(-\mathcal A_\alpha\Delta\alpha)
  \exp(-\mathcal A_\beta\Delta\beta)
  \exp(\mathcal A_\alpha\Delta\alpha)
  \exp(\mathcal A_\beta\Delta\beta)\\
  &=
  \exp\!\left(
    [\mathcal A_\alpha,\mathcal A_\beta]
    \Delta\alpha\,\Delta\beta
    +O(\Delta^3)
  \right),
\end{aligned}
\]
by the Baker-Campbell-Hausdorff formula. With the conventions above the
leading rotation vector is
\[
  -\mathcal F_{\alpha\beta}\Delta\alpha\,\Delta\beta
  =
  -ab\,\Delta\alpha\,\Delta\beta\,e_x;
\]
reversing the traversal reverses the sign. The essential content is the
holonomy: a closed loop in internal shape variables produces a rigid rotation
even though the rotors return to their starting angles. Like the falling cat,
this is a nonabelian \(\SO(3)\) holonomy over shape space, although the
physical constraints and shape coordinates are different.

\section{Swimming At Low Reynolds Number}
\label{sec:low-reynolds-swimming}

Shapere and Wilczek's low-Reynolds-number swimming problem gives a particularly
physical gauge theory. At very small Reynolds number, inertia drops out of
Navier-Stokes and the surrounding fluid satisfies Stokes equations:
\[
  -\nabla p+\mu\Delta u=0,\qquad \nabla\cdot u=0.
\]
Boundary velocity is prescribed by the swimmer's rigid motion plus deformation
rate. Because the Stokes problem is linear, the force and torque on the body
depend linearly on:
\[
  \xi=g^{-1}\dot g
  \qquad\text{and}\qquad
  \dot s.
\]
A freely swimming body has zero net force and torque, so the same algebra as
above gives
\[
  \xi=-\mathcal A_\alpha(s)\dot s^\alpha.
\]
The connection is now computed from a hydrodynamic boundary-value problem. More
explicitly, solving the Stokes problem with unit rigid velocity gives a
resistance matrix \(R(s)\), and solving it with unit shape velocity gives a
shape-force matrix \(B(s)\). The zero-force and zero-torque condition has the
linear form
\[
  R(s)\xi+B_\alpha(s)\dot s^\alpha=0,
\]
so
\[
  \xi=-R(s)^{-1}B_\alpha(s)\dot s^\alpha.
\]
Thus \(\mathcal A_\alpha=R^{-1}B_\alpha\).
The phrase ``solving the Stokes problem'' includes the no-slip boundary
condition on the swimmer surface, decay or wall conditions in the surrounding
fluid, and the force/torque integrals of the hydrodynamic stress over the
surface. Thus the connection is not inserted by hand; it is the
finite-dimensional summary of a linear elliptic boundary-value problem.
The Lorentz reciprocal theorem gives the symmetry of the Stokes resistance
operator under the usual boundary conditions, which is why the rigid-motion
resistance matrix is positive definite after quotienting out impossible rigid
motions.

A reciprocal one-parameter stroke encloses no area in shape space and produces
no net motion: this is Purcell's 1977 scallop theorem in gauge language. In
one shape coordinate \(s\), the reconstruction is
\[
  g^{-1}\dot g=-\mathcal A(s)\dot s.
\]
Traversing a path from \(s_0\) to \(s_1\) and then retracing it reverses
\(\dot s\,dt=\dd s\), so the ordered product of rigid motions is inverted on
the return leg. At zero Reynolds number there is no inertial memory to
distinguish the outward stroke from the reversed stroke, hence the net
displacement is trivial. A two-parameter stroke can enclose area and swim by
curvature. This is the same geometric-mechanics language developed in the
Marsden-Weinstein reduction tradition; later geometric swimmer models, including
work by Kanso, Marsden, and Rowley, make the connection computation explicit for
finite-dimensional shape families.

\section{Local Frames In Elastic Continua}
\label{sec:local-frames-elastic-continua}

The finite-dimensional gauge picture becomes a field theory when every material
point carries a local frame. A Cosserat body has, in addition to
\(\varphi(X,t)\), a rotation field
\[
  R(X,t)\in\SO(3).
\]
The directors are
\[
  d_a(X,t)=R(X,t)E_a,
\]
where \(E_a\) is a fixed reference basis. The rotational connection is
\[
  A_A=R^{-1}\partial_A R\in\mathfrak{so}(3),
  \qquad
  A_t=R^{-1}\partial_t R.
\]
The translational strain measured in the local frame is
\[
  \Gamma_A=R^{-1}\partial_A\varphi.
\]
For a rigid body, \(R\) depends only on time. For a Cosserat continuum, \(R\)
depends on \(X\), so neighboring material points can rotate relative to one
another.

This is the field-theoretic version of the rigid-body frame issue. In the
rigid body there is one moving frame; in a Cosserat body there is a moving frame
at every material point. The connection \(A_A\) compares nearby frames, while
\(A_t\) records how a given local frame evolves in time.

\begin{definition}[Local frame gauge transformation]
Let \(G(X,t)\in\SO(3)\) be a change of local material frame:
\[
  R'(X,t)=R(X,t)G(X,t).
\]
Then
\[
  A'_A=G^{-1}A_AG+G^{-1}\partial_A G,
  \qquad
  \Gamma'_A=G^{-1}\Gamma_A.
\]
Thus \(A_A\) is a gauge connection and \(\Gamma_A\) is a vector-valued
matter field in the local frame.
\end{definition}

The simplest elastic energy for a Cosserat continuum penalizes deviations of
these gauge-covariant strains from preferred values:
\[
  W_{\mathrm{Cosserat}}
  =
  \frac12 C^{AB}_{ab}
  \left(\Gamma_A^a-\Gamma_{0A}^a\right)
  \left(\Gamma_B^b-\Gamma_{0B}^b\right)
  +
  \frac12 D^{AB}_{ab}
  \left(K_A^a-K_{0A}^a\right)
  \left(K_B^b-K_{0B}^b\right).
\]
Here \(K_A\in\R^3\) is the axial-vector form of the skew matrix \(A_A\):
\[
  (A_A)^a{}_b=-\epsilon^a{}_{bc}K_A^c.
\]
The tensors \(C\) and \(D\) encode stretching/shearing and bending/twisting
stiffnesses. The preferred fields \(\Gamma_0,K_0\) represent intrinsic strain
and intrinsic curvature. This formula is schematic but important: it shows how
ordinary elastic energy becomes an energy on translational strain and gauge
connection data.

\begin{remark}[Rigid body as the zero-gradient special case]
If \(R(X,t)=R(t)\) and
\[
  \varphi(X,t)=r(t)+R(t)X,
\]
then
\[
  A_A=R^{-1}\partial_A R=0,
  \qquad
  \Gamma_A=R^{-1}\partial_A\varphi=E_A.
\]
Thus a rigid body is the special Cosserat field with no spatial frame
connection and no translational strain beyond the reference directors. A
deformable Cosserat body is obtained by allowing \(R\) and \(\Gamma_A\) to vary
with \(X\).
\end{remark}

\section{Curvature, Torsion, And Defects}
\label{sec:curvature-torsion-defects}

If a connection is globally of the form
\[
  A_A=R^{-1}\partial_A R,
\]
then it is pure gauge and satisfies the Maurer-Cartan identity:
\[
  F_{AB}
  =
  \partial_AA_B-\partial_BA_A+[A_A,A_B]
  =
  0.
\]
Conversely, on a simply connected material patch, \(F=0\) implies locally that
the connection is pure gauge: parallel transport is path independent, so one
can define \(R(X)\) by transporting a chosen initial frame from a base point to
\(X\), and this \(R\) satisfies \(A=R^{-1}\dd R\). Path independence follows
from applying Stokes' theorem to small spanning surfaces and using \(F=0\), then
subdividing a homotopy between two paths into such small surfaces.
Equivalently, the parallel transport around a small coordinate rectangle is
\[
  \exp\!\left(F_{AB}\,\Delta X^A\Delta X^B+O(\Delta^3)\right),
\]
so vanishing curvature removes the infinitesimal obstruction; simple
connectedness then removes the global path-dependence obstruction.
Nonzero curvature means that local frames cannot be integrated into a single
smooth orientation field without singularities or defects. In defect theory,
curvature is associated with disclinations, or rotational defects. More
precisely, in three dimensions the disclination-density tensor is the Hodge
dual of the curvature two-form, just as the dislocation-density tensor below is
the Hodge dual of torsion. Classical continuum defect theory was developed in
the 1950s and 1960s by Kondo, Bilby, Kroner, and others.

To describe translational defects, introduce a local coframe \(\theta^a\), which
measures material line elements in a local frame. The connection one-forms
\(\omega^a{}_b\) are the components of the same \(\mathfrak{so}(3)\)-valued
connection written above: if the matrix entries of \(A_A\) are
\((A_A)^a{}_b\), then
\[
  \omega^a{}_b=(A_A)^a{}_b\,\dd X^A.
\]
With this notation, the torsion two-form is
\[
  T^a=\dd\theta^a+\omega^a{}_b\wedge\theta^b.
\]
Torsion measures closure failure of infinitesimal parallelograms and is
associated with dislocation density. In three spatial dimensions the vector
defect-density fields often used in materials science are Hodge duals of these
two-forms; for example
\((*F^a{}_b)_i=\frac12\epsilon_i{}^{jk}F^a{}_{b\,jk}\) and
\((*T^a)_i=\frac12\epsilon_i{}^{jk}T^a{}_{jk}\). In this
language:
\[
\begin{aligned}
  \text{curvature} &\longleftrightarrow \text{rotational mismatch},\\
  \text{torsion} &\longleftrightarrow \text{translational mismatch},\\
  \text{incompatible strain} &\longleftrightarrow \text{nonintegrable local data}.
\end{aligned}
\]
The compatibility identities are the differential bookkeeping rules for these
defect densities. With covariant exterior derivative \(D\),
\[
  D F=0,
  \qquad
  D T^a=F^a{}_b\wedge\theta^b.
\]
They follow directly from the definitions. For curvature
\(F=\dd\omega+\omega\wedge\omega\),
\[
  DF=\dd F+\omega\wedge F-F\wedge\omega=0,
\]
because
\[
\begin{aligned}
  \dd F
  &=\dd\omega\wedge\omega-\omega\wedge\dd\omega,\\
  \omega\wedge F-F\wedge\omega
  &=\omega\wedge\dd\omega-\dd\omega\wedge\omega
    +\omega\wedge\omega\wedge\omega-\omega\wedge\omega\wedge\omega,
\end{aligned}
\]
so all terms cancel. For torsion,
\[
  DT^a
  =
  D(\dd\theta^a+\omega^a{}_b\wedge\theta^b)
  =
  F^a{}_b\wedge\theta^b.
\]
The first identity says that disclination density satisfies a Bianchi identity.
The second says that dislocation density is not independent of rotational
curvature when the local frame itself has curvature. These identities are the
continuum-defect analogue of \(\nabla\cdot(\nabla\times A)=0\): they tell which
defect fields can be prescribed consistently.

The gauge theory records which descriptions depend on arbitrary local choices of
frame and which quantities are invariant.

\section{Cosserat Rod As A One-Dimensional Gauge Theory}
\label{sec:cosserat-rod}

A planar rod has material coordinate \(s\), centerline \(r(s)\in\R^2\), and
orientation angle \(\theta(s)\). Its director is
\[
  d_1=(\cos\theta,\sin\theta).
\]
Eugene and Francois Cosserat (1909) introduced the idea that a continuum point
can carry an independent local frame, not only a position.
Let
\[
  d_2=(-\sin\theta,\cos\theta).
\]
Then the moving-frame equations are
\[
  d_1'=\theta' d_2,\qquad d_2'=-\theta' d_1.
\]
Thus curvature is the connection coefficient for the rotation of the director
frame along the rod:
\[
  \kappa=\theta'.
\]
For an inextensible unshearable rod,
\[
  r'(s)=d_1(s).
\]
Given \(\kappa\),
\[
  \theta(s)=\theta(0)+\int_0^s\kappa(\sigma)\,\dd\sigma,
\]
and
\[
  r(s)=r(0)+\int_0^s
  (\cos\theta(\sigma),\sin\theta(\sigma))\,\dd\sigma.
\]
This is the reconstruction problem: a gauge field \(\kappa\,\dd s\) determines
shape after choosing initial position and frame.

The bending energy
\[
  E_{\mathrm{bend}}
  =
  \frac12\int_0^L B(\kappa-\kappa_0)^2\,\dd s
\]
compares actual curvature with preferred curvature. Here \(B\) is bending
stiffness and \(\kappa_0\) is intrinsic curvature. In gauge language, the
energy penalizes mismatch between a connection and a preferred connection.

The Euler-Lagrange equation is short in this one-dimensional case. Since
\(\kappa=\theta'\),
\[
  \delta E_{\mathrm{bend}}
  =
  \int_0^L B(\kappa-\kappa_0)\delta\theta'\,\dd s
  =
  \left[M\,\delta\theta\right]_0^L
  -
  \int_0^L M'\delta\theta\,\dd s,
\]
where
\[
  M=B(\kappa-\kappa_0)
\]
is the bending moment. With fixed end angles and no distributed applied couple,
stationarity gives
\[
  M'=0.
\]
If endpoint angles are free, the boundary term supplies the natural moment
condition \(M=0\) at the free end. With external forces and the constraint
\(r'=d_1\), force balance adds the usual Cosserat equations
\[
  n'(s)+f(s)=0,\qquad M'(s)+d_1(s)\times n(s)+\ell(s)=0,
\]
where \(n\) is internal force, \(f\) is applied force per length, and
\(\ell\) is applied couple per length. These equations are the one-dimensional
balance laws behind the reconstruction formulas above.
In this planar notation \(d_1\times n\) denotes the scalar out-of-plane moment
component; in a spatial Cosserat rod the same balance is vector-valued.
They follow from the constrained variational functional
\[
  \int_0^L
  \left[
    \frac12B(\theta'-\kappa_0)^2
    +n\cdot(r'-d_1)
    -f\cdot r-\ell\,\theta
  \right]\dd s.
\]
Variation with respect to \(r\) gives \(n'+f=0\). Variation with respect to
\(\theta\) gives the moment equation, because
\(\delta d_1=\delta\theta\,d_2\) and
\[
  \delta\!\left[n\cdot(r'-d_1)\right]
  =
  -n\cdot d_2\,\delta\theta
  =
  -(d_1\times n)\delta\theta .
\]
The \(\theta'\)-variation contributes \(-M'\delta\theta\) after integration by
parts, while the applied couple contributes \(-\ell\,\delta\theta\). The
coefficient of \(\delta\theta\) is therefore
\[
  -M'-d_1\times n-\ell,
\]
which gives \(M'+d_1\times n+\ell=0\).

\section{What To Take Forward}
\label{sec:deforming-body-take-forward}

The progression from rigid body to the gauge theory of deforming bodies is:
\[
\begin{aligned}
  \text{rigid body:}&\quad R(t),\quad R^{-1}\dot R,\\
  \text{elastic body:}&\quad \varphi(X,t),\quad F=\partial_X\varphi,\quad W(F),\\
  \text{deformable body:}&\quad g(t),\ s(t),\quad
  g^{-1}\dot g=-\mathcal A(s)\dot s,\quad
  \mathcal F=\dd\mathcal A-\mathcal A\wedge\mathcal A,\\
  \text{Cosserat continuum:}&\quad R(X,t),\quad R^{-1}\partial_A R,\\
  \text{defect theory:}&\quad A,\quad F=\dd A+A\wedge A,\quad T=\dd\theta+\omega\wedge\theta.
\end{aligned}
\]
The first line is ordinary rigid-body mechanics. The second is
classical elasticity: deformation gradients, stored energy, stress, and elastic
waves. The third is Shapere-Wilczek gauge kinematics over shape space. The
fourth attaches a local rigid body to every material point. The fifth allows
the local frame and
translation data to be incompatible, producing curvature and torsion as physical
defect densities.

\section{Chapter-End Laboratories And Exercises}
\label{sec:deforming-body-chapter-end-labs}

These laboratories make the gauge language operational: a small shape loop
measures curvature, and a finite-dimensional rotor model shows how
noncommuting connection components produce body rotation.

\subsection{Mechanical Connection: Curvature From A Small Shape Loop}
\label{sec:detailed-connection-curvature}

Let \(s^1,s^2\) be two shape coordinates and let the zero-momentum
reconstruction equation be
\[
  g^{-1}\dot g
  =
  -\mathcal A_\alpha(s)\dot s^\alpha.
\]
For a tiny rectangular loop based at \(s_0\), first move by
\(\Delta s^1\), then \(\Delta s^2\), then back by \(-\Delta s^1\), then back
by \(-\Delta s^2\). To second order in the side lengths, the net group element
is
\[
  g_{\mathrm{loop}}
  =
  \exp(-\mathcal A_1\Delta s^1)
  \exp(-\mathcal A_2\Delta s^2)
  \exp(\mathcal A_1\Delta s^1)
  \exp(\mathcal A_2\Delta s^2)
\]
plus the correction from the variation of \(\mathcal A_\alpha(s)\) across the
loop. The Baker-Campbell-Hausdorff formula gives the commutator contribution,
while the spatial variation gives derivative terms.\footnote{We use only the
leading commutator term
\(\exp X\exp Y\exp(-X)\exp(-Y)=\exp([X,Y]+O(\|(X,Y)\|^3))\). Higher terms are
higher nested commutators. In the displayed loop
\(X=-\mathcal A_1\Delta s^1\) and
\(Y=-\mathcal A_2\Delta s^2\), so \([X,Y]=[\mathcal A_1,\mathcal A_2]\Delta
s^1\Delta s^2\).} The result is
\[
  g_{\mathrm{loop}}
  =
  \exp\left(
    -\mathcal F_{12}\Delta s^1\Delta s^2
    +
    O(|\Delta s|^3)
  \right),
\]
where
\[
  \boxed{
  \mathcal F_{12}
  =
  \partial_1\mathcal A_2-\partial_2\mathcal A_1
  -
  [\mathcal A_1,\mathcal A_2].
  }
\]
This is why curvature is the local density of geometric phase. Even if the
connection components are constant, a nonabelian structure group can give
nonzero holonomy through the commutator term. The helicopter rotor laboratory
below is a concrete instance of this constant-connection case.

\subsection{Helicopter Rotor Model As A Gauge Connection}
\label{sec:lab-helicopter-gauge}

A toy helicopter model makes the Shapere-Wilczek connection finite-dimensional.
This subsection does not introduce a second theory. It takes the connection
coefficients already derived in Section~\ref{sec:two-rotor-deforming-body-model}
and uses them as a worked laboratory: the shape coordinates are the two rotor
angles \((\alpha,\beta)\), the curvature is a commutator, and the rectangular
stroke can be compared directly with the small-loop holonomy formula above.

The calculation has one main checkpoint. The rotor angles return to their
initial values after a closed stroke, so any final rotation of the body cannot
be attributed to a changed shape. It must come from the noncommuting
connection components. This is the finite-dimensional version of geometric
phase.

As derived in Section~\ref{sec:two-rotor-deforming-body-model}, the
zero-momentum connection for this laboratory model is
\[
  \mathcal A_\alpha=
  \frac{I_1}{I_0+I_1}e_z,\qquad
  \mathcal A_\beta=
  \frac{I_2}{I_0+I_2}e_y.
\]
The components are constant, so the only curvature comes from noncommutativity.
With the reconstruction convention of this chapter,
\[
  \mathcal F_{\alpha\beta}
  =
  -[\mathcal A_\alpha,\mathcal A_\beta]
  =
  \frac{I_1I_2}{(I_0+I_1)(I_0+I_2)}\,e_x.
\]
For a rectangular stroke of signed area
\(\Delta\alpha\,\Delta\beta\), the leading rigid rotation is
\[
  \Delta g
  \approx
  \exp\left(
    -\mathcal F_{\alpha\beta}\Delta\alpha\,\Delta\beta
  \right).
\]
This is the point of the exercise: because the rotor angles return to their
initial values, any final rotation of the body is holonomy, not a changed
shape.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1.0, >=Latex, font=\small]
  \begin{scope}
    \draw[->] (-0.2,0) -- (4.4,0) node[right] {\(\alpha\)};
    \draw[->] (0,-0.2) -- (0,3.1) node[above] {\(\beta\)};
    \fill[blue!8] (0.85,0.65) rectangle (3.25,2.35);
    \draw[blue!70!black, very thick, -{Latex[length=3mm]}] (0.85,0.65) -- (3.25,0.65)
      node[midway, below] {\(\Delta\alpha\)};
    \draw[blue!70!black, very thick, -{Latex[length=3mm]}] (3.25,0.65) -- (3.25,2.35)
      node[midway, right] {\(\Delta\beta\)};
    \draw[blue!70!black, very thick, -{Latex[length=3mm]}] (3.25,2.35) -- (0.85,2.35)
      node[midway, above] {\(-\Delta\alpha\)};
    \draw[blue!70!black, very thick, -{Latex[length=3mm]}] (0.85,2.35) -- (0.85,0.65)
      node[midway, left] {\(-\Delta\beta\)};
    \node[align=center, fill=white, inner sep=1.5pt] at (2.05,3.35)
      {closed shape loop\\on the rotor torus};
  \end{scope}
  \draw[gray!85, thick, ->] (4.22,2.78) .. controls (4.70,2.42) .. (5.12,1.78);
  \node[above, fill=white, inner sep=1pt] at (4.35,2.98)
      {\([\mathcal A_\alpha,\mathcal A_\beta]\)};
  \begin{scope}[xshift=5.4cm]
    \draw[->, thick] (0,0) -- (1.6,0) node[right] {\(e_x\)};
    \draw[->, thick] (0,0) -- (0,1.6) node[above] {\(e_z\)};
    \draw[->, thick] (0,0) -- (-1.1,-0.8) node[left] {\(e_y\)};
    \draw[red!75!black, thick, ->] (0.95,0.48)
      arc[start angle=90, end angle=-230, x radius=0.22, y radius=0.48];
    \node[red!75!black, fill=white, inner sep=1pt] at (1.40,0.55)
      {rotation about \(e_x\)};
    \node[black!65, fill=white, inner sep=1pt] at (-0.55,1.85)
      {body-frame axes};
    \node[red!75!black, align=center, fill=white, inner sep=1.5pt] at (1.15,-1.62)
      {net body rotation\\from \([\mathcal A_\alpha,\mathcal A_\beta]\)};
  \end{scope}
\end{tikzpicture}
\caption{Helicopter-rotor gauge laboratory. The connection components are
constant, but they point in noncommuting body directions. A closed rectangle in
shape space therefore has curvature from the commutator term and can produce a
net body rotation.}
\label{fig:lab-helicopter-gauge-loop}
\end{figure}
\FloatBarrier

\section{Associated Demos}
\label{sec:deforming-body-demos}

\begin{itemize}
\item \path{demos/python/cosserat_rod_demo.py}, a planar frame reconstruction
from curvature.
\item \path{demos/python/deforming_body_gauge.py}: finite-dimensional
two-rotor gauge holonomy.
\item \path{demos/mathematica/GaugeDeformingBody.wl}, a finite-dimensional
gauge-connection toy model.
\item Suggested extension: a low-Reynolds swimmer whose connection is computed
from a Stokes boundary-value problem.
\end{itemize}
