\documentclass[11pt]{book}

\input{preamble}

\title{Notes on Classical Mechanics}
\author{GPT 5.5 and Xi Yin}
\date{\today}

\begin{document}

\frontmatter
\begin{titlepage}
\thispagestyle{empty}
\centering
\vspace*{0.16\textheight}
{\huge Notes on Classical Mechanics\par}
\vspace{1.25cm}
{\Large GPT 5.5 and Xi Yin\par}
\vspace{0.6cm}
{\large \today\par}
\vfill
\begin{minipage}{0.88\textwidth}
\small
\setlength{\parindent}{0pt}
These notes develop classical mechanics as a self-contained sequence of models,
structures, and examples.  The first layer establishes the common language:
configuration space, action principles, constraints, symmetries, momenta, the
Legendre transform, Hamiltonian flow, and symplectic phase space.  The second
layer uses Kepler motion and rigid bodies to make that language concrete, then
builds integrability through action-angle variables and invariant tori.  The
third layer studies perturbation: canonical transformations, resonances, KAM
mechanisms, breakdown of tori, and restricted three-body dynamics.  The final
layer carries the same structural ideas into continuum mechanics by treating
elastic media and shape-changing bodies as material systems before turning to
fluids, where Eulerian fields, Navier-Stokes dynamics, exact flows, stability
criteria, and vorticity become the natural variables.  Figures, exercises,
Python demonstrations, and Wolfram Language notebooks are placed next to the
derivations so that the architecture of the theory, its computations, and its
physical examples reinforce one another.
\end{minipage}
\vspace*{0.10\textheight}
\end{titlepage}

\tableofcontents

\chapter*{Preface}

Mechanics is developed here from the structures that give its equations
meaning. The starting point is the configuration space and the action. From
there the notes pass to phase space, symplectic geometry, canonical
transformations, integrability, perturbation, chaos, elasticity, gauge
kinematics of deformation, and fluids. Kepler motion, the rigid body, the
standard map, the restricted Sun-Jupiter-asteroid model, deforming bodies,
laminar flows, and point vortices are the main laboratories.

The intended reading style is active. At each derivation, identify the space,
the variables, the assumptions, the conserved quantities, and the geometric
object being preserved or varied. At each figure, ask which equation or
reduction the picture is meant to make visible. At each numerical demo, ask
which continuous structure the discretization respects and which quantities
must be monitored.

The exposition uses a repeated rhythm. First state the model in words. Then
name the mathematical space on which the model lives. Then derive the equation
from a principle or balance law. Coordinates, phase portraits, and simulations
come after those choices have been made. In this order, coordinate formulas
remain attached to the geometric or physical structures they represent.

The notes are also written for two passes. On a first pass, follow the
definitions, boxed equations, examples, and the ``what to take forward''
sections. On a second pass, return to the derivations and ask why each
assumption was needed. A formula may be a definition in one section, a
constraint in another, and an equation of motion only after a principle has
been applied. The text names these roles because mechanics becomes clearer when
definitions, identities, constraints, and dynamical laws are kept separate.

\paragraph{How the chapters fit together.}
The notes are meant to be read linearly, but each chapter also has a precise
role in the overall architecture:
\[
\begin{array}{ccl}
\text{reading rules and assumptions} &\longrightarrow&
\text{well-defined models introduced in context},\\
\text{Lagrangian mechanics} &\longrightarrow&
\text{equations from variation and symmetry},\\
\text{Hamiltonian mechanics} &\longrightarrow&
\text{phase space and generating functions},\\
\text{rigid body, tori, and actions} &\longrightarrow&
\text{integrability and action variables},\\
\text{perturbation and chaos} &\longrightarrow&
\text{canonical perturbation, KAM, and transport},\\
\text{three-body problem} &\longrightarrow&
\text{celestial reduction, restricted dynamics, and ejection},\\
\text{elastic bodies} &\longrightarrow&
\text{stress, strain, and boundary-value problems},\\
\text{deforming bodies} &\longrightarrow&
\text{shape space, connection, curvature, holonomy},\\
\text{fluids} &\longrightarrow&
\text{Eulerian fields, Navier-Stokes, exact flows, stability}.
\end{array}
\]
This dependency map is also a warning about shortcuts. For example, KAM theory
uses action-angle variables and near-identity generating functions. The
Liouville-torus construction explains the invariant tori, and canonical
perturbation theory tests what survives when the system is no longer exactly
integrable.
Similarly, the gauge theory of deforming bodies is clearest only after one has
separated configuration, symmetry, momentum, and connection in the rigid-body
chapters.

\paragraph{How to use the computational material.}
The scripts are controlled experiments attached to the derivations. A good run
tells the same story as the equations: which invariant is monitored, which
approximation is being made, which parameter controls the effect, and which
observed feature might be a numerical artifact. When a demo reports drift in
energy, Jacobi constant, symplectic area, or orthogonality, that diagnostic is
part of the mechanics.

\mainmatter
\input{00_notation_assumptions}
\input{01_lagrangian_mechanics}
\input{02_hamiltonian_phase_space}
\input{03_integrability_rigid_body}
\input{04_perturbation_kam_torus_breakdown}
\input{05_three_body_problem}
\input{06_elastic_bodies}
\input{07_deforming_body_gauge}
\input{08_fluid_mechanics}

\backmatter
\chapter*{Further Directions}

Natural continuations include Dirac constraints and symplectic reduction,
Delaunay variables for celestial mechanics, resonance normal forms, Arnold
diffusion, heavy-top dynamics, numerical Poincare-section construction, and
structure-preserving integrators for fluids and deforming bodies. Each addition
should follow the same standard used here: state the model, define the symbols,
derive the equations, draw the geometry, and connect the mathematics to a
reproducible computation.

\chapter*{Suggested Further Reading}

For Lagrangian and Hamiltonian mechanics, Arnold's \emph{Mathematical Methods
of Classical Mechanics} and Goldstein, Poole, and Safko's \emph{Classical
Mechanics} are natural companions. For symplectic geometry, reduction, and
momentum maps, see Abraham and Marsden's \emph{Foundations of Mechanics} and
Marsden and Ratiu's \emph{Introduction to Mechanics and Symmetry}. For
perturbation theory and KAM context, Arnold, Kozlov, and Neishtadt's
\emph{Mathematical Aspects of Classical and Celestial Mechanics} and Holmes's
\emph{Introduction to Perturbation Methods} are useful next steps. For fluids
and elasticity, standard references include Batchelor's \emph{An Introduction
to Fluid Dynamics}, Drazin and Reid's \emph{Hydrodynamic Stability}, and
Marsden and Hughes's \emph{Mathematical Foundations of Elasticity}. For
geometric phases and deforming bodies, read Shapere and Wilczek's original
gauge-kinematics papers alongside modern discussions of mechanical
connections.

\end{document}
